---
title: "Basic facts and notions of logic and set theory"
date: 2020-04-11 18:49:34
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}

library(tidyverse)
library(kableExtra)
library(tidyquant)
library(knitr)
```

# Logic and set theory

This section is based on [@mendelson_number_2008, Chapter 1, pp 1-52]

This is a very brief introduction to the basic concepts and terminology of mathematical logic and set theory

For another introduction to set theory see [@halmos_naive_1974] and for an extensive survey of the set theory see [@jech_set_2003]

For another introduction to mathematical logic see [@smullyan_beginners_2014] and for an extensive survey of mathematical logic see [@bilaniuk_problem_2009]

## Logical connectives

### Conjuction

Given sentences $A$ and $B$, we may form a new sentence $A \text{ and } B$, which we shall write $A \land B$

The relation between the new sentence $A \land B$ and the component sentences $A,B$ can be made clear by a __truth table__:

```{r truth_table_conjuction}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\land B$` = c("T", "F", "F", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

In this table, as in all tables to follow, T stands for __True__ and F for __False__

### Negation

If $A$ is a sentence, the $\text{not } A$ is also a sentence, which we shall denote $\neg A$

$\neg A$ is called the __negation__ of $A$

The corresponding __truth table__ is:

```{r truth_table_negation}

tibble(`$A$` = c("T", "F"),
       `$\\neg A$` = c("F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

### Disjuction

Given sentences $A$ and $B$, we may form a new sentence $A \text{ or } B$, which we shall write $A \lor B$

$\lor$ is understood in the __inclusive sense__, namely, $A \lor B$ is true when $A$ is true or $B$ is true or both

The corresponding __true table__ is

```{r truth_table_disjuction}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\lor B$` = c("T", "T", "T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

If a set of formulas take the same truth values we said that these formulas are __equivalent__. For example:

- $A \lor A$ and $A$

    + When $A$ is T, so is $A \lor A$, and viceversa

- $A \land A$ and $A$

    + $A$ is T if and only if $A \land A$ is T
    
- $A \lor (B \land A)$ and $A$

    + When $A$ is T, so is $A \lor (B \land A)$. If $A$ is F, then $A \lor (B \land A)$ is F. Using a __true table__ we can see that:
    
```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\land B$` = c("T", "F", "F", "F"),
       `$A \\lor (A \\land B)$` = c("T", "F", "T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

## Logical connectives exercises and solutions

1. Write the following English sentences in terms of our new notation. Use letters $A, B, C,...$ to stand for the basic sentences

    a. I will go to the movies or to the beach
    
        + $A = \text{I will go to the movies}$
        
        + $B = \text{I will go to the beach}$
        
        + $A \lor B = \text{I will go to the movies or to the beach}$
        
    b. Peter is handsome but Paul is not
    
        + $A = \text{Peter is handsome}$
        
        + $B = \text{Paul is handsome}$
        
        + $A \land \neg B = \text{Peter is handsome but Paul is not}$
        
    c. Either $x$ has a square root or $x$ is not positive
    
        + $A = x \text{ has a square root}$
        
        + $B = x \text{ is positive}$
        
        + $A \lor \neg B = \text{Either } x \text{ has a square root or } x \text{ is not positive}$

    d. It is not the case that $x$ is both positive and negative
    
        + $A = x \text{ is positive}$
        
        + $B = x \text{ is negative}$
        
        + $\neg (A \land B) = \text{It is not the case that } x \text{ is both positive and negative}$
        
    e. Neither John not James is here
    
        + $A = \text{John is here}$
        
        + $B = \text{James is here}$
        
        + $\neg (A \land B) = \text{Neither John not James is here}$
        
2. Render the following into idiomatic English

    a. $x > 2 \lor x < 2$
    
        + $x$ is greater than or less than $2$
        
    b. $\neg (y \text{ is rational } \lor y \text{ is trascendental})$
    
        + Neither $y$ is rational or transcendental
        
    c. $e > 2 \land e < 3$
    
        + $e$ is greater than $2$ and less than $3$
    
    d. $e > 2 \land e \ngtr 3$
    
        + $e$ is greater than $2$ but not greater than $3$
        
3. Show that the following pairs of formulas always take the same truth values, and, therefore, are, in a certain sense, equivalent:

    a. $\neg (A \lor B)$ and $(\neg A) \land (\neg B)$
    
```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\lor B$` = c("T", "T", "T", "F"),
       `$\\neg (A \\lor B)$` = c("F", "F", "F", "T"),
       `$(\\neg A)$` = c("F", "T", "F", "T"),
       `$(\\neg B)$` = c("F", "F", "T", "T"),
       `$(\\neg A) \\land (\\neg B)$` = c("F", "F", "F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(4, 7), background = palette_light()[[7]])
```

    b. $\neg (A \land B)$ and $(\neg A) \lor (\neg B)$

```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\land B$` = c("T", "F", "F", "F"),
       `$\\neg (A \\land B)$` = c("F", "T", "T", "T"),
       `$(\\neg A)$` = c("F", "T", "F", "T"),
       `$(\\neg B)$` = c("F", "F", "T", "T"),
       `$(\\neg A) \\lor (\\neg B)$` = c("F", "T", "T", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(4, 7), background = palette_light()[[7]])
```
   
    c. $\neg \neg B$ and $B$

```{r }

tibble(`$B$` = c("T", "F"),
       `$\\neg B$` = c("F", "T"),
       `$\\neg \\neg B$` = c("T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(1, 3), background = palette_light()[[7]])
```

    d. $A \land (B \lor C)$ and $(A \land B) \lor (A \land C)$

```{r }

tibble(`$A$` = c("T", "T", "T", "T", "F", "F", "F", "F"),
       `$B$` = c("T", "T", "F", "F", "T", "T", "F", "F"),
       `$C$` = c("T", "F", "T", "F", "T", "F", "T", "F"),
       `$B \\lor C$` = c("T", "T", "T", "F", "T", "T", "T", "F"),
       `$A \\land (B \\lor C)$` = c("T", "T", "T", "F", "F", "F", "F", "F"),
       `$A \\land B$` = c("T", "T", "F", "F", "F", "F", "F", "F"),
       `$A \\land C$` = c("T", "F", "T", "F", "F", "F", "F", "F"),
       `$(A \\land B) \\lor (A \\land C)$` = c("T", "T", "T", "F", "F", "F", "F", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(5, 8), background = palette_light()[[7]])
```

    e. $A \lor (B \land C)$ and $(A \lor B) \land (A \lor C)$

```{r }

tibble(`$A$` = c("T", "T", "T", "T", "F", "F", "F", "F"),
       `$B$` = c("T", "T", "F", "F", "T", "T", "F", "F"),
       `$C$` = c("T", "F", "T", "F", "T", "F", "T", "F"),
       `$B \\land C$` = c("T", "F", "F", "F", "T", "F", "F", "F"),
       `$A \\lor (B \\land C)$` = c("T", "T", "T", "T", "T", "F", "F", "F"),
       `$A \\lor B$` = c("T", "T", "T", "T", "T", "T", "F", "F"),
       `$A \\lor C$` = c("T", "T", "T", "T", "T", "F", "T", "F"),
       `$(A \\lor B) \\land (A \\lor C)$` = c("T", "T", "T", "T", "T", "F", "F", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(5, 8), background = palette_light()[[7]])
```

## Conditionals

Given sentences $A$ and $B$, we may form a new sentence $\text{If } A \text{ then } B$, which we shall write $A \implies B$

$A \implies B$ is know as a __conditional__, with __antecedent__ $A$ and __consequent__ $B$

The corresponding __true table__ is

```{r truth_table_conditional}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\implies B$` = c("T", "T", "F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

There is one line of this table which is clear, the third. For, if $A$ is true and $B$ is false, then $\text{If } A \text{ then } B$ is obviously false

However, there is no obvious reason for the entries in the other three rows of the table on the basis of ordinary usage. In that sense, the truth table we have established for $\implies$ is an extension of ordinary usage 

The above __true table__ reflects the meaning that $\implies$ has in contemporary mathematical logic. Two peculiar properties of $\implies$ sometimes cause problem:

- If $A$ is false, then $A \implies B$ is true (no matter what $B$ is)
- If $B$ is true, then $A \implies B$ is true (no matter what $A$ is)

Aside from these peculiar properties of $\implies$ the 2 most important properties of $\implies$ are:

- When $A$ and $A \implies B$ are both true, the $B$ must be true

    + $\{ A, A \implies B \} \vDash B$ [@bilaniuk_problem_2009, Chapter 2, Definition 2.4, p 10]

- In order to prove $A \implies B$, it suffices to assume $A$ true and then show that $B$ must also be true 

    + $\{ A, A \implies B \} \vdash B$ [@bilaniuk_problem_2009, Chapter 3, Definition 3.3, p 12]

### Necessary and sufficient condition

The sentence $\text{The truth of } A \text{ is a necessary condition for the truth of } B$ means that, if $B$ is true, then $A$ also must be true, and, thus, should be translated as $B \implies A$. For example,

- $A$ is $x > 5$
    
- $B$ is $(x > 10)$
    
- A __necesary condition__ for $B$ is $A$
    
    + For $B$ to be true it is necessary for $A$ to be true
        
The sentence $\text{The truth of } A \text{ is a sufficient condition for the truth of } B$ means that, if $A$ is true, $B$ also must be true, and, thus, should be translated as $A \implies B$. For example,

- $A$ is $(x = 2n + 1) \land n \in \mathbb{N} \land \mathbb{N} = \{1, 2, ... \}$
    
- $B$ is $(x^2 = 2m + 1) \land m \in \mathbb{N} \land \mathbb{N} = \{1, 2, ... \}$ 

- A __sufficient condition__ for $A$ is $B$

    + For $A$ to be true it is sufficient for $B$ to be true
    
- For example, assume that $A \cap B \neq \emptyset$

    + __Sufficient condition__ for $x \in A$: $x \in A \cap B$
    
    + __Necessary condition__ for $x \in A \cap B$: $x \in A$
    
    + __Sufficient__ and __necessary condition__ for $x \in A \cap B$: $x \in A \land x \in B$ 

In summary a __necessary condition__ is a condition that must be present in order for another condition to occur and a __sufficient condition__ is a condition that will produce another condition

## Conditionals exercises and solutions

1. Translate into our symbolism the following English sentences

a. If John is not sick and it is not raining, we shall go on a picnic

    + $A = \text{John is not sick}$
    
    + $B = \text{It is raining}$
    
    + $C = \text{We shall go on a picnic}$
    
    + $(A \land \neg B) \implies C = \text{If John is not sick and it is not raining, we shall go on a picnic}$
    
b. Sam goes to work only if the sun is not shinning

    + $A = \text{Sam goes to work}$
    
    + $B = \text{The sun is shinning}$
    
    + $\neg B \implies A = \text{Sam goes to work only if the sun is not shinning}$
    
c. A sufficient condition for $f$ to be continuous is that $f$ is differentiable

    + $A = f \text{ is a continuous function}$
    
    + $B = f \text{ is a differentiable function}$
    
    + $A \implies B = \text{A sufficient condition for } f \text{ to be continuous is that } f \text{ is differentiable}$
    
d. A necessary condition for $s$ to converge is that $s$ is bounded

    + $A = s \text{ converge}$
    
    + $B = s \text{ is bounded}$
    
    + $B \implies A = \text{A necessary condition for } s \text{ to converge is that } s \text{ is bounded}$
    
2. Show that the following sentences always have the same truth values and are, therefore, equivalent

    a. $A \implies B$
    
    b. $(\neg A) \lor B$
    
    c. $\neg (A \land (\neg B))$

```{r}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\implies B$` = c("T", "T", "F", "T"),
       `$\\neg A$` = c("F", "T", "F", "T"),
       `$(\\neg A) \\lor B$` = c("T", "T", "F", "T"),
       `$\\neg B$` = c("F", "F", "T", "T"),
       `$A \\land (\\neg B)$` = c("F", "F", "T", "F"),
       `$\\neg (A \\land (\\neg B))$` = c("T", "T", "F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(3, 5, 8), background = palette_light()[[7]])
```

3. Verify the following justification for our truth table for $A \implies B$. It is obvious that $(C \land D) \implies D$[^1] must always be true. Hence:

    + When $C$ and $D$ are both true we have that $(C \land D) \implies D$ is true. This gives the first row of the table for $\implies$:
    
        + $(T \implies T) = T$
    
    + When $C$ is false and $D$ is true we have that $(C \land D) \implies D$ is true. This gives the second row of the table for $\implies$:
    
        + $(F \implies T) = T$
    
    + When $D$ is false we have that $(C \land D) \implies D$ is true. This gives the fourth row of the table for $\implies$:
    
        + $(F \implies F) = T$

```{r }

tibble(`$C$` = c("T", "F", "T", "F"),
       `$D$` = c("T", "T", "F", "F"),
       `$C \\land D$` = c("T", "F", "F", "F"),
       `$(C \\land D) \\implies D$` = c("T", "F", "T", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```


[^1]: This is a rule of inference and it is know as __conjunction elimination__

4. Show that the following pairs of sentences are equivalent, that is, always take the same truth values:

    a. $\neg (A \implies B)$ and $\neg (A \land (\neg B))$
    
```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\implies B$` = c("T", "T", "F", "T"),
       `$\\neg (A \\implies B)$` = c("F", "F", "T", "F"),
       `$\\neg B$` = c("F", "F", "T", "T"),
       `$A \\land (\\neg B)$` = c("F", "F", "T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(4, 6), background = palette_light()[[7]]) 
```

    b. $A \implies (B \implies C)$ and $(A \land B) \implies C$
    
```{r }

tibble(`$A$` = c("T", "T", "T", "T", "F", "F", "F", "F"),
       `$B$` = c("T", "T", "F", "F", "T", "T", "F", "F"),
       `$C$` = c("T", "F", "T", "F", "T", "F", "T", "F"),
       `$B \\implies C$` = c("T", "F", "T", "T", "T", "F", "T", "T"),
       `$A \\implies (B \\implies C)$` = c("T", "F", "T", "T", "T", "T", "T", "T"),
       `$A \\land B$` = c("T", "T", "F", "F", "F", "F", "F", "F"),
       `$(A \\land B) \\implies C$` = c("T", "F", "T", "T", "T", "T", "T", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(5, 7), background = palette_light()[[7]]) 
```

    c. $A \implies (B \implies C)$ and $(A \land B) \implies C$

```{r}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\implies B$` = c("T", "T", "F", "T"),
       `$\\neg A$` = c("F", "T", "F", "T"),
       `$\\neg B$` = c("F", "F", "T", "T"),
       `$(\\neg B) \\implies (\\neg A)$` = c("T", "T", "F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>%
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")  %>% 
    column_spec(column = c(3, 6), background = palette_light()[[7]])
```

5. _Reductio ad adsurdum_

    a. Show that if $\neg A \implies (B \land (\neg B))$ is true, then $A$ is true. This justifies proofs by _reductio ad adsurdum_
    
        + To prove $A$, show that $\neg A$ leads to a contradiction
        
```{r}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$\\neg A$` = c("F", "T", "F", "T"),
       `$B \\land (\\neg B)$` = c("F", "F", "F", "F"),
       `$\\neg A \\implies (B \\land (\\neg B))$` = c("T", "F", "F", "F")
       ) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>%
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")  %>% 
    row_spec(row = 1, background = palette_light()[[7]])
```

    b. Prove by _reductio ad adsurdum_ that $A \implies A$ is always true
    
Assume that $\neg (A \implies A)$. This is equivalent to $A \land (\neg A)$ by exercise 4a. Therefore $\neg (A \implies A) \implies (A \land (\neg A))$ is true and we conclude by _reductio ad adsurdum_ that $A \implies A$ is true

### Making sense of the __truth table__ for conditionals

#### Possible __truth tables__ for 2 binary variables

- There are 16 possible __truth tables__ for two binary variables: $A, B$
    
```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$\\neg A$` = c("F", "T", "F", "T"),
       `$\\neg B$` = c("F", "F", "T", "T"),
       `$A \\land B$` = c("T", "F", "F", "F"),
       `$A \\lor B$` = c("T", "T", "T", "F"),
       `$A \\oplus B$` = c("F", "T", "T", "F"),
       `$A \\uparrow B$` = c("F", "T", "T", "T"),
       `$A \\downarrow B$` = c("F", "F", "F", "T"),
       `$A \\implies B$` = c("T", "T", "F", "T"),
       `$A \\Longleftarrow B$` = c("T", "F", "T", "T"),
       `$\\neg (A \\implies B)$` = c("F", "F", "T", "F"),
       `$\\neg (A \\Longleftarrow B)$` = c("F", "T", "F", "F"),
       `$A \\Longleftrightarrow B$` = c("T", "F", "F", "T"),
       `$\\top$` = c("T", "T", "T", "T"),
       `$\\bot$` = c("F", "F", "F", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    row_spec(row = 3, background = palette_light()[[7]]) %>% 
    column_spec(column = c(2, 3, 5, 9, 13, 14, 16), background = palette_light()[[7]])
```

- Where

    + $A$ is the case of a __truth function__  that only depends on the binary variable $A$
    
    + $B$ is the case of a __truth function__  that only depends on the binary variable $B$

    + $\neg A$ is the negation of $A$ (also is a __truth function__  that only depends on the binary variable $A$)
    
    + $\neg B$ is the negation of $B$ (also is a __truth function__  that only depends on the binary variable $B$)
    
    + $A \land B$ is a conjunction
    
    + $A \lor B$ is a disjunction
    
    + $A \oplus B$ is a exclusive disjunction and known as __XOR__: $(A \lor B) \land \neg (A \land B)$
    
    + $A \uparrow B$ is known as __NAND__: $\neg (A \land B)$
    
    + $A \downarrow B$ is known as __NOR__: $\neg (A \lor B)$
    
    + $A \implies B$ is know as the material conditional: $\neg (A \land \neg B)$
    
    + $A \Longleftarrow B$ is know as the converse of material conditional: $\neg (B \land \neg A)$
    
    + $\neg (A \implies B)$ is the negation of the material implication
    
    + $\neg (A \Longleftarrow B)$ is the negation of the converse of the material implication
    
    + $A \Longleftrightarrow B$ is a biconditional
    
    + $\top$ is a tautology
    
    + $\bot$ is a contradiction
    
#### Alternative __truth tables__ for $\implies$

- The columns colored in green from the above table are the alternative __truth tables__ for $A \implies B$

    + This columns are cases where $A$  is true, $B$ is false and the value of the __truth table__ is false  

        + We discard $B$ and $\neg A$ because these tables only depend on one binary variable
    
        + We discard $\bot$ because is a contradiction
    
        + We discard $A \land B$, and $A \Longleftrightarrow B$ becasue are logical connectives with a clear interpretation in mathematical logic
    
        + The remaining candidates are $A \downarrow B$ and $\neg (A \Longleftarrow B)$
        
            + The problem with this logical connectives as alternative candidates for $A \implies B$ is that when $A$ and $B$ are true the value of the __truth table__ is false  

- Therefore for 2 binary variables, we may conclude that $A \implies B$ doesn't have other suitable alternatives assuming that we want

    + $A$ true, $B$ false and false for the value of the __truth table__
    
    + $A$ true, $B$ true and true for the value of the __truth table__
    
## Biconditionals

Given sentences $A$ and $B$, we may form a new sentence $\text{If } A \text{ and only if } B$, which we shall write $A \Longleftrightarrow B$

The corresponding __true table__ is

```{r truth_table_biconditional}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\Longleftrightarrow B$` = c("T", "F", "F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```
    
## Biconditionals exercises and solutions

1. Translate into our symbolism:

    a. A necessary and suficient condition for $s$ to be convergent is that it is cauchy convergent
    
        + $A = s \text{ is convergent}$
        
        + $B = s \text{ is cauchy convergent}$
        
        + $A \Longleftrightarrow B = \text{A necessary and suficient condition for } s \text{ to be convergent is that it is cauchy convergent}$ 
        
    b. Horatio is happy if and only if the sun is shining and business is good
    
        + $A = \text{Horatio is happy}$
        
        + $B = \text{The sun is shining}$
        
        + $C = \text{Business is good}$
        
        + $A \Longleftrightarrow (B \land C) = \text{Horatio is happy if and only if the sun is shining and business is good}$
        
    c. $f$ is differentiable when and only when it is continuous
    
        + $A = f \text{ is a differentiable function}$
        
        + $B = f \text{ is a continuous function}$
        
        + $A \Longleftrightarrow B = f \text{ is differentiable when and only when it is continuous}$
        
2. Show that $A \Longleftrightarrow B$ always has the same truth values as $(A \implies B) \land (B \implies A)$. Hence to prove $A \Longleftrightarrow B$ it suffices to prove $A \implies B$ and $B \implies A$

```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\Longrightarrow B$` = c("T", "T", "F", "T"),
       `$B \\Longrightarrow A$` = c("T", "F", "T", "T"),
       `$(A \\Longrightarrow B) \\land (B \\Longrightarrow A)$` = c("T", "F", "F", "T"),
       `$A \\Longleftrightarrow B$` = c("T", "F", "F", "T")
       ) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")  %>% 
    column_spec(column = c(5, 6), background = palette_light()[[7]])
```
  
3. Assume that $A \lor B$ is true. What, if anything, can be said of the truth table values of the following?

    a. $A$
    
        + Nothing because $A$ can be true or false
        
    b. $A \implies B$
    
        + Nothing because $A$ can be true and $B$ false or $B$ can be true
        
    c. $\neg A \implies B$
    
        + $\neg A \implies B$ is true because $B$ is true or $A$ is false and $B is false$
        
    d. $\neg ((\neg A) \land (\neg B))$ 
    
        + $\neg A \implies B$ is true because $\neg A$ is false and $\neg B$ is false or $\neg A$ and $\neg B$ have different truth values
        
    e. $(A \land C) \lor (B \land (\neg C))$ 
    
        + Nothing because $A$ and $B$ can have the same true values
        
```{r }

tibble(`$A$` = c("T", "F", "T"),
       `$B$` = c("T", "T", "F"),
       `$A \\lor B$` = c("T", "T", "T"),
       `$A \\implies B$` = c("T", "T", "F"),
       `$\\neg A \\implies B$` = c("T", "T", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

```{r }

tibble(`$A$` = c("T", "T", "F", "F", "T", "T"),
       `$B$` = c("T", "T", "T", "T", "F", "F"),
       `$C$` = c("T", "F", "T", "F", "T", "F"),
       `$A \\land C$` = c("T", "F", "F", "F", "T", "F"),
       `$B \\land \\neg C$` = c("F", "T", "F", "T", "T", "F"),
       `$(A \\land C) \\lor (B \\land (\\neg C))$` = c("T", "T", "T", "T", "T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

4. Assume that $A \implies B$ is true. What, if anything, can be said of the truth values of the following?

    a. $B$
    
        + Nothing because $B$ can be true or false

    b. $A \lor B$
    
        + Nothing because $A$ and $B$ can have the same true values
        
    c. $(\neg A) \lor B$
    
        + $(\neg A) \lor B$ is true because $\neg A$ and $\neg B$ have different true values or $\neg A$ and $\neg B$ are false

    d. $B \implies A$
    
        + Nothing because $B$ is true and $A$ is false or $B$ is false or $A$ is true
        
    e. $(\neg B) \implies (\neg A)$
    
        + $(\neg B) \implies (\neg A)$ is true because it is not the case that $\neg B$ is true and $\neg A$ is false 

```{r }

tibble(`$A$` = c("T", "F", "F"),
       `$B$` = c("T", "T", "F"),
       `$A \\implies B$` = c("T", "T", "T"),
       `$A \\lor B$` = c("T", "T", "F"),
       `$\\neg A \\lor B$` = c("T", "T", "T"),
       `$B \\implies A$` = c("T", "F", "T"),
       `$\\neg B \\implies \\neg A$` = c("T", "T", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

5. Assume that $A$ is true, $B$ is false, and $C$ is true. What are the truth values of:

    a. $A \implies B$
    
        + False
        
    b. $B \implies C$
    
        + True
        
    c. $(A \lor \neg C) \lor B$
    
        + True
        
    d. $(A \Longleftrightarrow C) \Longleftrightarrow \neg B$
    
        + True
        
    e. $A \implies (B \implies A)$
    
        + True

6. For each of the following pairs of sentences show that they are equivalent, in the sense that they always take the same truth values

    a. $A \land (B \lor \neg A)$ and $A \land B$
    
        + $A \land (B \lor \neg A)$ is equivalent to $(A \land B) \lor (A \land \neg A)$ based on exercise 3d in section 1.1 Logical connectives
        
        + We have that $A \land \neg A$ is always false
        
        + Because $A \land B$ can be false of true we have that if $A \land B$ is false $(A \land B) \lor (A \land \neg A)$ is false. Also if $A \land B$ is true $(A \land B) \lor (A \land \neg A)$ is true
        
        + Therefore $A \land (B \lor \neg A)$ is equivalent to $A \land B$ 

    b. $A \lor (B \land \neg A)$ and $A \lor B$
    
        + $A \lor (B \land \neg A)$ is equivalent to $(A \lor B) \land (A \lor \neg A)$ based on exercise 3e in section 1.1 Logical connectives
        
        + We have that $A \lor \neg A$ is always true
        
        + Because $A \land B$ can be false of true we have that if $A \lor B$ is false $(A \lor B) \land (A \lor \neg A)$ is false. Also if $A \land B$ is true $(A \lor B) \land (A \lor \neg A)$ is true
        
        + Therefore $A \lor (B \land \neg A)$ is equivalent to $A \lor B$
        
    c. $A \lor (B \land A)$ and $A$
    
        + If $A$ is false we have that $B \land A$ is false therefore $A \lor (B \land A)$ is false
        
        + If $A$ is true we have that $B \land A$ can be true or false. However $A \lor (B \land A)$ is true
        
        + Therefore $A \lor (B \land A)$ is equivalent to $A$
        
    d. $A \land (B \lor A)$ and $A$
    
        + If $A$ is false we have that $B \lor A$ can be true of false. However $A \land (B \lor A)$ is false 
        
        + If $A$ is true we have that $B \lor A$ is true. Therefore $A \land (B \lor A)$ is true
        
        + Therefore $A \land (B \lor A)$ is equivalent to $A$
        
    e. $\neg (A \lor B)$ and $(\neg A) \land (\neg B)$
    
        + See exercise 3a in section 1.1 Logical connectives
        
    f. $\neg (A \land B)$ and $(\neg A) \lor (\neg B)$
    
        + See exercise 3b in section 1.1 Logical connectives
        
## Quantifiers

### Universal quantifier

The sentence

$$\text{All men are mortal } (1)$$ 

can be formulated as

$$\text{For all } x\text{, if } x \text{ is a man, then } x \text{ is mortal } (2)$$

If we let $M(x)$ stand for $x \text{ is a man}$ and $D(x)$ for $x \text{ is mortal}$ then we can rewritten $(3)$ as

$$\text{For all } x \; (M(x) \implies D(x)) \; (3)$$
Finally we abbreviate $\text{For all } x$ by $(\forall x)$ rewriting $(3)$ as

$$\forall x (M(x) \implies D(x)) \; (4)$$
The expression $(\forall x)$ is called a __universal quantifier__. It usually replaces the english expressions _for all x_, _for every x_, _for each x_, _for any x_. Therefore any english of the sentence of the form _All A's are B's_ can be written in the form $(\forall x)(A(x) \implies B(x))$

### Existential quantifier

There is another kind of quantifier occurring often in mathematics. Consider the sentence

$$\text{Some continuous functions are not differentiable } (1)$$
This can be written as 

$$\text{For some } f, f \text{ is continuous function and } f \text{ is not a differentiable function } (2)$$
This can be formulated as

$$\text{For some } f, \; C(f) \land \neg D(f) \; (3)$$
where $C(f)$ stands for $f \text{ is continuous function}$ and $D(f)$ stands for $f \text{ is a differentiable function }$ 

Finally, we introduce $(\exists f)$ as an abbreviation for $\text{for some } f$, obtaining

$$(\exists f)(C(f) \land \neg D(f)) \; (4)$$
The expression $(\exists x)$ is called a __existential quantifier__. It usually replaces the english expressions _for some f_, _there exists and f such that_, _for at least one f_. Therefore any english of the sentence of the form _Some A's are B's_ can be written in the form $(\exists f)(A(x) \implies B(x))$

For example, the sentence

$$\text{There is no largest integer}$$
can be written in the following ways where $I(X)$ means _x is an integer_

$$(\neg (\exists x))(I(x) \land (\forall y)(I(y) \implies y \leq x))$$

$$(\forall x)(I(x) \implies (\exists y)(I(y) \land x < y))$$
Sometimes it is convenient to apply a special uniqueness symbol:

- $(\exists !x)(A(x))$ shall mean _there is one and only one x such that A(x)_

- Equivalent english sentences are _there is a unique x such that A(x)_, _there is exactly one x such that A(x)_, _there is precisely one x such that A(x)_

- We can define $(\exists !x)(A(x))$ as

$$(\exists x)(A(x) \land (\forall z)(A(z) \implies z = x))$$

### Relation between the universal and existential quantifier

Notice that $\neg (\exists x)(A(x))$ has the same meaning as $(\forall x)(\neg A(x))$

For example, the sentence

$$\text{No integer is bigger than itself}$$
can be written in the following ways

$$\neg (\exists x)(I(x) \land x < x)$$
$$(\forall x)(\neg (I(x) \land x < x))$$
$$(\forall x)(I(x) \implies \neg(x < x)))$$

where $\neg (A \land B)$ is equivalent to $\neg (A \land \neg (\neg B))$. So it is also equivalent to $A \implies \neg B$   

For example, the sentence

$$\text{Not all primes are odd}$$
can be written in the following ways where $P(x)$ means _x is prime_ and $O(x)$ means _x is odd_

$$(\exists x)(P(x) \land \neg O(x))$$
$$(\exists x)(\neg \neg(P(x) \land \neg O(x)))$$
$$(\exists x)(\neg (P(x) \implies O(x)))$$
$$\neg (\forall x)(P(x) \implies O(x))$$

## Quantifiers exercises and solutions

1. Translate the following english sentences into our symbolism

    a. Not all american girls are beautiful
    
        + $A(x) = x \text{ is american}$
        
        + $G(x) = x \text{ is a girl}$
        
        + $B(x) = x \text{ is beautiful}$
        
        + $(\exists x)((A(x) \land G(x)) \land \neg B(x)) = \text{Not all american girls are beautiful}$
        
        + $(\exists x)\neg (A(x) \land G(x)) \implies B(x)) = \text{Not all american girls are beautiful}$
        
        + $\neg (\forall x) (A(x) \land G(x)) \implies B(x)) = \text{Not all american girls are beautiful}$
        
    b. All american girs are not beautiful
    
        + $A(x) = x \text{ is american}$
        
        + $G(x) = x \text{ is a girl}$
        
        + $B(x) = x \text{ is beautiful}$
        
        + $\neg (\exists x)((A(x) \land G(x)) \land B(x)) = \text{All american girs are not beautiful}$
        
        + $(\forall x) \neg ((A(x) \land G(x)) \land B(x)) = \text{All american girs are not beautiful}$
        
        + $(\forall x) \neg ((A(x) \land G(x)) \land \neg \neg B(x)) = \text{All american girs are not beautiful}$
        
        + $(\forall x) (A(x) \land G(x)) \implies \neg B(x)) = \text{All american girs are not beautiful}$
        
    c. There is a smallest positive integer
    
        + $P(x) = x \text{ is a positive integer} = x \in \mathbb{P}$
        
        + $S(x) = x \text{ is small or equal to } y = x \leq y$
        
        + $(\forall x)(x \in \mathbb{P} \implies (\exists y)(y \in \mathbb{P} \land y \leq x))$
        
    d. There is a rational number between any two real numbers
    
        + $Q(x) = x \text{ is a rational number} = x \in \mathbb{Q}$
        
        + $R(x) = x \text{ is a real number} = x \in \mathbb{R}$
        
        + $B(x,y,z) = x \text{ is between } y \text{ and } z = y < x < z$
        
        + $(\forall y)(\forall z)((y \in \mathbb{R} \land z \in \mathbb{R} \land y < z) \implies (\exists x)(x \in \mathbb{Q} \land y < x < z))$

# Bibliography
















