---
title: "Basic facts and notions of logic and set theory"
date: 2020-04-11 18:49:34
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}

library(tidyverse)
library(kableExtra)
```

# Logic and set theory

This section is based on [@mendelson_number_2008, Chapter 1, pp 1-52]

This is a very brief introduction to the basic concepts and terminology of mathematical logic and set theory

For another introduction to set theory see [@halmos_naive_1974] and for an extensive survey of the set theory see [@jech_set_2003]

For another introduction to mathematical logic see [@smullyan_beginners_2014] and for an extensive survey of mathematical logic see [@bilaniuk_problem_2009]

## Logical connectives

### Conjuction

Given sentences $A$ and $B$, we may form a new sentence $A \text{ and } B$, which we shall write $A \land B$

The relation between the new sentence $A \land B$ and the component sentences $A,B$ can be made clear by a __truth table__:

```{r truth_table_conjuction}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\land B$` = c("T", "F", "F", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

In this table, as in all tables to follow, T stands for __True__ and F for __False__

### Negation

If $A$ is a sentence, the $\text{not } A$ is also a sentence, which we shall denote $\neg A$

$\neg A$ is called the __negation__ of $A$

The corresponding __truth table__ is:

```{r truth_table_negation}

tibble(`$A$` = c("T", "F"),
       `$\\neg A$` = c("F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

### Disjuction

Given sentences $A$ and $B$, we may form a new sentence $A \text{ or } B$, which we shall write $A \lor B$

$\lor$ is understood in the __inclusive sense__, namely, $A \lor B$ is true when $A$ is true or $B$ is true or both

The corresponding __true table__ is

```{r truth_table_disjuction}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\lor B$` = c("T", "T", "T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

If a set of formulas take the same truth values we said that these formulas are __equivalent__. For example:

- $A \lor A$ and $A$

    + When $A$ is T, so is $A \lor A$, and viceversa

- $A \land A$ and $A$

    + $A$ is T if and only if $A \land A$ is T
    
- $A \lor (B \land A)$ and $A$

    + When $A$ is T, so is $A \lor (B \land A)$. If $A$ is F, then $A \lor (B \land A)$ is F. Using a __true table__ we can see that:
    
```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\land B$` = c("T", "F", "F", "F"),
       `$A \\lor (A \\land B)$` = c("T", "F", "T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

## Logical connectives exercises



# Bibliography
