---
title: "Basic facts and notions of logic and set theory"
date: 2020-04-11 18:49:34
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}

library(tidyverse)
library(kableExtra)
library(tidyquant)
```

# Logic and set theory

This section is based on [@mendelson_number_2008, Chapter 1, pp 1-52]

This is a very brief introduction to the basic concepts and terminology of mathematical logic and set theory

For another introduction to set theory see [@halmos_naive_1974] and for an extensive survey of the set theory see [@jech_set_2003]

For another introduction to mathematical logic see [@smullyan_beginners_2014] and for an extensive survey of mathematical logic see [@bilaniuk_problem_2009]

## Logical connectives

### Conjuction

Given sentences $A$ and $B$, we may form a new sentence $A \text{ and } B$, which we shall write $A \land B$

The relation between the new sentence $A \land B$ and the component sentences $A,B$ can be made clear by a __truth table__:

```{r truth_table_conjuction}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\land B$` = c("T", "F", "F", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

In this table, as in all tables to follow, T stands for __True__ and F for __False__

### Negation

If $A$ is a sentence, the $\text{not } A$ is also a sentence, which we shall denote $\neg A$

$\neg A$ is called the __negation__ of $A$

The corresponding __truth table__ is:

```{r truth_table_negation}

tibble(`$A$` = c("T", "F"),
       `$\\neg A$` = c("F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

### Disjuction

Given sentences $A$ and $B$, we may form a new sentence $A \text{ or } B$, which we shall write $A \lor B$

$\lor$ is understood in the __inclusive sense__, namely, $A \lor B$ is true when $A$ is true or $B$ is true or both

The corresponding __true table__ is

```{r truth_table_disjuction}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\lor B$` = c("T", "T", "T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

If a set of formulas take the same truth values we said that these formulas are __equivalent__. For example:

- $A \lor A$ and $A$

    + When $A$ is T, so is $A \lor A$, and viceversa

- $A \land A$ and $A$

    + $A$ is T if and only if $A \land A$ is T
    
- $A \lor (B \land A)$ and $A$

    + When $A$ is T, so is $A \lor (B \land A)$. If $A$ is F, then $A \lor (B \land A)$ is F. Using a __true table__ we can see that:
    
```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\land B$` = c("T", "F", "F", "F"),
       `$A \\lor (A \\land B)$` = c("T", "F", "T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

## Logical connectives exercises and solutions

1. Write the following English sentences in terms of our new notation. Use letters $A, B, C,...$ to stand for the basic sentences

    a. I will go to the movies or to the beach
    
        + $A = \text{I will go to the movies}$
        
        + $B = \text{I will go to the beach}$
        
        + $A \lor B = \text{I will go to the movies or to the beach}$
        
    b. Peter is handsome but Paul is not
    
        + $A = \text{Peter is handsome}$
        
        + $B = \text{Paul is handsome}$
        
        + $A \land \neg B = \text{Peter is handsome but Paul is not}$
        
    c. Either $x$ has a square root or $x$ is not positive
    
        + $A = x \text{ has a square root}$
        
        + $B = x \text{ is positive}$
        
        + $A \lor \neg B = \text{Either } x \text{ has a square root or } x \text{ is not positive}$

    d. It is not the case that $x$ is both positive and negative
    
        + $A = x \text{ is positive}$
        
        + $B = x \text{ is negative}$
        
        + $\neg(A \land B) = \text{It is not the case that } x \text{ is both positive and negative}$
        
    e. Neither John not James is here
    
        + $A = \text{John is here}$
        
        + $B = \text{James is here}$
        
        + $\neg(A \land B) = \text{Neither John not James is here}$
        
2. Render the following into idiomatic English

    a. $x > 2 \lor x < 2$
    
        + $x$ is greater than or less than $2$
        
    b. $\neg (y \text{ is rational } \lor y \text{ is trascendental})$
    
        + Neither $y$ is rational or transcendental
        
    c. $e > 2 \land e < 3$
    
        + $e$ is greater than $2$ and less than $3$
    
    d. $e > 2 \land e \ngtr 3$
    
        + $e$ is greater than $2$ but not greater than $3$
        
3. Show that the following pairs of formulas always take the same truth values, and, therefore, are, in a certain sense, equivalent:

    a. $\neg(A \lor B)$ and $(\neg A) \land (\neg B)$
    
```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\lor B$` = c("T", "T", "T", "F"),
       `$\\neg (A \\lor B)$` = c("F", "F", "F", "T"),
       `$(\\neg A)$` = c("F", "T", "F", "T"),
       `$(\\neg B)$` = c("F", "F", "T", "T"),
       `$(\\neg A) \\land (\\neg B)$` = c("F", "F", "F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(4, 7), background = palette_light()[[7]])
```

    b. $\neg(A \land B)$ and $(\neg A) \lor (\neg B)$

```{r }

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\land B$` = c("T", "F", "F", "F"),
       `$\\neg (A \\land B)$` = c("F", "T", "T", "T"),
       `$(\\neg A)$` = c("F", "T", "F", "T"),
       `$(\\neg B)$` = c("F", "F", "T", "T"),
       `$(\\neg A) \\lor (\\neg B)$` = c("F", "T", "T", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(4, 7), background = palette_light()[[7]])
```
   
    c. $\neg \neg B$ and $B$

```{r }

tibble(`$B$` = c("T", "F"),
       `$\\neg B$` = c("F", "T"),
       `$\\neg \\neg B$` = c("T", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(1, 3), background = palette_light()[[7]])
```

    d. $A \land (B \lor C)$ and $(A \land B) \lor (A \land C)$

```{r }

tibble(`$A$` = c("T", "T", "T", "T", "F", "F", "F", "F"),
       `$B$` = c("T", "T", "F", "F", "T", "T", "F", "F"),
       `$C$` = c("T", "F", "T", "F", "T", "F", "T", "F"),
       `$B \\lor C$` = c("T", "T", "T", "F", "T", "T", "T", "F"),
       `$A \\land (B \\lor C)$` = c("T", "T", "T", "F", "F", "F", "F", "F"),
       `$A \\land B$` = c("T", "T", "F", "F", "F", "F", "F", "F"),
       `$A \\land C$` = c("T", "F", "T", "F", "F", "F", "F", "F"),
       `$(A \\land B) \\lor (A \\land C)$` = c("T", "T", "T", "F", "F", "F", "F", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(5, 8), background = palette_light()[[7]])
```

    e. $A \lor (B \land C)$ and $(A \lor B) \land (A \lor C)$

```{r }

tibble(`$A$` = c("T", "T", "T", "T", "F", "F", "F", "F"),
       `$B$` = c("T", "T", "F", "F", "T", "T", "F", "F"),
       `$C$` = c("T", "F", "T", "F", "T", "F", "T", "F"),
       `$B \\land C$` = c("T", "F", "F", "F", "T", "F", "F", "F"),
       `$A \\lor (B \\land C)$` = c("T", "T", "T", "T", "T", "F", "F", "F"),
       `$A \\lor B$` = c("T", "T", "T", "T", "T", "T", "F", "F"),
       `$A \\lor C$` = c("T", "T", "T", "T", "T", "F", "T", "F"),
       `$(A \\lor B) \\land (A \\lor C)$` = c("T", "T", "T", "T", "T", "F", "F", "F")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered") %>% 
    column_spec(column = c(5, 8), background = palette_light()[[7]])
```

## Conditionals

Given sentences $A$ and $B$, we may form a new sentence $\text{If } A \text{ then } B$, which we shall write $A \implies B$

$A \implies B$ is know as a __conditional__, with __antecedent__ $A$ and __consequent__ $B$

The corresponding __true table__ is

```{r truth_table_conditional}

tibble(`$A$` = c("T", "F", "T", "F"),
       `$B$` = c("T", "T", "F", "F"),
       `$A \\implies B$` = c("T", "T", "F", "T")) %>% 
    kable(format = "html", escape = FALSE, align = "c") %>% 
    kable_styling(full_width = FALSE, bootstrap_options = "bordered")
```

There is one line of this table which is clear, the third. For, if $A$ is true and $B$ is false, then $\text{If } A \text{ then } B$ is obviously false

However, there is no obvious reason for the entries in the other three rows of the table on the basis of ordinary usage. In that sense, the truth table we have established for $\implies$ is an extension of ordinary usage 

The above __true table__ reflects the meaning that $\implies$ has in contemporary mathematical logic. Two peculiar properties of $\implies$ sometimes cause problem:

- If $A$ is false, then $A \implies B$ is true (no matter what $B$ is)
- If $B$ is true, then $A \implies B$ is true (no matter what $A$ is)

# Bibliography
