---
title: "The natural numbers"
date: 2020-06-02 23:02:09
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}


```

# The natural numbers

This section is based on [@mendelson_number_2008, Chapter 2, pp 53-85]

- We define the natural numbers as the positive integers, $\mathbb{P} = \{1, 2, 3, \ldots \}$

- In this part we present an axiomatic development of the natural numbers

## Peano systems

By a __peano system__ we mean

- A set $P$

- A particular element $1 \in P$

- A singularity operation $S$ on $P$ $(S: P \longrightarrow P) \land \mathscr{D}(S) = P$

Such that the following axioms are satisfied:

- $(P1) \; 1$ is not the _successor_ $S(x)$ of any object $x \in P$:

    + $(\forall x)(x \in P \implies S(x) \neq 1)$
    
        + In that sense $1 \notin \mathscr{R}(S)$ so $\mathscr{R}(S) \subsetneq P$

- $(P2)$ Different objects in $P$ have different successors:

    + $(\forall x)(\forall y)((x \in P \land y \in P \land x \neq y) \implies S(x) \neq S(y))$
    
        + $S$ is one-one $(S: P \xrightarrow[]{\text{1-1}} P)$

- $(P3)$ _Principle of Mathematical Induction_: Any subset of $P$ containing $1$ and closed under the successor operation must be identical with $P$

    + $(\forall B)((B \subseteq P \land 1 \in B \land (\forall x)(x \in B \implies S(x) \in P)) \implies B = P)$

Such a peano system will be denoted by the ordered triple $(P, S, 1)$

- $P$ is called the __underlying set__

- $S$ the __successor operation__

- $1$ the __distinguished element__

    + The distinguished element $1$ need not to have anything to do with the ordinary integer 1

### Example of peano systems

- Let $P = \mathbb{P}$, $1$ the ordinary integer $1$ and $S$ equal to $S(x) = x + 1$ for all $x \in \mathbb{P}$

    + $S(x) = x + 1 \neq 1$
    
    + If $x \neq y$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \mathbb{P}$, $1 \in B$ and assume that if $x \in B$ then $x + 1 \in B$
    
        + If $1 \in B$ then $1 + 1 = 2 \in B$, $2 + 1  = 3 \in B$ and so on, where any $n \in \mathbb{P}$ can be reach
        
- Let $P = \mathbb{N}$ where $\mathbb{N} = \mathbb{P} \cup \{ 0 \}$, $1$ the ordinary integer $0$ and $S$ equal to $S(x) = x + 1$ for all $x \in \mathbb{N}$ 

    + $S(x) = x + 1 \neq 0$
    
    + If $x \neq y$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \mathbb{N}$, $0 \in B$ and assume that if $x \in B$ then $x + 1 \in B$
    
        + If $0 \in B$ then $0 + 1 = 1 \in B$, $1 + 1  = 2 \in B$ and so on, where any $n \in \mathbb{N}$ can be reach

- Let $P = \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$, $1$ the ordinary integer $1000$ and $S$ equal to $S(x) = x + 1$ for all $x \in \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$

    + $S(x) = x + 1 \neq 1000$
    
    + If $x \neq y$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$, $1000 \in B$ and assume that if $x \in B$ then $x + 1 \in B$
    
        + If $1000 \in B$ then $1000 + 1 = 1001 \in B$, $1001 + 1  = 1002 \in B$ and so on, where any $x \in \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$ can be reach
        
- Let $P = \mathbb{Z}^-$, $1$ the ordinary integer $-1$ and $S$ equal to $S(x) = x - 1$ for all $x \in \mathbb{Z}^-$

    + $S(x) = x - 1 \neq -1$
    
    + If $x \neq y$ then $x - 1 \neq y - 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \mathbb{Z}^-$, $-1 \in B$ and assume that if $x \in B$ then $x - 1 \in B$
    
        + If $-1 \in B$ then $-1 - 1 = -2 \in B$, $-2 -1 = -3 \in B$ and so on, where any $z \in \mathbb{Z}^-$ can be reach
        
- Let $P = \{z: z = 2n \land n \in \mathbb{P} \}$, $1$ the ordinary integer $2$ and $S$ equal to $S(x) = x + 2$ for all $x \in \{z: z = 2n \land n \in \mathbb{P} \}$

    + $S(x) = x + 2 \neq 2$
    
    + If $x \neq y$ then $x + 2 \neq y + 2$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \{z: z = 2n \land n \in \mathbb{P} \}$, $2 \in B$ and assume that if $x \in B$ then $x + 2 \in B$
    
        + If $2 \in B$ then $2 + 2 = 4 \in B$, $4 + 2 = 6 \in B$ and so on, where any $x \in \{z: z = 2n \land n \in \mathbb{P} \}$ can be reach

__Theorem__ Every element different from $1$ and belonging to $P$ is a successor $$(\forall x)((x \in P \land x \neq 1) \implies (\exists y)(y \in P \land x = S(y)))$$ 

Let $B = \{x: x = 1 \lor (\exists y)(x = S(y)) \}$

- $1 \in B$

- If $x \in B$ then $x = 1 \lor x = S(y)$ where $S(y) \neq 1 \land S(y) \in P$

    + Because $1 \in P \land S(y) \in P$ then $B \subseteq P$

- $\begin{split}
    x \in B \land x \neq 1 & \implies x = S(y) \\
    & \implies S(x) = S(S(y)) \\
    & \implies S(x) \in B
    \end{split}$
    
    + Therefore by the principle of mathematical induction $B = P$

__Theorem__ No object is its own successor, that is, $(\forall x)(x \in P \land S(x) \neq x)$

Let $B = \{x: x \in P \land S(x) \neq x \}$

- $1 \in B$ because $1 \in P \land S(x) \neq 1$
    
- If $x \in B$ then $x \in P$ so $B \subseteq P$
    
- $\begin{split}
    x \in B \land & \implies S(x) \neq x \land S(x) \in P \land x \in P \\
    & \implies S(S(x)) \neq S(x) \land S(x) \in P \\
    & \implies S(x) \in B
    \end{split}$
    
    + Therefore by the principle of mathematical induction $B = P$
    
## Peano systems exercises and solutions

1. Prove that every object different from $1$ is the successor of a unique object, that is, $$(\forall x)(x \in P \implies (x = 1 \lor (\exists !y)(y \in P \land x = S(y))))$$

- If $1 \in P$ then $1 = 1$

- If $x \in P \land x \neq 1$ then $(\exists y)(y \in P \land x = S(y))))$

    + Assume that there is another $z$ such that $z \in P \land x = S(z)$
    
        + Therefore $S(y) = S(z) \implies y = z$ because $S$ is a one-one function
        
            + Thus if $x \in P \land x \neq 1$ then $(\exists !y)(y \in P \land x = S(y))))$
            
2. Prove $(\forall x)(x \in P \implies (x = 1 \lor x = S(1) \lor x = S(S(y))))$

- If $1 \in P$ then $1 = 1$

- If $x \in P \land x \neq y$ then $(\exists !z)(z \in P \land x = S(z))$

    + If $z = 1$ then $x = S(1)$
    
    + If $z \neq 1$ then $(\exists !y)(y \in P \land z = S(y))$. Therefore $(\exists !y)(y \in P \land x = S(S(y)))$
    
3. Determine whether or not the following structures $(P, S, 1)$ are peano systems

    + $P$ is the set of all integers greater than $9$, $1$ stands for the integer $10$ and $S(u) = u + 1$ for any $u \in P$
    
        + $P = \{u: u \in \mathbb{Z} \land x > 9 \}$
        
            + $(\forall u)(u \in P \implies u + 1 \neq 10)$
            
            + $(\forall u)(\forall v)((u \in P \land v \in P \land u \neq v) \implies u + 1 \neq v + 1)$
            
            + Assume that $B \subseteq P$, $10 \in B$ and $(\forall u)(u \in B \implies u + 1 \in B)$
                + $10 \in P$, $10 + 1 = 11 \in P$, $11 + 1 = 12 \in P$ and so on where any $u \in P$ can be reach
                
    + $P$ is the set of all integers, $1$ stands for the ordinary integer $1$ and $S(u) = u + 1$ for all $u$ in $P$
    
        + $P = \mathbb{Z}$
        
            + $0 \in \mathbb{Z} \land S(0) = 0 + 1 = 1$. Therefore $S(0) = 1$
        
    + $P$ is the set of all integers, $1$ stands for the ordinary integer $0$ and $$S(u) = \begin{cases}
    1        & \text{if } u = 0 \\
    -u       & \text{if } u > 0 \\
    -(u - 1) & \text{if } u < 0
    \end{cases}$$
    
        + $P = \mathbb{Z}$
            
            + $S(u) = 1 \lor S(u) < 0 \lor S(u) > 0$ so $S(x) \neq 0$
            
            + Let $u \in P \land v \in P \land u \neq v$
            
                + If $u = 0$ then $S(1) = 0 \land (v > 0  \lor v < 0)$
                
                    + If $v > 0$ then $S(v) = -v < 0$ so $S(v) \neq S(u)$
                    
                    + If $v < 0$ then $S(v) = -(v-1) > 1$ so $S(v) \neq S(u)$
                    
                + If $u > 0$ then $S(u) = -u < 0 \land (v = 0  \lor v < 0)$
                
                    + If $v = 0$ then $S(v) = 1$ so $S(v) \neq S(u)$
                    
                    + If $v < 0$ then $S(v) = -(v-1) > 1$ so $S(v) \neq S(u)$
                    
                + If $u < 0$ then $S(u) = -(u - 1) > 1 \land (v = 0  \lor v > 0)$
                
                    + If $v = 0$ then $S(v) = 1$ so $S(v) \neq S(u)$
                    
                    + If $v < 0$ then $S(v) = -(v-1) > 1$ so $S(v) \neq S(u)$
                    
            + Assume that $B \subseteq P$, $0 \in B$ and $(\forall u)(u \in B \implies S(u)  \in B)$ where $$S(u) = \begin{cases}
    1        & \text{if } u = 0 \\
    -u       & \text{if } u > 0 \\
    -(u - 1) & \text{if } u < 0
    \end{cases}$$
    
                + $0 \in B$, $S(0) = 1 > 0 \land 1 \in P$, $S(1) = -1 < 0 \land -1 \in P$,  $S(-1) = -(-1 - 1) = 2 > 0 \land 2 \in P$, $S(2) = -2 < 0 \land -2 \in P$ and so on where any $u \in P$ can be reach
        
    + $P$ is the set of all rational numbers of the form $\frac{1}{2^n}$ where $n$ is a nonnegative integer, $1$ stands for the integer $1$ and $S(u) = \frac{1}{2}u$ for all $u \in P$
    
        + $P = \{u: u = \frac{1}{2^n} \land n \in \mathbb{Z^+}\}$
        
            + $1 \in P$ because $0 \in \mathbb{Z^+}$ and $\frac{1}{2^0} = \frac{1}{1} = 1$
            
            + $S(u) = \frac{1}{2}u$ where $u = \frac{1}{2^n}$ and $n \in \mathbb{Z^+}$. Therefore $S(u) = \frac{1}{2} \frac{1}{2^n} = \frac{1}{2^{n + 1}} \neq 1$
            
            + Assume that $B \subseteq P$, $1 \in B$ and $(\forall u)(u \in B \implies \frac{1}{2}u \in B)$
            
                + $1 \in B$, $\frac{1}{2} * 1 = \frac{1}{2} = \frac{1}{2^1}$, $\frac{1}{2} \frac{1}{2} = \frac{1}{2^2}$ and so on where any $u \in P$ can be reach
                
    + $P = \{ 1,2,3,4 \}$, $1$ stands for the ordinary integer $1$ and $S(1) = 2, S(2) = 3, S(4) = 1$
      
        + $4 \in P \land S(4) = 1$ so it is not true that for all $x \in P$ then $S(x) \neq 1$
          
    + $P = \{ 1,2,3, \ldots \} \cup \{ \#1,\#2,\#3, \ldots \}$, where $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers. Let $1$ denote the integer $1$. Let $S(n) = n + 1$ for any natural number $n$ and let $S(\#n) = (n + 1)\#$ for any natural number $n$. Thus, $S(1) = 2, S(2) = 3, \ldots$ and $S(\#1) = \#2, S(\#2) = \#3, \ldots$
 
        + If $n \in P$ then $n = m \in \{ 1,2,3, \ldots \}$ or $n = \#m \in \{ \#1,\#2,\#3, \ldots \}$ but not both because $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers
        
            + If $n = m$ then $S(m) = m + 1 \neq 1$
            
            + If $n = \#m$ then $S(\#m) = (m + 1)\# \neq 1$ because $1 \in \{ 1,2,3, \ldots \}$ and $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers
            
        + Let $x \in P \land y \in P \land x \neq y$
        
            + If $x \neq y$ then $y \in \{ 1,2,3, \ldots \}$ or $y \in \{ \#1,\#2,\#3, \ldots \}$
            
                + If $y \in \{ 1,2,3, \ldots \}$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
            
                + If $y \in \{ \#1,\#2,\#3, \ldots \}$ then $S(x) = x + 1 \land S(y) = (n + 1)\#$ for some $n \in \{ 1,2,3, \ldots \}$. So, $(x + 1) \neq (n + 1)\#$ because $(x + 1) \in \{ 1,2,3, \ldots \}$ and $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers. Therefore $S(x) \neq S(y)$
                
        + Assume that $B \subseteq P$, $1 \in B$ and $(\forall u)(u \in B \implies S(u) \in B)$
            
            + Let $B = \{ 1,2,3, \ldots \}$
                
                + $B \subseteq P$
                    
                + $1 \in B$
                    
                + If $x \in B$ then $S(x) = x + 1 \in B$
                    
                    + However $B \neq P = \{ 1,2,3, \ldots \} \cup \{ \#1,\#2,\#3, \ldots \}$

## The iteration theorem



# Bibliography

