---
title: "The natural numbers"
date: 2020-06-02 23:02:09
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}


```

# The natural numbers

This section is based on [@mendelson_number_2008, Chapter 2, pp 53-85]

- We define the natural numbers as the positive integers, $\mathbb{P} = \{1, 2, 3, \ldots \}$

- In this part we present an axiomatic development of the natural numbers

## Peano systems

By a __peano system__ we mean

- A set $P$

- A particular element $1 \in P$

- A singularity operation $S$ on $P$ $(S: P \longrightarrow P) \land \mathscr{D}(S) = P$

Such that the following axioms are satisfied:

- $(P1) \; 1$ is not the _successor_ $S(x)$ of any object $x \in P$:

    + $(\forall x)(x \in P \implies S(x) \neq 1)$
    
        + In that sense $1 \notin \mathscr{R}(S)$ so $\mathscr{R}(S) \subsetneq P$

- $(P2)$ Different objects in $P$ have different successors:

    + $(\forall x)(\forall y)((x \in P \land y \in P \land x \neq y) \implies S(x) \neq S(y))$
    
        + $S$ is one-one $(S: P \xrightarrow[]{\text{1-1}} P)$

- $(P3)$ _Principle of Mathematical Induction_: Any subset of $P$ containing $1$ and closed under the successor operation must be identical with $P$

    + $(\forall B)((B \subseteq P \land 1 \in B \land (\forall x)(x \in B \implies S(x) \in P)) \implies B = P)$

Such a peano system will be denoted by the ordered triple $(P, S, 1)$

- $P$ is called the __underlying set__

- $S$ the __successor operation__

- $1$ the __distinguished element__

    + The distinguished element $1$ need not to have anything to do with the ordinary integer 1

### Example of peano systems

- Let $P = \mathbb{P}$, $1$ the ordinary integer $1$ and $S$ equal to $S(x) = x + 1$ for all $x \in \mathbb{P}$

    + $S(x) = x + 1 \neq 1$
    
    + If $x \neq y$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \mathbb{P}$, $1 \in B$ and assume that if $x \in B$ then $x + 1 \in B$
    
        + If $1 \in B$ then $1 + 1 = 2 \in B$, $2 + 1  = 3 \in B$ and so on, where any $n \in \mathbb{P}$ can be reach
        
- Let $P = \mathbb{N}$ where $\mathbb{N} = \mathbb{P} \cup \{ 0 \}$, $1$ the ordinary integer $0$ and $S$ equal to $S(x) = x + 1$ for all $x \in \mathbb{N}$ 

    + $S(x) = x + 1 \neq 0$
    
    + If $x \neq y$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \mathbb{N}$, $0 \in B$ and assume that if $x \in B$ then $x + 1 \in B$
    
        + If $0 \in B$ then $0 + 1 = 1 \in B$, $1 + 1  = 2 \in B$ and so on, where any $n \in \mathbb{N}$ can be reach

- Let $P = \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$, $1$ the ordinary integer $1000$ and $S$ equal to $S(x) = x + 1$ for all $x \in \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$

    + $S(x) = x + 1 \neq 1000$
    
    + If $x \neq y$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$, $1000 \in B$ and assume that if $x \in B$ then $x + 1 \in B$
    
        + If $1000 \in B$ then $1000 + 1 = 1001 \in B$, $1001 + 1  = 1002 \in B$ and so on, where any $x \in \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$ can be reach
        
- Let $P = \mathbb{Z}^-$, $1$ the ordinary integer $-1$ and $S$ equal to $S(x) = x - 1$ for all $x \in \mathbb{Z}^-$

    + $S(x) = x - 1 \neq -1$
    
    + If $x \neq y$ then $x - 1 \neq y - 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \mathbb{Z}^-$, $-1 \in B$ and assume that if $x \in B$ then $x - 1 \in B$
    
        + If $-1 \in B$ then $-1 - 1 = -2 \in B$, $-2 -1 = -3 \in B$ and so on, where any $z \in \mathbb{Z}^-$ can be reach
        
- Let $P = \{z: z = 2n \land n \in \mathbb{P} \}$, $1$ the ordinary integer $2$ and $S$ equal to $S(x) = x + 2$ for all $x \in \{z: z = 2n \land n \in \mathbb{P} \}$

    + $S(x) = x + 2 \neq 2$
    
    + If $x \neq y$ then $x + 2 \neq y + 2$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \{z: z = 2n \land n \in \mathbb{P} \}$, $2 \in B$ and assume that if $x \in B$ then $x + 2 \in B$
    
        + If $2 \in B$ then $2 + 2 = 4 \in B$, $4 + 2 = 6 \in B$ and so on, where any $x \in \{z: z = 2n \land n \in \mathbb{P} \}$ can be reach

__Theorem__ Every element different from $1$ and belonging to $P$ is a successor $$(\forall x)((x \in P \land x \neq 1) \implies (\exists y)(y \in P \land x = S(y)))$$ 

Let $B = \{x: x = 1 \lor (\exists y)(x = S(y)) \}$

- $1 \in B$

- If $x \in B$ then $x = 1 \lor x = S(y)$ where $S(y) \neq 1 \land S(y) \in P$

    + Because $1 \in P \land S(y) \in P$ then $B \subseteq P$

- $\begin{split}
    x \in B \land x \neq 1 & \implies x = S(y) \\
    & \implies S(x) = S(S(y)) \\
    & \implies S(x) \in B
    \end{split}$
    
    + Therefore by the principle of mathematical induction $B = P$

__Theorem__ No object is its own successor, that is, $(\forall x)(x \in P \land S(x) \neq x)$

Let $B = \{x: x \in P \land S(x) \neq x \}$

- $1 \in B$ because $1 \in P \land S(x) \neq 1$
    
- If $x \in B$ then $x \in P$ so $B \subseteq P$
    
- $\begin{split}
    x \in B \land & \implies S(x) \neq x \land S(x) \in P \land x \in P \\
    & \implies S(S(x)) \neq S(x) \land S(x) \in P \\
    & \implies S(x) \in B
    \end{split}$
    
    + Therefore by the principle of mathematical induction $B = P$
    
## Peano systems exercises and solutions

1. Prove that every object different from $1$ is the successor of a unique object, that is, $$(\forall x)(x \in P \implies (x = 1 \lor (\exists !y)(y \in P \land x = S(y))))$$

- If $1 \in P$ then $1 = 1$

- If $x \in P \land x \neq 1$ then $(\exists y)(y \in P \land x = S(y))))$

    + Assume that there is another $z$ such that $z \in P \land x = S(z)$
    
        + Therefore $S(y) = S(z) \implies y = z$ because $S$ is a one-one function
        
            + Thus if $x \in P \land x \neq 1$ then $(\exists !y)(y \in P \land x = S(y))))$
            
2. Prove $(\forall x)(x \in P \implies (x = 1 \lor x = S(1) \lor x = S(S(y))))$

- If $1 \in P$ then $1 = 1$

- If $x \in P \land x \neq y$ then $(\exists !z)(z \in P \land x = S(z))$

    + If $z = 1$ then $x = S(1)$
    
    + If $z \neq 1$ then $(\exists !y)(y \in P \land z = S(y))$. Therefore $(\exists !y)(y \in P \land x = S(S(y)))$
    
3. Determine whether or not the following structures $(P, S, 1)$ are peano systems

    + $P$ is the set of all integers greater than $9$, $1$ stands for the integer $10$ and $S(u) = u + 1$ for any $u \in P$
    
        + $P = \{u: u \in \mathbb{Z} \land x > 9 \}$
        
            + $(\forall u)(u \in P \implies u + 1 \neq 10)$
            
            + $(\forall u)(\forall v)((u \in P \land v \in P \land u \neq v) \implies u + 1 \neq v + 1)$
            
            + Assume that $B \subseteq P$, $10 \in B$ and $(\forall u)(u \in B \implies u + 1 \in B)$
                + $10 \in P$, $10 + 1 = 11 \in P$, $11 + 1 = 12 \in P$ and so on where any $u \in P$ can be reach
                
    + $P$ is the set of all integers, $1$ stands for the ordinary integer $1$ and $S(u) = u + 1$ for all $u$ in $P$
    
        + $P = \mathbb{Z}$
        
            + $0 \in \mathbb{Z} \land S(0) = 0 + 1 = 1$. Therefore $S(0) = 1$
        
    + $P$ is the set of all integers, $1$ stands for the ordinary integer $0$ and $$S(u) = \begin{cases}
    1        & \text{if } u = 0 \\
    -u       & \text{if } u > 0 \\
    -(u - 1) & \text{if } u < 0
    \end{cases}$$
    
        + $P = \mathbb{Z}$
            
            + $S(u) = 1 \lor S(u) < 0 \lor S(u) > 0$ so $S(x) \neq 0$
            
            + Let $u \in P \land v \in P \land u \neq v$
            
                + If $u = 0$ then $S(1) = 0 \land (v > 0  \lor v < 0)$
                
                    + If $v > 0$ then $S(v) = -v < 0$ so $S(v) \neq S(u)$
                    
                    + If $v < 0$ then $S(v) = -(v-1) > 1$ so $S(v) \neq S(u)$
                    
                + If $u > 0$ then $S(u) = -u < 0 \land (v = 0  \lor v < 0)$
                
                    + If $v = 0$ then $S(v) = 1$ so $S(v) \neq S(u)$
                    
                    + If $v < 0$ then $S(v) = -(v-1) > 1$ so $S(v) \neq S(u)$
                    
                + If $u < 0$ then $S(u) = -(u - 1) > 1 \land (v = 0  \lor v > 0)$
                
                    + If $v = 0$ then $S(v) = 1$ so $S(v) \neq S(u)$
                    
                    + If $v < 0$ then $S(v) = -(v-1) > 1$ so $S(v) \neq S(u)$
                    
            + Assume that $B \subseteq P$, $0 \in B$ and $(\forall u)(u \in B \implies S(u)  \in B)$ where $$S(u) = \begin{cases}
    1        & \text{if } u = 0 \\
    -u       & \text{if } u > 0 \\
    -(u - 1) & \text{if } u < 0
    \end{cases}$$
    
                + $0 \in B$, $S(0) = 1 > 0 \land 1 \in P$, $S(1) = -1 < 0 \land -1 \in P$,  $S(-1) = -(-1 - 1) = 2 > 0 \land 2 \in P$, $S(2) = -2 < 0 \land -2 \in P$ and so on where any $u \in P$ can be reach
        
    + $P$ is the set of all rational numbers of the form $\frac{1}{2^n}$ where $n$ is a nonnegative integer, $1$ stands for the integer $1$ and $S(u) = \frac{1}{2}u$ for all $u \in P$
    
        + $P = \{u: u = \frac{1}{2^n} \land n \in \mathbb{Z^+}\}$
        
            + $1 \in P$ because $0 \in \mathbb{Z^+}$ and $\frac{1}{2^0} = \frac{1}{1} = 1$
            
            + $S(u) = \frac{1}{2}u$ where $u = \frac{1}{2^n}$ and $n \in \mathbb{Z^+}$. Therefore $S(u) = \frac{1}{2} \frac{1}{2^n} = \frac{1}{2^{n + 1}} \neq 1$
            
            + Assume that $B \subseteq P$, $1 \in B$ and $(\forall u)(u \in B \implies \frac{1}{2}u \in B)$
            
                + $1 \in B$, $\frac{1}{2} * 1 = \frac{1}{2} = \frac{1}{2^1}$, $\frac{1}{2} \frac{1}{2} = \frac{1}{2^2}$ and so on where any $u \in P$ can be reach
                
    + $P = \{ 1,2,3,4 \}$, $1$ stands for the ordinary integer $1$ and $S(1) = 2, S(2) = 3, S(4) = 1$
      
        + $4 \in P \land S(4) = 1$ so it is not true that for all $x \in P$ then $S(x) \neq 1$
          
    + $P = \{ 1,2,3, \ldots \} \cup \{ \#1,\#2,\#3, \ldots \}$, where $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers. Let $1$ denote the integer $1$. Let $S(n) = n + 1$ for any natural number $n$ and let $S(\#n) = (n + 1)\#$ for any natural number $n$. Thus, $S(1) = 2, S(2) = 3, \ldots$ and $S(\#1) = \#2, S(\#2) = \#3, \ldots$
 
        + If $n \in P$ then $n = m \in \{ 1,2,3, \ldots \}$ or $n = \#m \in \{ \#1,\#2,\#3, \ldots \}$ but not both because $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers
        
            + If $n = m$ then $S(m) = m + 1 \neq 1$
            
            + If $n = \#m$ then $S(\#m) = (m + 1)\# \neq 1$ because $1 \in \{ 1,2,3, \ldots \}$ and $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers
            
        + Let $x \in P \land y \in P \land x \neq y$
        
            + If $x \neq y$ then $y \in \{ 1,2,3, \ldots \}$ or $y \in \{ \#1,\#2,\#3, \ldots \}$
            
                + If $y \in \{ 1,2,3, \ldots \}$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
            
                + If $y \in \{ \#1,\#2,\#3, \ldots \}$ then $S(x) = x + 1 \land S(y) = (n + 1)\#$ for some $n \in \{ 1,2,3, \ldots \}$. So, $(x + 1) \neq (n + 1)\#$ because $(x + 1) \in \{ 1,2,3, \ldots \}$ and $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers. Therefore $S(x) \neq S(y)$
                
        + Assume that $B \subseteq P$, $1 \in B$ and $(\forall u)(u \in B \implies S(u) \in B)$
            
            + Let $B = \{ 1,2,3, \ldots \}$
                
                + $B \subseteq P$
                    
                + $1 \in B$
                    
                + If $x \in B$ then $S(x) = x + 1 \in B$
                    
                    + However $B \neq P = \{ 1,2,3, \ldots \} \cup \{ \#1,\#2,\#3, \ldots \}$

## The iteration theorem

Arithmetical notions such as $+$ and $\times$ are not mention in the definition of a peano system

- These notions will be _defined by induction_

- The idea of definition by induction, is different from a proof by the principle of mathematical induction: Axiom $(P3)$

- To define a function $f$ by inductionwe apply the following process:

    + We specify $f(1)$
    
    + Then indicate a rule for obtaining $f(S(x))$ from a previous value $f(x)$
    
        + This determines a unique function $f$ defined on $P$
        
__Theorem__ _(Iteration Theorem)_ Consider any peano system $(P, S, 1)$. Let $W$ be any arbitary set, let $c$ be a fixed element of $W$ and let $g$ be a singulary operation on $W$, $g:W \longrightarrow W$. Then there is a unique function $F:P \longrightarrow W$ such that

- $F(1) = c$

- $F(S(x)) = g(F(x))$ for all $x \in P$

__Examples of the iteration theorem__

1. Let $W = \mathbb{R}$, $c = \sqrt{2}$ and $g(x) = x + \sqrt{2}$ for all $x \in \mathbb{R}$

    + $F(1) = \sqrt{2}$
    
    + $F(S(1)) = g(F(1)) = F(1) + \sqrt{2} = \sqrt{2} + \sqrt{2} = 2\sqrt{2}$
    
    + $F(S(2)) = g(F(2)) = 2\sqrt{2} + \sqrt{2} = 3\sqrt{2}$
    
    + $\cdots$
    
    + $F(S(n-1)) = g(F(n-1)) = (n - 1)\sqrt{2} + \sqrt{2} = n\sqrt{2}$
    
2. Let $W = \mathbb{Q}$, $c = \frac{1}{2}$ and $g(x) = \frac{x}{2}$ for all $x \in \mathbb{Q}$

    + $F(1) = \frac{1}{2}$
    
    + $F(S(1)) = g(F(1)) = \frac{F(1)}{2} = \frac{1}{2^2}$
    
    + $F(S(2)) = g(F(2)) = \frac{F(2)}{2} = \frac{1}{2^3}$
    
    + $\cdots$
    
    + $F(S(n-1)) = g(F(n-1)) = \frac{F(n-1)}{2} = \frac{1}{2^n}$
    
__Proof of the iteration theorem__

- First we will prove the __existence__ of a suitable function $F$ and the its __uniqueness__

    + __Existence__
    
        + Take any $n$ in $P$. A function $f: A \longrightarrow W$ is said to be __n-admissible__ if and only if:
        
            + $A \subseteq P$
            
            + $1 \in A$
            
            + $n \in A$
            
            + $(\forall u)(S(u) \in A \implies u \in A)$
            
            + $f(1) = c$
            
            + $(\forall u)(S(u) \in A \implies f(S(u)) = g(f(u)))$
            
                + For example let $(\mathbb{P}, 1, S(u))$ where $S(u) = u + 1$, $W = \mathbb{R}$, $c = 1$, $g(x) = x + 1$, $A = \{ 1,2 \}$ then $f(x) = x$ is $2$-admissible because:
                
                    + $\{ 1,2 \} \subseteq \mathbb{P}$
                    
                    + $1 \in \{ 1,2 \}$
                    
                    + $2 \in \{ 1,2 \}$
                    
                    + $S(1) = 2 \in A \land 1 \in A$
                    
                    + $f(1) = 1$
                    
                    + $S(1) = 2 \in A \land f(S(1)) = 2 \land g(f(1)) = f(1) + 1 = 2$
                    
        + If $f$ is $S(n)$-admissible then $f$ is $n$-admissible
        
            + $f: A \longrightarrow W$, $g: W \longrightarrow W$, $S(n) \in P$
            
            + $A \subseteq P$
            
            + $1 \in A$
            
            + $S(n) \in A$
            
            + $(\forall u)(S(u) \in A \implies u \in A)$ then $S(n) \in A \implies n \in A$
            
            + $f(1) = c$
            
            + $(\forall u)(S(u) \in A \implies f(S(u)) = g(f(u)))$
            
                + Therefore $f$ is $n$-admissible 
                                    
                    + For example let $(\mathbb{P}, 1, S(u))$ where $S(u) = u + 1$, $W = \mathbb{R}$, $c = 1$, $g(x) = x + 1$, $A = \{ 1,2 \}$ then $f(x) = x$ is $2$-admissible. Therefore $f(x) = x$ is also $1$-admissible
                    
          + For any $n \in P$, there exists at least one $n$-admissible function
          
              + Let $B = \{n: n \in P \land (\exists f)(f \text{ is } n \text{-admissible}) \}$
              
                  + $B \subseteq P$ because if $n \in B$ then $n \in P$
              
                  + $1 \in B$ because $f: \{1 \} \longrightarrow W$ where $f(1) = c$ with $c \in W$ is $1$-admissible:
          
                      + $\{ 1 \} \subseteq P$
              
                      + $1 \in P$
              
                      + $S(u) \neq 1$ so $S(1) \notin \{ 1 \}$
              
                      + $f(1) = c$
                      
                  + Assume $n \in B$ then there exists an $f$ that is $n$-admissible. Lets define the following function:
                  
                      + $f: A^* \longrightarrow W$ where $A^* = A \cup \{ S(n) \}$ 
                  
                      + $f^*(u) = \begin{cases}
                                   f(u)    & \text{ if } u \in A \\
                                   g(f(n)) & \text{ if } u = S(n) 
                                   \end{cases}$
                                   
                          + $n \in A \land A \subseteq P$ so $n \in P$. Therefore $S(n) \in P$ and $A^* \subseteq P$
                          + $S(n) \in A^*$
                      
                          + $1 \in A \land A \subseteq A^*$ so $1 \in A^*$
                      
                          + If $u \in A$ then $(\forall u)(S(u) \in A \implies u \in A)$. Also $S(n) \in A^* \land n \in A^* \land A \subseteq A^*$. Therefore $(\forall u)(S(u) \in A^* \implies u in A^*)$
                          + $f^*(1) = f(1) = c$
                      
                          + If $u \in A$ then $(\forall u)(S(u) \in A \implies f(S(u)) = g(f(u)))$. Also $f^*(S(n)) = g(f(n))$ so $(\forall u)(S(u) \in A^* \implies f^*(S(u)) = g(f^*(u)))$
                      
                              + Therefore $f^*$ is $S(n)$-admissible and $S(n) \in B$
                              
              + By the principle of mathematical induction $B = P$

          + If $f$ is $n$-admissible and $h$ is $n$-admissible then $f(n) = h(n)$ 
          
              + Let $B = \{n: n \in P \land (\forall f)(\forall g)((f \text{ is } n \text{-admissible} \land g \text{ is } n \text{-admissible}) \implies f(n) = g(n))\}$
              
                  + $B \subseteq P$ because if $n \in B$ then $n \in P$
              
                  + $1 \in B$ because if $f$ and $h$ are $n$-admissible then $f(1) = c = h(1)$
                  + Assume $n \in B$ then $f(n) = h(n)$. 
                  
                      + Because $S(n) \in P$ we can use $f^*$ as the $S(n)$-admissible function defined above. Therefore $f^*(S(n)) = g(f^*(n))) = g(f(n)) = g(h(n))$
                      
                      + Because $S(n) \in P$ we can use $h^*$ as the $S(n)$-admissible function defined above. Therefore $h^*(S(n)) = g(h^*(n))) = g(h(n))$
                      
                          + Thus we have that $f^*(S(n)) = h^*(S(n))$ and $S(n) \in B$

              + By the principle of mathematical induction $B = P$
              
          + Let any $n \in P$ then $S(n) \in P$. Also there exists a $S(n)$-admissible function $f^*$. Because $f^*$ is an $S(n)$-admissible function then $f^*$ is also a $n$-admissible function.
          
              + Lets define $F(n) = f^*(n)$ 
              
                  + $F(1) = f^*(1) = c$
                  
                  + $F(S(n)) = f^*(S(n)) = g(f^*(n)) = g(F(n))$ 
                  
                  + Because $f^*(n) \in \mathscr{f^*}$ and $mathscr{f^*} \subseteq W$ we have $F(n) \in W$
                  
                    + Therefore for any peano system $(P, S, 1)$, an arbitrary set $W$m an element $c \in W$ and some $g: W \longrightarrow W$ exists a function $F: P \longrightarrow W$ such that $F(1) = c$ and $F(S(n)) = g(F(n))$ for all $x \in P$

    + __Uniqueness__
    
        + Assume $F$ and $H$ are functions satisfying the two conditions of the theorem. For $n \in P$ we have that $F(n) = f(n)$ and $H(n) = h(n)$ where $f$ and $g$ are $n$-admissible functions. Therefore we have that $f(n) = h(n)$ and this means that $F(n) = H(n)$
        
            + Let $B = \{n: n \in P \land F(n) = H(n) \}$
            
                + If $n \in B$ then $n \in P$ so $B \subseteq P$
                
                + $1 \in B$ because $F(1) = c = H(1)$
                
                + Assume $n \in B$ then $F(S(n)) = g(F(n)) = g(H(n)) = H(S(n))$. Therefore $S(n) \in B$
                
            + By the principle of mathematical induction $B = P$
            
        + Thus, $F(x) = H(x)$ for all $x \in P$
        
        + Also $\mathscr{D}(F) = P = \mathscr{D}(H)$
        
            + We can conclude that $F = H$
            
## Aplication of the iteration theorem: addition

__Theorem__ _(Addition)_ Consider any peano system $(P, S, 1)$. Then there is a unique binary operation $+$ on $P$, $+ : P \times P \longrightarrow P$ such that

- $(\alpha) \; x + 1 = S(x)$ for all $x \in P$
- $x + S(y) = S(x + y)$ for all $x \in P$ and $y \in P$

We adopt the standard convention of writing $x + y = z$ instead of $+((x,y)) = z$

__Proof__


# Bibliography

