---
title: "The natural numbers"
date: 2020-06-02 23:02:09
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}


```

# The natural numbers

This section is based on [@mendelson_number_2008, Chapter 2, pp 53-85]

- We define the natural numbers as the positive integers, $\mathbb{P} = \{1, 2, 3, \ldots \}$

- In this part we present an axiomatic development of the natural numbers

## Peano systems

By a __peano system__ we mean

- A set $P$

- A particular element $1 \in P$

- A singularity operation $S$ on $P$ $(S: P \longrightarrow P) \land \mathscr{D}(S) = P$

Such that the following axioms are satisfied:

- $(P1) \; 1$ is not the _successor_ $S(x)$ of any object $x \in P$:

    + $(\forall x)(x \in P \implies S(x) \neq 1)$
    
        + In that sense $1 \notin \mathscr{R}(S)$ so $\mathscr{R}(S) \subsetneq P$

- $(P2)$ Different objects in $P$ have different successors:

    + $(\forall x)(\forall y)((x \in P \land y \in P \land x \neq y) \implies S(x) \neq S(y))$
    
        + $S$ is one-one $(S: P \xrightarrow[]{\text{1-1}} P)$

- $(P3)$ _Principle of Mathematical Induction_: Any subset of $P$ containing $1$ and closed under the successor operation must be identical with $P$

    + $(\forall B)((B \subseteq P \land 1 \in B \land (\forall x)(x \in B \implies S(x) \in P)) \implies B = P)$

Such a peano system will be denoted by the ordered triple $(P, S, 1)$

- $P$ is called the __underlying set__

- $S$ the __successor operation__

- $1$ the __distinguished element__

    + The distinguished element $1$ need not to have anything to do with the ordinary integer 1

### Example of peano systems

- Let $P = \mathbb{P}$, $1$ the ordinary integer $1$ and $S$ equal to $S(x) = x + 1$ for all $x \in \mathbb{P}$

    + $S(x) = x + 1 \neq 1$
    
    + If $x \neq y$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \mathbb{P}$, $1 \in B$ and assume that if $x \in B$ then $x + 1 \in B$
    
        + If $1 \in B$ then $1 + 1 = 2 \in B$, $2 + 1  = 3 \in B$ and so on, where any $n \in \mathbb{P}$ can be reach
        
- Let $P = \mathbb{N}$ where $\mathbb{N} = \mathbb{P} \cup \{ 0 \}$, $1$ the ordinary integer $0$ and $S$ equal to $S(x) = x + 1$ for all $x \in \mathbb{N}$ 

    + $S(x) = x + 1 \neq 0$
    
    + If $x \neq y$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \mathbb{N}$, $0 \in B$ and assume that if $x \in B$ then $x + 1 \in B$
    
        + If $0 \in B$ then $0 + 1 = 1 \in B$, $1 + 1  = 2 \in B$ and so on, where any $n \in \mathbb{N}$ can be reach

- Let $P = \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$, $1$ the ordinary integer $1000$ and $S$ equal to $S(x) = x + 1$ for all $x \in \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$

    + $S(x) = x + 1 \neq 1000$
    
    + If $x \neq y$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$, $1000 \in B$ and assume that if $x \in B$ then $x + 1 \in B$
    
        + If $1000 \in B$ then $1000 + 1 = 1001 \in B$, $1001 + 1  = 1002 \in B$ and so on, where any $x \in \{ z: z \geq 1000 \land z \in \mathbb{Z} \}$ can be reach
        
- Let $P = \mathbb{Z}^-$, $1$ the ordinary integer $-1$ and $S$ equal to $S(x) = x - 1$ for all $x \in \mathbb{Z}^-$

    + $S(x) = x - 1 \neq -1$
    
    + If $x \neq y$ then $x - 1 \neq y - 1$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \mathbb{Z}^-$, $-1 \in B$ and assume that if $x \in B$ then $x - 1 \in B$
    
        + If $-1 \in B$ then $-1 - 1 = -2 \in B$, $-2 -1 = -3 \in B$ and so on, where any $z \in \mathbb{Z}^-$ can be reach
        
- Let $P = \{z: z = 2n \land n \in \mathbb{P} \}$, $1$ the ordinary integer $2$ and $S$ equal to $S(x) = x + 2$ for all $x \in \{z: z = 2n \land n \in \mathbb{P} \}$

    + $S(x) = x + 2 \neq 2$
    
    + If $x \neq y$ then $x + 2 \neq y + 2$ so $S(x) \neq S(y)$
    
    + Let $B \subseteq \{z: z = 2n \land n \in \mathbb{P} \}$, $2 \in B$ and assume that if $x \in B$ then $x + 2 \in B$
    
        + If $2 \in B$ then $2 + 2 = 4 \in B$, $4 + 2 = 6 \in B$ and so on, where any $x \in \{z: z = 2n \land n \in \mathbb{P} \}$ can be reach

__Theorem__ Every element different from $1$ and belonging to $P$ is a successor $$(\forall x)((x \in P \land x \neq 1) \implies (\exists y)(y \in P \land x = S(y)))$$ 

Let $B = \{x: x = 1 \lor (\exists y)(x = S(y)) \}$

- $1 \in B$

- If $x \in B$ then $x = 1 \lor x = S(y)$ where $S(y) \neq 1 \land S(y) \in P$

    + Because $1 \in P \land S(y) \in P$ then $B \subseteq P$

- $\begin{split}
    x \in B \land x \neq 1 & \implies x = S(y) \\
    & \implies S(x) = S(S(y)) \\
    & \implies S(x) \in B
    \end{split}$
    
    + Therefore by the principle of mathematical induction $B = P$

__Theorem__ No object is its own successor, that is, $(\forall x)(x \in P \land S(x) \neq x)$

Let $B = \{x: x \in P \land S(x) \neq x \}$

- $1 \in B$ because $1 \in P \land S(x) \neq 1$
    
- If $x \in B$ then $x \in P$ so $B \subseteq P$
    
- $\begin{split}
    x \in B \land & \implies S(x) \neq x \land S(x) \in P \land x \in P \\
    & \implies S(S(x)) \neq S(x) \land S(x) \in P \\
    & \implies S(x) \in B
    \end{split}$
    
    + Therefore by the principle of mathematical induction $B = P$
    
## Peano systems exercises and solutions

1. Prove that every object different from $1$ is the successor of a unique object, that is, $$(\forall x)(x \in P \implies (x = 1 \lor (\exists !y)(y \in P \land x = S(y))))$$

- If $1 \in P$ then $1 = 1$

- If $x \in P \land x \neq 1$ then $(\exists y)(y \in P \land x = S(y))))$

    + Assume that there is another $z$ such that $z \in P \land x = S(z)$
    
        + Therefore $S(y) = S(z) \implies y = z$ because $S$ is a one-one function
        
            + Thus if $x \in P \land x \neq 1$ then $(\exists !y)(y \in P \land x = S(y))))$
            
2. Prove $(\forall x)(x \in P \implies (x = 1 \lor x = S(1) \lor x = S(S(y))))$

- If $1 \in P$ then $1 = 1$

- If $x \in P \land x \neq y$ then $(\exists !z)(z \in P \land x = S(z))$

    + If $z = 1$ then $x = S(1)$
    
    + If $z \neq 1$ then $(\exists !y)(y \in P \land z = S(y))$. Therefore $(\exists !y)(y \in P \land x = S(S(y)))$
    
3. Determine whether or not the following structures $(P, S, 1)$ are peano systems

    + $P$ is the set of all integers greater than $9$, $1$ stands for the integer $10$ and $S(u) = u + 1$ for any $u \in P$
    
        + $P = \{u: u \in \mathbb{Z} \land x > 9 \}$
        
            + $(\forall u)(u \in P \implies u + 1 \neq 10)$
            
            + $(\forall u)(\forall v)((u \in P \land v \in P \land u \neq v) \implies u + 1 \neq v + 1)$
            
            + Assume that $B \subseteq P$, $10 \in B$ and $(\forall u)(u \in B \implies u + 1 \in B)$
                + $10 \in P$, $10 + 1 = 11 \in P$, $11 + 1 = 12 \in P$ and so on where any $u \in P$ can be reach
                
    + $P$ is the set of all integers, $1$ stands for the ordinary integer $1$ and $S(u) = u + 1$ for all $u$ in $P$
    
        + $P = \mathbb{Z}$
        
            + $0 \in \mathbb{Z} \land S(0) = 0 + 1 = 1$. Therefore $S(0) = 1$
        
    + $P$ is the set of all integers, $1$ stands for the ordinary integer $0$ and $$S(u) = \begin{cases}
    1        & \text{if } u = 0 \\
    -u       & \text{if } u > 0 \\
    -(u - 1) & \text{if } u < 0
    \end{cases}$$
    
        + $P = \mathbb{Z}$
            
            + $S(u) = 1 \lor S(u) < 0 \lor S(u) > 0$ so $S(x) \neq 0$
            
            + Let $u \in P \land v \in P \land u \neq v$
            
                + If $u = 0$ then $S(1) = 0 \land (v > 0  \lor v < 0)$
                
                    + If $v > 0$ then $S(v) = -v < 0$ so $S(v) \neq S(u)$
                    
                    + If $v < 0$ then $S(v) = -(v-1) > 1$ so $S(v) \neq S(u)$
                    
                + If $u > 0$ then $S(u) = -u < 0 \land (v = 0  \lor v < 0)$
                
                    + If $v = 0$ then $S(v) = 1$ so $S(v) \neq S(u)$
                    
                    + If $v < 0$ then $S(v) = -(v-1) > 1$ so $S(v) \neq S(u)$
                    
                + If $u < 0$ then $S(u) = -(u - 1) > 1 \land (v = 0  \lor v > 0)$
                
                    + If $v = 0$ then $S(v) = 1$ so $S(v) \neq S(u)$
                    
                    + If $v < 0$ then $S(v) = -(v-1) > 1$ so $S(v) \neq S(u)$
                    
            + Assume that $B \subseteq P$, $0 \in B$ and $(\forall u)(u \in B \implies S(u)  \in B)$ where $$S(u) = \begin{cases}
    1        & \text{if } u = 0 \\
    -u       & \text{if } u > 0 \\
    -(u - 1) & \text{if } u < 0
    \end{cases}$$
    
                + $0 \in B$, $S(0) = 1 > 0 \land 1 \in P$, $S(1) = -1 < 0 \land -1 \in P$,  $S(-1) = -(-1 - 1) = 2 > 0 \land 2 \in P$, $S(2) = -2 < 0 \land -2 \in P$ and so on where any $u \in P$ can be reach
        
    + $P$ is the set of all rational numbers of the form $\frac{1}{2^n}$ where $n$ is a nonnegative integer, $1$ stands for the integer $1$ and $S(u) = \frac{1}{2}u$ for all $u \in P$
    
        + $P = \{u: u = \frac{1}{2^n} \land n \in \mathbb{Z^+}\}$
        
            + $1 \in P$ because $0 \in \mathbb{Z^+}$ and $\frac{1}{2^0} = \frac{1}{1} = 1$
            
            + $S(u) = \frac{1}{2}u$ where $u = \frac{1}{2^n}$ and $n \in \mathbb{Z^+}$. Therefore $S(u) = \frac{1}{2} \frac{1}{2^n} = \frac{1}{2^{n + 1}} \neq 1$
            
            + Assume that $B \subseteq P$, $1 \in B$ and $(\forall u)(u \in B \implies \frac{1}{2}u \in B)$
            
                + $1 \in B$, $\frac{1}{2} * 1 = \frac{1}{2} = \frac{1}{2^1}$, $\frac{1}{2} \frac{1}{2} = \frac{1}{2^2}$ and so on where any $u \in P$ can be reach
                
    + $P = \{ 1,2,3,4 \}$, $1$ stands for the ordinary integer $1$ and $S(1) = 2, S(2) = 3, S(4) = 1$
      
        + $4 \in P \land S(4) = 1$ so it is not true that for all $x \in P$ then $S(x) \neq 1$
          
    + $P = \{ 1,2,3, \ldots \} \cup \{ \#1,\#2,\#3, \ldots \}$, where $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers. Let $1$ denote the integer $1$. Let $S(n) = n + 1$ for any natural number $n$ and let $S(\#n) = (n + 1)\#$ for any natural number $n$. Thus, $S(1) = 2, S(2) = 3, \ldots$ and $S(\#1) = \#2, S(\#2) = \#3, \ldots$
 
        + If $n \in P$ then $n = m \in \{ 1,2,3, \ldots \}$ or $n = \#m \in \{ \#1,\#2,\#3, \ldots \}$ but not both because $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers
        
            + If $n = m$ then $S(m) = m + 1 \neq 1$
            
            + If $n = \#m$ then $S(\#m) = (m + 1)\# \neq 1$ because $1 \in \{ 1,2,3, \ldots \}$ and $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers
            
        + Let $x \in P \land y \in P \land x \neq y$
        
            + If $x \neq y$ then $y \in \{ 1,2,3, \ldots \}$ or $y \in \{ \#1,\#2,\#3, \ldots \}$
            
                + If $y \in \{ 1,2,3, \ldots \}$ then $x + 1 \neq y + 1$ so $S(x) \neq S(y)$
            
                + If $y \in \{ \#1,\#2,\#3, \ldots \}$ then $S(x) = x + 1 \land S(y) = (n + 1)\#$ for some $n \in \{ 1,2,3, \ldots \}$. So, $(x + 1) \neq (n + 1)\#$ because $(x + 1) \in \{ 1,2,3, \ldots \}$ and $\#1,\#2,\#3, \ldots$ are pairwise-distinct new objects different from the natural numbers. Therefore $S(x) \neq S(y)$
                
        + Assume that $B \subseteq P$, $1 \in B$ and $(\forall u)(u \in B \implies S(u) \in B)$
            
            + Let $B = \{ 1,2,3, \ldots \}$
                
                + $B \subseteq P$
                    
                + $1 \in B$
                    
                + If $x \in B$ then $S(x) = x + 1 \in B$
                    
                    + However $B \neq P = \{ 1,2,3, \ldots \} \cup \{ \#1,\#2,\#3, \ldots \}$

## The iteration theorem

Arithmetical notions such as $+$ and $\times$ are not mention in the definition of a peano system

- These notions will be _defined by induction_

- The idea of definition by induction, is different from a proof by the principle of mathematical induction: Axiom $(P3)$

- To define a function $f$ by inductionwe apply the following process:

    + We specify $f(1)$
    
    + Then indicate a rule for obtaining $f(S(x))$ from a previous value $f(x)$
    
        + This determines a unique function $f$ defined on $P$
        
__Theorem__ _(Iteration Theorem)_ Consider any peano system $(P, S, 1)$. Let $W$ be any arbitary set, let $c$ be a fixed element of $W$ and let $g$ be a singulary operation on $W$, $g:W \longrightarrow W$. Then there is a unique function $F:P \longrightarrow W$ such that

- $F(1) = c$

- $F(S(x)) = g(F(x))$ for all $x \in P$

__Examples of the iteration theorem__

1. Let $W = \mathbb{R}$, $c = \sqrt{2}$ and $g(x) = x + \sqrt{2}$ for all $x \in \mathbb{R}$

    + $F(1) = \sqrt{2}$
    
    + $F(S(1)) = g(F(1)) = F(1) + \sqrt{2} = \sqrt{2} + \sqrt{2} = 2\sqrt{2}$
    
    + $F(S(2)) = g(F(2)) = 2\sqrt{2} + \sqrt{2} = 3\sqrt{2}$
    
    + $\cdots$
    
    + $F(S(n-1)) = g(F(n-1)) = (n - 1)\sqrt{2} + \sqrt{2} = n\sqrt{2}$
    
2. Let $W = \mathbb{Q}$, $c = \frac{1}{2}$ and $g(x) = \frac{x}{2}$ for all $x \in \mathbb{Q}$

    + $F(1) = \frac{1}{2}$
    
    + $F(S(1)) = g(F(1)) = \frac{F(1)}{2} = \frac{1}{2^2}$
    
    + $F(S(2)) = g(F(2)) = \frac{F(2)}{2} = \frac{1}{2^3}$
    
    + $\cdots$
    
    + $F(S(n-1)) = g(F(n-1)) = \frac{F(n-1)}{2} = \frac{1}{2^n}$
    
__Proof of the iteration theorem__

- First we will prove the __existence__ of a suitable function $F$ and the its __uniqueness__

    + __Existence__
    
        + Take any $n$ in $P$. A function $f: A \longrightarrow W$ is said to be __n-admissible__ if and only if:
        
            + $A \subseteq P$
            
            + $1 \in A$
            
            + $n \in A$
            
            + $(\forall u)(S(u) \in A \implies u \in A)$
            
            + $f(1) = c$
            
            + $(\forall u)(S(u) \in A \implies f(S(u)) = g(f(u)))$
            
                + For example let $(\mathbb{P}, 1, S(u))$ where $S(u) = u + 1$, $W = \mathbb{R}$, $c = 1$, $g(x) = x + 1$, $A = \{ 1,2 \}$ then $f(x) = x$ is $2$-admissible because:
                
                    + $\{ 1,2 \} \subseteq \mathbb{P}$
                    
                    + $1 \in \{ 1,2 \}$
                    
                    + $2 \in \{ 1,2 \}$
                    
                    + $S(1) = 2 \in A \land 1 \in A$
                    
                    + $f(1) = 1$
                    
                    + $S(1) = 2 \in A \land f(S(1)) = 2 \land g(f(1)) = f(1) + 1 = 2$
                    
        + If $f$ is $S(n)$-admissible then $f$ is $n$-admissible
        
            + $f: A \longrightarrow W$, $g: W \longrightarrow W$, $S(n) \in P$
            
            + $A \subseteq P$
            
            + $1 \in A$
            
            + $S(n) \in A$
            
            + $(\forall u)(S(u) \in A \implies u \in A)$ then $S(n) \in A \implies n \in A$
            
            + $f(1) = c$
            
            + $(\forall u)(S(u) \in A \implies f(S(u)) = g(f(u)))$
            
                + Therefore $f$ is $n$-admissible 
                                    
                    + For example let $(\mathbb{P}, 1, S(u))$ where $S(u) = u + 1$, $W = \mathbb{R}$, $c = 1$, $g(x) = x + 1$, $A = \{ 1,2 \}$ then $f(x) = x$ is $2$-admissible. Therefore $f(x) = x$ is also $1$-admissible
                    
          + For any $n \in P$, there exists at least one $n$-admissible function
          
              + Let $B = \{n: n \in P \land (\exists f)(f \text{ is } n \text{-admissible}) \}$
              
                  + $B \subseteq P$ because if $n \in B$ then $n \in P$
              
                  + $1 \in B$ because $f: \{1 \} \longrightarrow W$ where $f(1) = c$ with $c \in W$ is $1$-admissible:
          
                      + $\{ 1 \} \subseteq P$
              
                      + $1 \in P$
              
                      + $S(u) \neq 1$ so $S(1) \notin \{ 1 \}$
              
                      + $f(1) = c$
                      
                  + Assume $n \in B$ then there exists an $f$ that is $n$-admissible. Lets define the following function:
                  
                      + $f: A^* \longrightarrow W$ where $A^* = A \cup \{ S(n) \}$ 
                  
                      + $f^*(u) = \begin{cases}
                                   f(u)    & \text{ if } u \in A \\
                                   g(f(n)) & \text{ if } u = S(n) 
                                   \end{cases}$
                                   
                          + $n \in A \land A \subseteq P$ so $n \in P$. Therefore $S(n) \in P$ and $A^* \subseteq P$
                          + $S(n) \in A^*$
                      
                          + $1 \in A \land A \subseteq A^*$ so $1 \in A^*$
                      
                          + If $u \in A$ then $(\forall u)(S(u) \in A \implies u \in A)$. Also $S(n) \in A^* \land n \in A^* \land A \subseteq A^*$. Therefore $(\forall u)(S(u) \in A^* \implies u in A^*)$
                          + $f^*(1) = f(1) = c$
                      
                          + If $u \in A$ then $(\forall u)(S(u) \in A \implies f(S(u)) = g(f(u)))$. Also $f^*(S(n)) = g(f(n))$ so $(\forall u)(S(u) \in A^* \implies f^*(S(u)) = g(f^*(u)))$
                      
                              + Therefore $f^*$ is $S(n)$-admissible and $S(n) \in B$
                              
              + By the principle of mathematical induction $B = P$

          + If $f$ is $n$-admissible and $h$ is $n$-admissible then $f(n) = h(n)$ 
          
              + Let $B = \{n: n \in P \land (\forall f)(\forall g)((f \text{ is } n \text{-admissible} \land g \text{ is } n \text{-admissible}) \implies f(n) = g(n))\}$
              
                  + $B \subseteq P$ because if $n \in B$ then $n \in P$
              
                  + $1 \in B$ because if $f$ and $h$ are $n$-admissible then $f(1) = c = h(1)$
                  + Assume $n \in B$ then $f(n) = h(n)$. 
                  
                      + Because $S(n) \in P$ we can use $f^*$ as the $S(n)$-admissible function defined above. Therefore $f^*(S(n)) = g(f^*(n))) = g(f(n)) = g(h(n))$
                      
                      + Because $S(n) \in P$ we can use $h^*$ as the $S(n)$-admissible function defined above. Therefore $h^*(S(n)) = g(h^*(n))) = g(h(n))$
                      
                          + Thus we have that $f^*(S(n)) = h^*(S(n))$ and $S(n) \in B$

              + By the principle of mathematical induction $B = P$
              
          + Let any $n \in P$ then $S(n) \in P$. Also there exists a $S(n)$-admissible function $f^*$. Because $f^*$ is an $S(n)$-admissible function then $f^*$ is also a $n$-admissible function.
          
              + Lets define $F(n) = f^*(n)$ 
              
                  + $F(1) = f^*(1) = c$
                  
                  + $F(S(n)) = f^*(S(n)) = g(f^*(n)) = g(F(n))$ 
                  
                  + Because $f^*(n) \in \mathscr{f^*}$ and $mathscr{f^*} \subseteq W$ we have $F(n) \in W$
                  
                    + Therefore for any peano system $(P, S, 1)$, an arbitrary set $W$m an element $c \in W$ and some $g: W \longrightarrow W$ exists a function $F: P \longrightarrow W$ such that $F(1) = c$ and $F(S(n)) = g(F(n))$ for all $x \in P$

    + __Uniqueness__
    
        + Assume $F$ and $H$ are functions satisfying the two conditions of the theorem. For $n \in P$ we have that $F(n) = f(n)$ and $H(n) = h(n)$ where $f$ and $g$ are $n$-admissible functions. Therefore we have that $f(n) = h(n)$ and this means that $F(n) = H(n)$
        
            + Let $B = \{n: n \in P \land F(n) = H(n) \}$
            
                + If $n \in B$ then $n \in P$ so $B \subseteq P$
                
                + $1 \in B$ because $F(1) = c = H(1)$
                
                + Assume $n \in B$ then $F(S(n)) = g(F(n)) = g(H(n)) = H(S(n))$. Therefore $S(n) \in B$
                
            + By the principle of mathematical induction $B = P$
            
        + Thus, $F(x) = H(x)$ for all $x \in P$
        
        + Also $\mathscr{D}(F) = P = \mathscr{D}(H)$
        
            + We can conclude that $F = H$
            
## Aplication of the iteration theorem (addition)

__Theorem__ _(Addition)_ Consider any peano system $(P, S, 1)$. Then there is a unique binary operation $+$ on $P$, $+ : P \times P \longrightarrow P$ such that

- $(\alpha) \; x + 1 = S(x)$ for all $x \in P$
- $(\beta)  \; x + S(y) = S(x + y)$ for all $x \in P$ and $y \in P$

We adopt the standard convention of writing $x + y = z$ instead of $+((x,y)) = z$

__Proof__

Let $P = W$, $S(x) = c \in W$ and $S = g$ then using the iteration theorem there is a unique function $F: P \longrightarrow P$ such that:

- $F((x,1)) = S(x)$
- $F((x,S(y))) = S(F((x,y)))$

Let $F = +$ and $+((z,w)) \equiv z + w$ then we can rewrite the above expressions as: 

- $x + 1 = S(x)$
- $x + S(y) = S(x + y)$

__Terminology__ $+$ is called the __addition__ operation on $P$. $x + y$ is called the __sum__ of $x$ and $y$

- Also because $x + S(y) = S(x + y)$, $S(y) = y + 1$ and $S(x + y) = (x + y) + 1$ then $x + (y + 1) = (x + y) + 1$

- Furthermore if $P = \mathbb{P}$, $x = 1$ and $S(x) = 2$ then the definition of $+$ implies $2 = 1 + 1 = S(1)$, $1 + 2 = 1 + S(1) = S(1 + 1) = S(2) \equiv 3$, $1 + 3 = 1 + S(2) = S(1 + 2) = S(3) \equiv 4$ and so on.

__Theorem__ _(Associativity of Addition)_ $x + (y + z) = (x + y) + z$ for all $x,y$ and $z$ in $P$

- Let $B = \{z: z \in P \land x + (y + z) = (x + y) + z \text{ where } x \in P \land y \in P\}$ 

    + $B \subseteq P$ because if $z \in B$ then $z \in P$

    + $\begin{split}
        x + (y + 1) & = x + S(y) \\
        & = S(x + y) \\
        & = (x + y) + 1
        \end{split}$
    
        + Therefore $1 \in B$
    
    + Assume $z \in B$ therefore $x + (y + z) = (x + y) + z$ where $x \in P \land y \in P$

    + $\begin{split}
        x + (y + S(z)) & = x + S(y + z) \\
        & = S(x + (y + z)) \\
        & = S((x + y) + z) \\
        & = (x + y) + S(z)
        \end{split}$
    
        + Therefore $S(z) \in B$

- Thus by the principle of mathematical induction $B = P$

From the associative law it follows that it does not matter how parentheses are inserted in a sum of three or more terms , and that, as a consequence, parentheses may be omitted.

__Lemma__ 

a. $x + 1 = 1 + x$ for all $x$ in $P$ (Hence $1 + x = S(x)$)

    + Let $B = \{x: x \in P \land x + 1 = 1 + x \}$
    
        + $B \subseteq P$ because if $x \in B$ then $x \in P$
        
        + $1 \in B$ because $1 + 1 = 1 + 1$
        
        + Assume $x \in B$ then $x + 1 = 1 + x$
        
            + $\begin{split}
                S(x) + 1 & = (x + 1) + 1 \\
                & = (1 + x) + 1 \\
                & = 1 + (x + 1) \\
                & = 1 + S(x)
                \end{split}$
                
                + Therefore $S(x) \in B$
                
    + Thus by the principle of mathematical induction $B = P$
    
b. $S(y) + x = S(y + x)$ for all $x$ and $y$ in $P$
        
    + $\begin{split}
        S(y) + x & = (y + 1) + x \\
        & = y + (1 + x) \\
        & = y + (x + 1) \\
        & = (y + x) + 1 \\
        & = S(y + x)
        \end{split}$
        
__Theorem__ _(Commutativity of Addition)_ $x + y = y + x$ for all $x$ and $y$ in $P$

- Let $B = \{x: x \in P \land x + y = y + x \text{ where } y \in P \}$

    + $B \subseteq P$ because if $x \in B$ then $x \in P$
        
    + $1 \in B$ because $1 + y = y + 1$
    
    + Assume $x \in P$ therefore $x + y = y + x$
    
        + $\begin{split}
            S(x) + y & = (x + 1) + y \\
            & = (1 + x) + y \\
            & = 1 + (x + y) \\
            & = 1 + (y + x) \\
            & = (1 + y) + x \\
            & = (y + 1) + x \\
            & = y + (1 + x) \\
            & = y + S(x) \\
            \end{split}$
        
            + Therefore $S(x) \in B$
                
- Thus by the principle of mathematical induction $B = P$

__Theorem__ _(Cancellation Law for Addition)_ $x + z = y + z \implies x = y$ for all $x, y$ and $z$ in $P$

- Let $B = \{z: z \in P \land (x + z = y + z \implies x = y) \text{ where } x \in P \land y \in P \}$

    + $B \subseteq P$ because if $x \in B$ then $x \in P$
        
    + $\begin{split}
        x + 1 = y + 1 & \iff S(x) = S(y)
        & \iff x = y 
        \end{split}$
        
        + Therefore $1 \in B$
        
    + Assume $z \in B$ then $x + z = y + z \implies x = y$
    
        + $\begin{split}
            x + S(z) = y + S(z) & \iff S(x + z) = S(y + z) \\
            & \iff x + z = y + z \\
            & \iff x = y
            \end{split}$
            
            + Therefore $S(z) \in B$
                
- Thus by the principle of mathematical induction $B = P$

__Theorem__ $y \neq x + y$ for all $x$ and $y$ in $P$

- Let $B = \{y: y \in P \land y \neq x + y \text{ where } x \in P \}$

    + $B \subseteq P$ because if $x \in B$ then $x \in P$
        
    + $\begin{split}
        S(x) \neq 1 & \iff \neg(S(x) = 1) \\
        & \iff \neg(1 = S(x)) \\
        & \iff \neg(1 = x + 1) \\
        & \iff 1 \neq x + 1 \\
        \end{split}$
        
        + Therefore $1 \in B$
        
    + Assume $x \in B$ then $y \neq x + y$
    
        + $\begin{split}
            y \neq x + y & \implies S(y) \neq S(x + y) \\
            & \iff S(y) \neq x + S(y)
            \end{split}$ 
            
            + Therefore $S(y) \in B$
            
- Thus by the principle of mathematical induction $B = P$

## Aplication of the iteration theorem: addition exercises and solutions

1. Compute $4 + 2$ and $2 + 4$. (We do not know yet that these two values must be the same.)

    + $\begin{split}
        4 + 2 & = 4 + S(1) \\
        & = S(4 + 1) \\
        & = S(5) \\
        & = 6
        \end{split}$
        
    + $\begin{split}
        2 + 4 & = 2 + S(3) \\
        & = S(2 + 3) \\
        & = S(5) \\
        & = 6
        \end{split}$
        
2. Prove $(x + y) + (u + v) = (x + u) + (y + v)$

    + $\begin{split}
        (x + y) + (u + v) & \iff ((x + y) + u) + v \\
        & \iff (x + (y + u)) + v \\
        & \iff (x + (u + y)) + v \\
        & \iff ((x + u) + y) + v \\
        & \iff (x + u) + (y + v) 
        \end{split}$

## The order relation

__Definition__ $x < y$ if and only if $(\exists z)(z \in P \land x + z = y)$

- $x < y$ is also equivalent to $y > x$

__Definition__ $x > y$ if and only if $(\exists w)(w \in P \land x = y + w)$

__Definition__ $x \not < y$ if and only if $\neg(x < y)$

- $\begin{split}
    \neg(x < y) & \iff \neg(\exists z)(x + z = y) \\
    & \iff (\forall z)\neg(z \in P \land x + z = y) \\
    & \implies (\forall z)(x + z \neq y)
    \end{split}$
    
    + $x \not < y$ is also equivalent to $y \not > x$ 
    
__Theorem__

a. $x \not < x$ for all $x$ in $P$ (irreflexivity of $<$)

- Assume that $x < x$. Therefore $(\exists z)(z \in P \land x + z = z)$. However previously we proved that $x + z \neq x$

b. $(x < y \land y < z) \implies x < z$ for all $x, y$ and $z$ in $P$ (transitivity of $<$)

- $\begin{split}
    x < y \land y < z & \iff (\exists u)(u \in P \land x + u = y) \land (\exists v)(v \in P \land y + v = z) \\
    & \implies (\exists u)(\exists v)(u \in P \land v \in P \land x + u = y \land y + v = z) \\
    & \implies (\exists u)(\exists v)(u \in P \land v \in P \land (x + u) + v = z) \\
    & \implies (\exists u)(\exists v)(u \in P \land v \in P \land x + (u + v) = z) \\
    & \implies (\exists w)(w \in P \land x + w = z) \\
    & \implies x < z
    \end{split}$
    
c. For any $x,y$ in $P$, exactly one of the following conditions hold: $x < y$, $x = y$, $y < x$ (trichotomy)

- Let $B = \{x: x \in P \land (x = y \lor x < y \lor x > y) \text{ where } y \in P \}$

    + __At least one__

        + $B \subseteq P$ because if $x \in B$ then $x \in P$
            
        + If x = 1 then $y = 1 \lor (\exists z)(y = S(z))$. Therefore $y = 1 \lor y = z + 1$ so $x = 1 \lor x < y$
        
        + Assume $x \in B$ therefore $x = y \lor x < y \lor x > y$
        
            + If $x = y$ then $S(x) = S(y)$ and $S(x) = y + 1$. Therefore $S(x) > y$
            
            + If $x > y$ then $(\exists z)(z \in P \land y + z = x)$. Therefore $S(x) = S(y + z) = S(z + y) = S(z) + y$ so $S(x) > y$
            
            + If $x < y$ then $(\exists z)(z \in P \land x + z = y)$. Also $z = 1 \lor (\exists w)(w \in P \land z = S(w))$
            
                + If $z = 1$ then $S(x) = y$
                
                + If $z = S(w)$ then $y = x + S(w) = (x + 1) + w = S(x) + w$. Thus $S(x) < y$
                
                    + Therefore $S(x) \in B$
                
        + Thus by the principle of mathematical induction $B = P$
        
    + __Exactly one__
    
        + If $x = y$
        
            + Is not possible also that $x < y$ or $x > y$ because $y < y$ or $y > y$ is not true
        + If $x < y$ then $x + z = y$
        
            + If $x = y$ also then $y + z = y$ but $y + z \neq y$
            
            + If $x > y$ also then $y + w = x$ and $(y + w) + z = y + (w + z) = y$ but $y + (w + z) \neq y$
        
        + If $x > y$ then $y + z = x$
        
            + If $x = y$ also then $x + z = x$ but $x + z \neq x$
            
            + If $x < y$ also then $x + w = x$ and $(x + w) + z = x + (w + z) = x$ but $x + (w + z) \neq x$

__Theorem__

- $x < S(x)$

    + $\begin{split}
        S(x) = x + 1 & \implies (\exists z)(z \in P \land x + z = S(x)) \\
        & \implies x < S(x)
       \end{split}$
       
- $\neg(\exists y)(x < y < S(x))$ 

    + $\begin{split}
        (\exists y)(x < y < S(x)) & \implies (\exists z_1)(z_1 \in P \land x + z_1 = y) \land (\exists z_2)(z_2 \in P \land y + z_2 = S(x)) \\
        & \implies (\exists z_1)(\exists z_2)((z_1 + z_2) \in P \land x + (z_1 + z_2) = S(x) \\
        & \implies (\exists z_1)(\exists z_2)(x + (z_1 + z_2) = S(x) \land z_1 = 1 \lor (\exists w)(w \in P \land z_1 = S(w)) \\
        & \implies (x + 1) + z_2 = S(x) \lor x + (S(w) + z_2) = S(x) \\
        & \implies S(x) + z_2 = S(x) \lor S(x) + (w + z_2) = S(x) \\
        & \implies \text{ There is a contradiction}
       \end{split}$
       
        + Therefore it is false that $(\exists y)(x < y < S(x)$ and it is true that $\neg(\exists y)(x < y < S(x))$
        
__Definitions__

- $x \leq y$ for $x < y \lor x = y$
- $x \not \leq y$ for $\neg(x \leq y)$
- $x > y$ for $y < x$
- $x \geq y$ for $y \leq x$
       
__Theorem__

- $x \leq x$

    + $\begin{split}
       x = x & \implies x = x \lor x < x \\
       & \implies x \leq x
       \end{split}$
       
- $(x < y \land y \leq z) \implies x < z$

    + $\begin{split}
       x < y \land y \leq z & \iff x < y \land (y < z \lor y = z) \\
       & \implies (x < y \land y < z) \lor (x < y \land y = z) \\
       & \implies x < y \land y < z \\
       & \implies x < z
       \end{split}$
       
- $(x \leq y \land y < z) \implies x < z$

    + $\begin{split}
       x \leq y \land y < z & \iff (x < y \lor x = y) \land y < z \\
       & \implies (x < y \land y < z) \lor (x = y \land y < z) \\
       & \implies x < y \land y < z \\
       & \implies x < z
       \end{split}$
       
-  $(x \leq y \land y < z) \implies x < z$

    + $\begin{split}
       x \leq y \land y \leq z & \iff (x < y \lor x = y) \land (y < z \lor y = z) \\
       & \iff ((x < y \lor x = y) \land y < z) \lor ((x < y \lor x = y) \land y = z) \\
       & \iff (x \leq y \land y < z) \lor (x < y \land y = z) \lor (x = y \land y = z) \\
       & \implies x < z \lor x < z \lor x = z \\
       & \iff x < z \lor x = z \\
       & \iff x \leq z
       \end{split}$
       
- $x \leq y \lor y \leq x$

    + For $x \in P \land y \in P$ exactly one of the following holds:
    
      + $x < y$
      
        + $\begin{split}
           x < y & \implies x < y \lor x = y \\
           & \implies x \leq y
           \end{split}$
      
      + $x = y$
      
        + $\begin{split}
           x = y & \implies x = y \lor x < y \\
           & \implies x \leq y
           \end{split}$
        
      + $x > y$
      
        + $\begin{split}
           x > y & \implies x > y \lor x = y \\
           & \implies x \geq y           
           \end{split}$
           
          + Therefore $x \leq y \lor x \geq y$
          
- $(x \leq y \land x \geq y) \implies x = y$

    + $\begin{split}
        x \leq y \land x \geq y & \iff (x < y \lor x = y) \land (x > y \lor x = y) \\
        & \iff ((x < y \lor x = y) \land x > y) \lor ((x < y \lor x = y) \land x = y) \\
        & \iff (x \leq y \land x > y) \lor (x < y \land  x = y) \lor (x = y \land x = y) \\ 
        & \iff x = y \land x = y \\
        & \iff x = y
        \end{split}$
        
__Theorem__

- $x \neq 1 \implies 1 < x$

    + $\begin{split}
       x \neq 1 & \implies (\exists y)(y \in P \land x = S(y)) \\
       & \implies (\exists y)(y \in P \land x = y + 1)
       & \implies 1 < x
       \end{split}$
       
- $x < x + y$

    + $\begin{split}
       y \in P & \implies y = 1 \lor (\exists w)(w \in P \land y = S(w)) \\
       & \implies x + 1 = x + y \lor x + S(w) = x + y \\
       & \implies x < x + y
       \end{split}$
       
- $x < y \iff x + z < x + y$

    + $\begin{split}
       x < y & \iff (\exists w)(w \in P \land x + w = y) \\
       & \iff  (\exists w)(w \in P \land (x + w) + z = y + z) \\
       & \iff  (\exists w)(w \in P \land (x + z) + w = y + z) \\
       & \iff x + z < y + z  
       \end{split}$
       
        + Checkout _Substitutivity_ of inequality in __Appendix A__ and _Cancellation Law for Addition_ in this chapter to examine why $x + w = y \iff (x + w) + z = y + z$ 

- $x \leq y \iff x + z \leq x + y$

    + In the case of $x < y \lor x + z < y + z$ then $x < y \iff x + z < x + y$
    
    + In the case of $x = y \lor x + z = y + z$ use the _Substitutivity_ of inequality in __Appendix A__ and _Cancellation Law for Addition_ in this chapter to prove that $x = y \iff x + z = y + z$
    
        + Therefore $x \leq y \iff x + z \leq x + y$
        
- $(x < y \land u < v) \implies x + u < y + v$

    + $\begin{split}
       x < y \land u < v & \iff (\exists w_1)(\exists w_2)(w_1 \in P \land w_2 \in P \land x + w_1 = y \land u + w_2 = v) \\
       & \implies (\exists w_1)(\exists w_2)(w_1 \in P \land w_2 \in P \land (x + w_1) + (u + w_2) = y + v) \\
       & \implies (\exists (w_1 + w_2))((w_1 + w_2) \in P \land (x + u) + (w_1 + w_2) = y + v) \\
       & \implies x + u < y + v
       \end{split}$
       
- $(x < y \land u \leq v) \implies x + u < y + v$

    + $\begin{split}
       (x < y \land u \leq v) & \iff x < y \land (u < v \lor u = v) \\
       & \implies (x < y \land u < v) \lor (x < y \land u = v) \\
       & \implies (x + u < y + v) \lor (x + u < y + z ) \\
       & \implies x + u < y + v
       \end{split}$
       
- $(x \leq y \land u < v) \implies x + u < y + v$

    + $\begin{split}
       (x \leq y \land u < v) & \iff (u < v \land x \leq y)  \\
       & \implies u + x < v + y \\
       & \iff x + u < y + v
       \end{split}$
       
- $(x \leq y \land u \leq v) \implies x + u \leq y + v$

    + If $(x \leq y \land u < v)$ then $x + u < y + v$ so  $x + u \leq y + v$ is true
    
    + If $(x \leq y \land u = v)$ then $x + u \leq y + u$ is true and $x + u \leq y + v$ is also true
    
        + Therefore in any case $(x \leq y \land u \leq v) \implies x + u \leq y + v$ is true
        
__Definition__ Let $A \subseteq P$. By a __least element__ of $A$ we mean an object $z \in A$ such that $z$ is less than or equal to every element of $A$. Thus $z$ is a least element of $A$ if and only if

- $z \in A$
- $(\forall u)(u \in A \implies z \leq u)$

If $A = \{a \}$ then $a \leq a$ so a is a least element. Also if $1 \in A$ then $1$ is the least element because $(\forall x)(x \in P \land 1 \leq x)$. The following theorem provides a very useful fact about the order relation.

__Theorem__ _(Least Number Principle)_ Any nonempty subset of $P$ has a least element

__*Proof*__

- Let $\emptyset \neq A \subseteq P$. Assume $A$ has no least element, and let us show that this leads to a contradiction.

    + Let $B$ be the set of all elements $x \in P$ such that $x$ and all smaller elements are not in $A$: $B = \{x: x \in P \land (\forall u)(u \leq x \implies u \notin A) \}$
    
        + If $x \in B$ then $x \notin A$ because $x \leq x$
        
        + Let us show now by mathematical induction that $B = P$
        
            + $B \subseteq P$ because if $x \in B$ then $x \in P$
            
            + If $1 \notin B$ then $(\forall u)(u \leq 1 \land u \in A)$. Also if $u \leq 1$, because $u \neq 1 \implies 1 < u$ is true, then $1 \geq u \implies 1 = u$ is true. Therefore $1 \in A$ and $A$ has a least element contradicting the supposition that $A$ has no elements. Thus, $1 \in B$ if $A$ has no least element
            
            + Assume that $x \in B$ then $(\forall u)(u \leq x \implies u \notin A)$. Therefore $x \notin A$. Because $x < S(x)$ and $\neg(\exists y)(x < y < S(x))$ then $S(x) \notin A$. If $S(x) \in A$ then $S(x)$ will be a least element of $A$ but we have assumed that $A$ has no least elements
            
                + Furthermore because $u \leq x < S(x)$ then $(\forall u)(u \leq S(x) \implies u \notin A)$ will be true if $A$ has no least element so $S(x) \in B$
                
    + By the principle of mathematical induction $B = P$
    
    + Also because $B \cap A = \emptyset$ then $A = \emptyset$. However we have assumed that $A \neq \emptyset$ so there is a contradiction
        
## The order relation exercises and solutions

1. Prove that $x < y \iff (x \leq y \land x \neq y)$

    + $\begin{split}
        x \leq y \land x \neq y & \iff (x < y \lor x = y) \land x \neq y \\
        & \iff (x < y \land x \neq y) \lor (x = y \land x \neq y) \\
        & \iff x < y \land x \neq y \\
        & \iff (\exists z)(z \in P \land x  + z = y) \land x \neq y \\
        & \iff (\exists z)(z \in P \land x  + z = y) \land x \neq x  + z \\
        & \iff (\exists z)(z \in P \land x  + z = y) \\
        & \iff x < y
        \end{split}$
       
2. Prove that $1 \leq x$

    + $\begin{split}
        x \in P & \implies x = 1 \lor (\exists y)(y \in P \land x = S(y)) \\
        & \implies x = 1 \lor (\exists y)(y \in P \land x = y + 1) \\
        & \implies x = 1 \lor 1 < x \\
        & \implies 1 \leq x
        \end{split}$
        
3. Prove that $x \leq 1 \iff x = 1$

    + $\begin{split}
        x \in P & \implies x = 1 \lor (\exists y)(y \in P \land x = S(y)) \\
        & \implies x = 1 \lor (\exists y)(y \in P \land x = y + 1)
        \end{split}$
        
        + Therefore $1 \neq y + 1$. So it is not the case that $x = 1 \land x < 1$

    + $\begin{split}
        x \leq 1 & \iff x < 1 \lor x = 1 \\
        & \iff x = 1
        \end{split}$
        
4. Prove that $x < S(y) \iff x \leq y$

    + $\begin{split}
        x < S(y) & \iff (\exists w)(w \in P \land x + w = S(y)) \\
        & \iff (\exists w)(1 \leq w \land x + w = S(y)) \\ 
        & \iff (\exists w)((1 = w  \lor 1 < w) \land x + w = S(y)) \\ 
        & \iff (\exists w)((1 = w \land x + w = S(y)) \lor (1 < w \land x + w = S(y))) \\
        & \iff (\exists w)((1 = w \land x + w = y + 1) \lor (1 < w \land x + w = y + 1)) \\
        & \iff (\exists w)((1 = w \land x + w = y + 1) \lor ((\exists z)(w = z + 1) \land x + w = y + 1)) \\
        & \iff (\exists w)((1 = w \land x + w = y + 1) \lor ((\exists z)(w = z + 1) \land x + (z + 1) = y + 1)) \\
        & \iff x + 1 = y + 1 \lor (\exists z)((x + z) + 1 = y + 1) \\
        & \iff x = y  \lor (\exists z)((x + z) = y) \\
        & \iff x = y \lor x < y \\
        & \iff x \leq y
        \end{split}$
       
5. Prove that $x < y \iff S(x) \leq y$

    + $\begin{split}
        x < y & \iff (\exists z)(z \in P \land x + z = y) \\
        & \iff (\exists z)((z = 1 \lor (\exists w)(z = S(w))) \land x + z = y) \\
        & \iff (\exists z)((z = 1 \land x + z = y) \lor ((\exists w)(z = S(w)) \land x + z = y)) \\
        & \iff (\exists z)((z = 1 \land x + z = y) \lor ((\exists w)(z = S(w)) \land (x + 1) + w  = y)) \\
        & \iff (\exists z)((z = 1 \land x + z = y) \lor ((\exists w)(z = S(w)) \land S(x) + w  = y)) \\
        & \iff S(x) = y \lor S(x) < y \\
        & \iff S(x) \leq y
        \end{split}$

6. Prove that $x + u < y + v \implies x < y \lor u < v$ 

    + $x + u < y + v \implies x < y \lor u < v$ is equivalent to $x \geq y \land u \geq v \implies x + u \geq y + v$
    
    + Aso we already prove that $x \geq y \land u \geq v \implies x + u \geq y + v$ is true.
    
    + Therefore $x + u < y + v \implies x < y \lor u < v$ is true
    
7. Prove that $1 < 2 < 3 < 4$, etc

    + Because $(\forall x)(x \in P \implies x < S(x))$
    
        + $1 < S(1) \equiv 2$
        
        + $2 < S(2) \equiv 3$
        
        + $3 < S(3) \equiv 4$
        
        + $\vdots$
        
        + $n < S(n) = n + 1$
        
8. If $g: P \longrightarrow P$ and $(\forall x)(\forall y)(x < y \implies g(x) < g(y))$ prove that $(\forall u)(g(u) \geq u)$

- Let $B = \{u: (\forall u)(g(u) \geq u) \text{ where } g: P \longrightarrow P \land (\forall x)(\forall y)(x < y \implies g(x) < g(y)) \}$

    + $B \subseteq P$ because if $u \in B$ then $u \in \mathscr{D}(g) = P$
    
    + $1 \in B$ because 
    
        + $\begin{split}
           g(1) \in & \implies g(1) = 1 \lor (\exists z)(z \in P \land g(1) = S(z)) \\
           & \implies g(1) = 1 \lor (\exists z)(z \in P \land g(1) = z + 1) \\
           & \implies g(1) = 1 \lor g(1) > 1 \\
           & \implies g(1) \geq 1
           \end{split}$
           
    + Assume $u \in B$ then $g(u) \geq u$
    
        + $\begin{split}
           u < S(u)  & \implies u < u + 1 \\ 
           & \implies u \leq g(u) \land g(u) < g(u + 1) \\
           & \implies u < g(u + 1) \\
           & \implies (\exists z)(z \in P \land u  + z  = g(u + 1)) \\
           & \implies (\exists z)((z = 1 \lor (\exists w)(z = S(w))) \land u  + z  = g(u + 1)) \\
           & \implies u + 1 = g(u + 1) \lor u + (1 + w) = g(u + 1) \\
           & \implies u + 1 = g(u + 1) \lor (u + 1) + w = g(u + 1) \\
           & \implies u + 1 = g(u + 1) \lor (u + 1) < g(u + 1) \\
           & \implies u + 1 \leq g(u + 1) \\
           & \implies g(u + 1) \geq u + 1 \\
           & \implies g(u + 1) \geq S(u)  \\
           \end{split}$
           
            + Thus $S(u) \in B$
            
- By the principle of mathematical induction $B = P$ so $(\forall u)(g(u) \geq u)$ when $u \in P$

9. Prove the _Principle of Complete Induction_: $$(\forall B)((B \subseteq P \land (\forall x)((\forall y)(y < x \implies y \in B) \implies x \in B)) \implies P = B)$$
(_Hint_: Assume $P \neq B$. Then, $P - B \neq \emptyset$. Apply the Least Number Principle. Notice that, when $x = 1$, $(\forall y)(y < x \implies y \in B)$ is trivially true, since $y < 1$ is always false)

    + Let $B = \{x: x \in P \land (\forall x)((\forall y)(y < x \implies y \in B) \implies x \in B) \}$

        + $B \subseteq P$ because if $x \in B$ then $x \in P$
    
        + $1 \in B$ because $y < 1$ is false so $y < 1 \implies y \in B$ is true. Furthermore we are assuming that $(\forall y)(y < 1 \implies y \in B) \implies 1 \in B$ is true, thus, $1 \in B$ is true.
    
        + Assume $x \in B$ then because $x < S(x)$ we have that $x < S(x) \implies x \in B$ is true. Because $(\forall y)(x < S(x) \implies x \in B) \implies S(x) \in B$ is true, thus, $S(x) \in B$ is true.
    
    + By the principle of mathematical induction $B = P$

10. By a __greatest__ element of a set $B \subseteq P$ we mean an object $z$ such that $z \in B \land (\forall u)(u \in B \implies u \leq z)$. Prove:

    + A set has at most one greatest element
    
        + Assume that a set has a greatest element $z$. If this set has another gretest element $w$ then $z \leq w \land w \leq z$. Thus, $z = w$
        
            + In that sense if a set has a greatest element then it is unique
        
        + __Observation__: we can use a similar argument to prove that a set has at most one least element 
                    
    + A set need not have a greatest element
    
        + Let $z \in P$ then $S(z) \in P \land z < S(z)$. Therefore there is no $z \in P$ such that $(\forall u)(u \in P \implies u \leq z)$
        
    + If $\emptyset \neq A \subseteq P$ and $A$ is bounded above, that is $(\exists w)((\forall u)(u \in A \implies u \leq w))$ then $A$ has a greatest element. (_Hint_: Let $B = \{w: (\forall u)(u \in A \implies u \leq w) \}$). By hypothesis, $B \neq \emptyset$. Apply the Least Number Principle)
    
        +  Let $B = \{w: (\forall u)(u \in A \implies u \leq w) \}$. Because $A$ is bounded above there exists at least one $w \in B$. Therefore $B \neq \emptyset$
        
            + $B \subseteq P \land B \neq \emptyset$ so by the least number principle $B$ has a least element $z$. Also this $z$ is unique.
            
                + Therefore $(\exists u)(u \in A \land u = z)$ because is the least element of $B$. Also because $x \in B$ we have that $(\forall u)(u \in A \implies u \leq z)$
                
                + We conclude that $z \in A \land (\forall u)(u \in A \implies u \leq z)$. Thus, $z$ is a greatest element of $A$
                
## Multiplication

__Theorem __ Consider any Peano system $(P, S, 1)$. There is a unique operation $\times$ on $P$ such that

- $(\gamma) \; x \times 1 = x$ for all $x \in P$
- $(\beta)  \; x \times S(y) = (x \times y) + x$ for all $x \in P \land y \in P$

__Proof__

Let $P = W$, $x = c \in P$ and $x + u = g(u)$ then using the iteration theorem there is a unique function $F:P \longrightarrow P$ such that:

- $F((x,1)) = x$
- $F((x, S(y))) = x + F((x,y))$

Let $F = \times$ and $+((x,y)) \equiv x \times y$ we can rewrite the above expressions as:

- $x \times 1 = x$
- $x \times S(y) = x + (x \times y)$

__Terminology__ $\times$ is called the __multiplication__ operation on $P$. $x \times y$ is called the __product__ of $x$ and $y$

- Also because $S(y) = y + 1$  and the commutativity of addition we have that $x \times (y + 1) = x \times S(y) = x + (x \times y) = (x \times y) + x$

- Furthermore if $P = \mathbb{P}$, $x = 1$ and $S(x) = 2$ then the definition of $\times$ implies:

    + $\begin{split}
        S(1) \times S(1) & = 2 \times S(1) \\
        & = 2 + (2 \times 1) \\
        & = 2 + 2 \\
        & = 4
    \end{split}$
    
__Theorem__ _(Distributive Laws)_

- $x \times (y + z) = (x \times y) + (x \times z)$

    + Let $B = \{z: z \in P \land x \times (y + z) = (x \times y) + (x \times z) \}$
    
        + $B \subseteq P$ because if $z \in B$ then $z \in P$
        
        + $\begin{split}
            x \times (y + 1) & = (x \times y) + x \\
            & = (x \times y) + (x \times 1)
            \end{split}$
        
            + Therefore $1 \in B$
            
        + Let assume $z \in B$
        
            + $\begin{split}
                x \times (y + S(z)) & = x \times S(y + z) \\
                & = x + x \times (y + z) \\
                & = x + ((x \times y) + (x \times z)) \\
                & = (x + (x \times z)) + (x \times y) \\
                & = (x \times S(z)) + (x \times y) \\
                & = (x \times y) + (x \times S(z)) 
                \end{split}$
                
                + Therefore $S(z) \in B$
    
    + Therefore by the principle of mathematical induction $B = P$

- $(y + z) \times x = (y \times x) + (z \times x)$

    + Let $B = \{x: x \in P \land (y + z) \times x = (y \times x) + (z \times x) \}$
    
        + $B \subseteq P$ because if $x \in B$ then $x \in P$
        
        + $\begin{split}
            (y + z) \times 1 & = y + z \\
            & = (y \times 1) + (z \times 1) \\
            \end{split}$
        
            + Therefore $1 \in B$
            
        + Let assume $x \in B$
        
            + $\begin{split}
                (y + z) \times S(x) & = (y + z) + ((y + z) \times x) \\
                & = (y + z) + ((y \times x) + (z \times x)) \\
                & = (y + (y \times x)) + (z + (z \times x)) \\
                & = (y \times S(x)) + (z \times S(x))
                \end{split}$
        
            + Therefore $S(x) \in B$
    
    + Therefore by the principle of mathematical induction $B = P$
    
## Multiplication exercises and solutions

1. Compute $3 \times 2$ (Observe that at this point we have not established the commutativity of multiplication: $x \times y = y \times x$)

    + $\begin{split}
        3 \times 2 & = 3 \times S(1) \\
        & = 3 + (3 \times 1) \\
        & = 3 + 3 \\
        & = 6
    \end{split}$
    
2. Compute $3 \times 3$

    + $\begin{split}
        3 \times 3 & = 3 \times S(2) \\
        & = 3 + (3 \times 2) \\
        & = 3 + 6 \\
        & = 9
    \end{split}$
    
3. Compute $1 \times 3$

    + $\begin{split}
        1 \times 3 & = 3 \times S(2) \\
        & = 3 + (3 \times 2) \\
        & = 3 + 6 \\
        & = 9
    \end{split}$
    
4. Prove that  $2 \times x = x + x$

    + Let $B = \{x: x \in P \land 2 \times x = x + x \}$
    
        + $B \subseteq P$ because if $x \in B$ then $x \in P$
        
        + $\begin{split}
            2 \times 1 & \implies 2 \\
            & \implies S(1) \\
            & \implies 1 + 1 
            \end{split}$
            
            + Therefore $1 \in B$
            
        + Let assume $x \in B$
        
            + $\begin{split}
                2 \times S(x) & \implies 2 \times (x + 1) \\
                & \implies (2 \times x) + (2 \times 1) \\
                & \implies (x + x) + 2 \\
                & \implies (x + x) + S(1) \\
                & \implies (x + x) + (1 + 1) \\
                & \implies (x + 1) + (x + 1) \\
                & \implies S(x) + S(x) \\
                \end{split}$
                
                + Therefore $S(x) \in B$
                
    + Therefore by the principle of mathematical induction $B = P$

5. Prove that  $3 \times x = (x + x) + x$



# Bibliography

