---
title: "The natural numbers"
date: 2020-06-02 23:02:09
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}


```

# The natural numbers

This section is based on [@mendelson_number_2008, Chapter 2, pp 53-85]

- We define the natural numbers as the positive integers, $\mathbb{P} = \{1, 2, 3, \ldots \}$

- In this part we present an axiomatic development of the natural numbers

## Peano systems

By a __peano system__ we mean

- A set $P$

- A particular element $1 \in P$

- A singularity operation $S$ on $P$ $(S: P \longrightarrow P) \land \mathscr{D}(S) = P$

Such that the following axioms are satisfied:

- $(P1) \; 1$ is not the _successor_ $S(x)$ of any object $x \in P$:

    + $(\forall x)(x \in P \implies S(x) \neq 1)$
    
        + In that sense $1 \notin \mathscr{R}(S)$ so $\mathscr{R}(S) \subsetneq P$

- $(P2)$ Different objects in $P$ have different successors:

    + $(\forall x)(\forall y)((x \in P \land y \in P \land x \neq y) \implies S(x) \neq S(y))$
    
        + $S$ is one-one $(S: P \xrightarrow[]{\text{1-1}} P)$

- $(P3)$ _Principle of Mathematical Induction_: Any subset of $P$ containing $1$ and closed under the successor operation must be identical with $P$

    + $(\forall B)((B \subseteq P \land 1 \in B \land (\forall x)(x \in B \implies S(x) \in P)) \implies B = P)$

Such a peano system will be denoted by the ordered triple $(P, S, 1)$

- $P$ is called the __underlying set__

- $S$ the __successor operation__

- $1$ the __distinguished element__

    + The distinguished element $1$ need not to have anything to do with the ordinary integer 1

### Example of peano systems

# Bibliography

