---
title: "The integers"
date: 2020-07-09 18:51:36
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = FALSE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}

```

# The integers

This section is based on [@mendelson_number_2008, Chapter 3, pp 86-155]

- In this part we present a way to construct the integers from the natural numbers, that is from a peano system.

- We may assume the existence of a peano system $(P, S, 1)$ where the elements of $P$ will be called __natural numbers__

## Definition of the integers

- From an intuitive knowledge of the integers we can represent an integer $r$ as the difference $n - k$ of some natural numbers $n$ and $k$. For example:

    + $\begin{split}
        3 &  = 4 - 1 \\
        0 &  = 1 - 1 \\
        -7 & = 1 - 8
        \end{split}$
        
- Of course, the same integer can be represented as the difference of many different pairs:

    + $\begin{split}
        3 &  = 4 - 1 & = 5 - 2 & = 6 - 3 \\
        0 &  = 1 - 1 & = 1 - 1 & = 3 - 3 \\
        -7 & = 1 - 8 & = 2 - 9 & = 3 - 10
        \end{split}$
        
        + It then appears plausible to define the integer $r$ to be the set of all pairs of natural numbers $(n, k)$ such that $r = n - k$
        
            + For example, $-7$ will be the set of all ordered pairs $(n, k)$ of natural numbers such that $-7 = n - k$. However, we cannot proceed directly as indicated because:
            
                + We have not defined $-7$ so the meaning $-7 = n - k$ is still not clear
                
                + We have not defined the substraction operation
                
                    + Therefore we shall avoid these difficulties with the following definition

__Definition__ For any natural numbers $n, k, j$ and $i$ let $$(n, j) \sim (k, i)$$ mean that $$n + i = k + j$$ This defines a relation $\sim$ 

- Examples

    + $(1, 3) \sim (5, 7)$ because $1 + 7 = 5 + 3$
    
    + $(2, 2) \sim (3, 3)$ because $2 + 3 = 3 + 2$
    
    + For any natural numbers $n$ and $j$: $(n, j) \sim (1, 1) \iff n = j$
    
        + $\begin{split}
            (n, j) \sim (1, 1) & \iff n + 1 = 1 + j \\
            & \iff n + 1 = j + 1 \\
            & \iff n = j \\
            \end{split}$
            
    + For any natural numbers $n$ and $j$: $(n, j) \sim (2, 1) \iff n = j + 1$
    
        + $\begin{split}
            (n, j) \sim (2, 1) & \iff n + 1 = 2 + j \\
            & \iff n + 1 = j + 2 \\
            & \iff n + 1 = (j + 1) + 1 \\
            & \iff n = j + 1 
            \end{split}$

From our intuitive knowledge of the integers we see that $(n, j) \sim (k, i)$ corresponds to the equation $n - j = k - i$. In the definition instead of $n - j = k - i$, we have used the equivalent equation $n + i = k + j$, whic involves only natural numbers and the addition operation, $+$, on natural numbers. In this way we are trying to make precise and to justify the intuitive knowledge about the integers using the conception of natural numbers as a peano system. 

__Theorem__ $\sim$ is an equivalence relation in the set $P \times P$ of ordered pairs of natural numbers, that is, for all natural numbers $h, i, j, k, m$ and $n$:

- $(h, i) \sim (h, i)$ (Reflexivity)

    + $h + i = h + i \iff (h, i) \sim (h, i)$
    
- $(h, i) \sim (j, k) \implies (j, k) \sim (h, i)$ (Symmetry)

    + \begin{split}
       (h, i) \sim (j, k) & \iff h + k = j + i \\
       & \iff j + i = h + k \\
       & \iff (j, k) \sim (h, i)
       \end{split}
       
- $((h, i) \sim (j, k) \land (j, k) \sim (m, n)) \implies (h, i) \sim (m, n)$ (Transitivity)

    + $\begin{split}
        (h, i) \sim (j, k) \land (j, k) \sim (m, n) & \implies h + k = j + i \land j + n = m + k \\
        & \implies (h + k) + (j + n) = (j + i) + (j + n) \\
        & \implies (h + k) + (j + n) = (j + i) + (m + k) \\
        & \implies (h + n) + (k + j) = (m + i) + (k + j) \\
        & \implies h + n = m + i \\
        & \implies (h, i) \sim (m, n)
        \end{split}$
        
Therefore $\sim$ is an equivalence relation in $P \times P$ which implies that $\sim$ divides $P \times P$ into a collection of nonempty and mutually exclusive subsets. If $(n, j) \in P \times P$ then the equivalence class of $(n, j)$ relative to the relation $\sim$ is: $$[(n, j)] = \{(k,j): (k,j) \in P \times P \land (n, j) \sim (k,i) \}$$ 

__Examples__

- $\begin{split}
    [(2, 3)] & = \{(k,i): (k,i) \in P \times P \land (2, 3) \sim (k,i) \} \\
    & = \{(k,i): (k,i) \in P \times P \land 2 + i = k + 3 \} \\
    & = \{(1,2), (2,3), (3,4), (4,5), \ldots \}
    \end{split}$
    
- $\begin{split}
    [(4, 4)] & = \{(k,i): (k,i) \in P \times P \land (4, 4) \sim (k,i) \} \\
    & = \{(k,i): (k,i) \in P \times P \land 4 + i = k + 4 \} \\
    & = \{(1,1), (2,2), (3,3), (4,4), \ldots \}
    \end{split}$
    
We let $Z$ denote the set of all equivalence classes with respect to $\sim$. The elements of $Z$ will be called __integers__. We shall use lowercase greek letters $\alpha, \beta, \gamma, \ldots$ as variable for integers. 
    
__Definitions__ $0_Z = [(1,1)]$ and $1_Z = [(2,1)]$

- $\begin{split}
    (n, j) \in 0_Z & \iff (n, j) \in [(1,1)] \\
    & \iff (1, 1) \sim (n,j) \\
    & \iff 1 + j = n + 1 \\
    & \iff j + 1 = n + 1 \\
    & \iff j = n \\
    & \iff n = j 
    \end{split}$
    
    + Thus, $0_Z = \{ (1,1), (2,2), (3,3), \ldots \}$
    
- $\begin{split}
    (n, j) \in 1_Z & \iff (n, j) \in [(2,1)] \\
    & \iff (2, 1) \sim (n,j) \\
    & \iff 2 + j = n + 1 \\
    & \iff (j + 1) + 1 = n + 1 \\
    & \iff j + 1 = n \\
    & \iff n = j + 1 \\
    & \iff n = S(j)
    \end{split}$
    
    + Thus, $1_Z = \{ (2,1), (3,2), (4,3), \ldots \}$
    
- Also because $(1,1) \in 0_Z$ and $1 \neq S(j)$ then $(1,1) \in 1_Z$. Therefore $0_Z \neq 1_Z$

## Addition and multiplication of integers

Returning to our intuitive picture of the integers we have that:

- $3 = 4 - 1$
- $-7 = 1 - 8$

Therefore we can obtain the following:

- $3 + (-7) = (4 + 1) - (1 + 8) = 5 - 9 = -4$

More generally we have the following argument which suggest a definition of addition of integers:

- $r = n - j$
- $s = k - i$
- $r + s = (n + k) - (j + i)$

However we must first confirm the following fact:

__Lemma__ Let $n, j, k, i, n_1, j_1, k_1, i_1$ be natural numbers. If $(n, j) \sim (n_1, j_1)$ and $(k, i) \sim (k_1, i_1)$ then $$(n + k, j + i) \sim (n_1 + k_1, j_1 + i_1)$$

__Proof__ 

- $\begin{split}
    (n, j) \sim (n_1, j_1) \land (k, i) \sim (k_1, i_1) & \iff n + j_1 = n_1 + j \land k + i_1 = k_1 + i \\
    & \iff (n + j_1) + (k + i_1) = (n_1 + j) + (k_1 + i) \\
    & \iff (n + k) + (j_1 + i_1) = (n_1 + k_1) + (j + i) \\
    & \iff (n + k, j + i) \sim (n_1 + k_1, j_1 + i_1) 
    \end{split}$
    
__Example__

# Bibliography