---
title: "The integers"
date: 2020-07-09 18:51:36
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = FALSE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}

```

# The integers

This section is based on [@mendelson_number_2008, Chapter 3, pp 86-155]

- In this part we present a way to construct the integers from the natural numbers, that is from a peano system.

- We may assume the existence of a peano system $(P, S, 1)$ where the elements of $P$ will be called __natural numbers__

## Definition of the integers

- From an intuitive knowledge of the integers we can represent an integer $r$ as the difference $n - k$ of some natural numbers $n$ and $k$. For example:

    + $\begin{split}
        3 &  = 4 - 1 \\
        0 &  = 1 - 1 \\
        -7 & = 1 - 8
        \end{split}$
        
- Of course, the same integer can be represented as the difference of many different pairs:

    + $\begin{split}
        3 &  = 4 - 1 & = 5 - 2 & = 6 - 3 \\
        0 &  = 1 - 1 & = 1 - 1 & = 3 - 3 \\
        -7 & = 1 - 8 & = 2 - 9 & = 3 - 10
        \end{split}$
        
        + It then appears plausible to define the integer $r$ to be the set of all pairs of natural numbers $(n, k)$ such that $r = n - k$
        
            + For example, $-7$ will be the set of all ordered pairs $(n, k)$ of natural numbers such that $-7 = n - k$. However, we cannot proceed directly as indicated because:
            
                + We have not defined $-7$ so the meaning $-7 = n - k$ is still not clear
                
                + We have not defined the substraction operation
                
                    + Therefore we shall avoid these difficulties with the following definition

__Definition__ For any natural numbers $n, k, j$ and $i$ let $$(n, j) \sim (k, i)$$ mean that $$n + i = k + j$$ This defines a relation $\sim$ 

- Examples

    + $(1, 3) \sim (5, 7)$ because $1 + 7 = 5 + 3$
    
    + $(2, 2) \sim (3, 3)$ because $2 + 3 = 3 + 2$
    
    + For any natural numbers $n$ and $j$: $(n, j) \sim (1, 1) \iff n = j$
    
        + $\begin{split}
            (n, j) \sim (1, 1) & \iff n + 1 = 1 + j \\
            & \iff n + 1 = j + 1 \\
            & \iff n = j \\
            \end{split}$
            
    + For any natural numbers $n$ and $j$: $(n, j) \sim (2, 1) \iff n = j + 1$
    
        + $\begin{split}
            (n, j) \sim (2, 1) & \iff n + 1 = 2 + j \\
            & \iff n + 1 = j + 2 \\
            & \iff n + 1 = (j + 1) + 1 \\
            & \iff n = j + 1 
            \end{split}$

From our intuitive knowledge of the integers we see that $(n, j) \sim (k, i)$ corresponds to the equation $n - j = k - i$. In the definition instead of $n - j = k - i$, we have used the equivalent equation $n + i = k + j$, whic involves only natural numbers and the addition operation, $+$, on natural numbers. In this way we are trying to make precise and to justify the intuitive knowledge about the integers using the conception of natural numbers as a peano system. 

__Theorem__ $\sim$ is an equivalence relation in the set $P \times P$ of ordered pairs of natural numbers, that is, for all natural numbers $h, i, j, k, m$ and $n$:

- $(h, i) \sim (h, i)$ (Reflexivity)

    + $h + i = h + i \iff (h, i) \sim (h, i)$
    
- $(h, i) \sim (j, k) \implies (j, k) \sim (h, i)$ (Symmetry)

    + \begin{split}
       (h, i) \sim (j, k) & \iff h + k = j + i \\
       & \iff j + i = h + k \\
       & \iff (j, k) \sim (h, i)
       \end{split}
       
- $((h, i) \sim (j, k) \land (j, k) \sim (m, n)) \implies (h, i) \sim (m, n)$ (Transitivity)

    + $\begin{split}
        (h, i) \sim (j, k) \land (j, k) \sim (m, n) & \implies h + k = j + i \land j + n = m + k \\
        & \implies (h + k) + (j + n) = (j + i) + (j + n) \\
        & \implies (h + k) + (j + n) = (j + i) + (m + k) \\
        & \implies (h + n) + (k + j) = (m + i) + (k + j) \\
        & \implies h + n = m + i \\
        & \implies (h, i) \sim (m, n)
        \end{split}$
        
Therefore $\sim$ is an equivalence relation in $P \times P$ which implies that $\sim$ divides $P \times P$ into a collection of nonempty and mutually exclusive subsets. If $(n, j) \in P \times P$ then the equivalence class of $(n, j)$ relative to the relation $\sim$ is: $$[(n, j)] = \{(k,j): (k,j) \in P \times P \land (n, j) \sim (k,i) \}$$ 

__Examples__

- $\begin{split}
    [(2, 3)] & = \{(k,i): (k,i) \in P \times P \land (2, 3) \sim (k,i) \} \\
    & = \{(k,i): (k,i) \in P \times P \land 2 + i = k + 3 \} \\
    & = \{(1,2), (2,3), (3,4), (4,5), \ldots \}
    \end{split}$
    
- $\begin{split}
    [(4, 4)] & = \{(k,i): (k,i) \in P \times P \land (4, 4) \sim (k,i) \} \\
    & = \{(k,i): (k,i) \in P \times P \land 4 + i = k + 4 \} \\
    & = \{(1,1), (2,2), (3,3), (4,4), \ldots \}
    \end{split}$
    
We let $Z$ denote the set of all equivalence classes with respect to $\sim$. The elements of $Z$ will be called __integers__. We shall use lowercase greek letters $\alpha, \beta, \gamma, \ldots$ as variable for integers. 
    
__Definitions__ $0_\mathbb{Z} = [(1,1)]$ and $1_\mathbb{Z} = [(2,1)]$

- $\begin{split}
    (n, j) \in 0_\mathbb{Z} & \iff (n, j) \in [(1,1)] \\
    & \iff (1, 1) \sim (n,j) \\
    & \iff 1 + j = n + 1 \\
    & \iff j + 1 = n + 1 \\
    & \iff j = n \\
    & \iff n = j 
    \end{split}$
    
    + Thus, $0_\mathbb{Z} = \{ (1,1), (2,2), (3,3), \ldots \}$
    
- $\begin{split}
    (n, j) \in 1_\mathbb{Z} & \iff (n, j) \in [(2,1)] \\
    & \iff (2, 1) \sim (n,j) \\
    & \iff 2 + j = n + 1 \\
    & \iff (j + 1) + 1 = n + 1 \\
    & \iff j + 1 = n \\
    & \iff n = j + 1 \\
    & \iff n = S(j)
    \end{split}$
    
    + Thus, $1_Z = \{ (2,1), (3,2), (4,3), \ldots \}$
    
- Also because $(1,1) \in 0_Z$ and $1 \neq S(j)$ then $(1,1) \in 1_Z$. Therefore $0_Z \neq 1_Z$

## Addition and multiplication of integers

Returning to our intuitive picture of the integers we have that:

- $3 = 4 - 1$
- $-7 = 1 - 8$

Therefore we can obtain the following:

- $3 + (-7) = (4 + 1) - (1 + 8) = 5 - 9 = -4$

More generally we have the following argument which suggest a definition of addition of integers:

- $r = n - j$
- $s = k - i$
- $r + s = (n + k) - (j + i)$

However we must first confirm the following fact:

__Lemma__ Let $n, j, k, i, n_1, j_1, k_1, i_1$ be natural numbers. If $(n, j) \sim (n_1, j_1)$ and $(k, i) \sim (k_1, i_1)$ then $$(n + k, j + i) \sim (n_1 + k_1, j_1 + i_1)$$

__Proof__ 

- $\begin{split}
    (n, j) \sim (n_1, j_1) \land (k, i) \sim (k_1, i_1) & \iff n + j_1 = n_1 + j \land k + i_1 = k_1 + i \\
    & \iff (n + j_1) + (k + i_1) = (n_1 + j) + (k_1 + i) \\
    & \iff (n + k) + (j_1 + i_1) = (n_1 + k_1) + (j + i) \\
    & \iff (n + k, j + i) \sim (n_1 + k_1, j_1 + i_1) 
    \end{split}$
    
__Example__

- We know that $(1,2) \sim (3,4)$ and $(2,2) \sim (3,3)$. Hence by the above lemma $(1 + 2,2 + 2) \sim (3 + 3,4 + 3)$. Thus $(3,4) \sim (6,7)$ which is correct because $3 + 7 = 6 + 4$

The above lemma allows us to make the following definition

__Definition__ Let $\alpha$ and $\beta$ be integers. Choose any element $(n, j) \in \alpha$ and choose any elements $(k, i) \in \beta$. Then $\alpha +_\mathbb{Z} \beta$ stands for $[(n + k, j + i)]$.

The reason that the above lemma is required for this definition to make sense is that we must be sure that our particular choice of pairs $(n, j) \in \alpha$ and $(k, i) \in \beta$ does not affect the outcome $\alpha +_\mathbb{Z} \beta$

Assume that $(n_1, j_1) \in \alpha$ and $(k_1, i_1) \in \beta$. Therefore $(n, j) \sim (n_1, j_1)$ and $(k, i) \sim (k_1, i_1)$ so by the lemma $(n + k, j + i) \sim (n_1 + k_1, j_1 + i_1)$. This implies that $[(n + k, j + i)] = [(n_1 + k_1, j_1 + i_1)]$

__Theorem__ _(Properties of Addition)_ Consider any $\alpha, \beta, \gamma$ in $\mathbb{Z}$

- $\alpha +_\mathbb{Z} \beta = \beta +_\mathbb{Z} \alpha$ (Commutativity)

    + $\begin{split}
        \alpha +_\mathbb{Z} \beta & \iff [(n + k, j + i)] \text{ where } (n,j) \in \alpha \land (k, i) \in \beta \\
        & \iff [(k + n, i + j)] \text{ where } (k, i) \in \beta \land (n,j) \in \alpha \\
        & \iff \beta +_\mathbb{Z} \alpha
        \end{split}$
        
- $\alpha +_\mathbb{Z} (\beta +_\mathbb{Z} \gamma)= (\alpha +_\mathbb{Z} \beta) +_\mathbb{Z} \gamma$ (Associativity)

    + $\begin{split}
        \alpha +_\mathbb{Z} (\beta +_\mathbb{Z} \gamma) & \iff [(n + (k + m), j + (i + h))] \text{ where } (n, j) \in \alpha \land (k + m, i + h) \in \beta +_\mathbb{Z} \gamma \\
        & \iff [((n + k) + m, (j + i) + h)] \text{ where } (n + k, j + i) \in \alpha +_\mathbb{Z} \beta \land (m, h) \in \gamma \\
        & \iff (\alpha +_\mathbb{Z} \beta) +_\mathbb{Z} \gamma
        \end{split}$
        
- $\alpha +_\mathbb{Z} 0_\mathbb{Z} = \alpha$

    + $\begin{split}
        \alpha +_\mathbb{Z} 0_\mathbb{Z} & \iff [(n + 1, j + 1)] \text{ where } (n, j) \in \alpha \land (1,1) \in 0_\mathbb{Z} \\
        & \implies (n, j) \in [(n + 1, j + 1)] \text{ because } (n + 1, j + 1) \sim (n, j) \\
        & \implies (n, j) \in [(n + 1, j + 1)] \text{ because } (n + 1) + j = n + (j + 1) \\
        & \implies [(n + 1, j + 1)] = [(n,j)] \\
        & \implies \alpha +_\mathbb{Z} 0_\mathbb{Z} = \alpha
        \end{split}$
        
- $(\exists ! \delta)(\alpha +_\mathbb{Z} \delta = 0_\mathbb{Z})$ (This unique $\delta$ is denoted $-_\mathbb{Z}\alpha$). If $(h, i) \in \alpha$ then $(i, h) \in -_\mathbb{Z}\alpha$

    + If $(n, j) \in \alpha$ then let $[(j,n)] = \delta$. Therefore:
    
        + $\begin{split}
            \alpha +_\mathbb{Z} \delta & \iff [(n + j, j + n)] \\
            & \implies (1,1) \in [(n + j, j + n)] \text{ because } (n + j, j + n) \sim (1, 1) \\
            & \implies (1,1) \in [(n + j, j + n)] \text{ because } (n + j) + 1 = (j + n) +1 \\
            & \implies [(n + j, j + n)] = [(1,1)] \\
            & \implies \alpha +_\mathbb{Z} \delta = 0_\mathbb{Z}
            \end{split}$
            
    + Assume $\alpha +_\mathbb{Z} \delta = 0_\mathbb{Z}$ and $\alpha +_\mathbb{Z} \beta = 0_\mathbb{Z}$
    
        + $\begin{split}
            \delta & = \delta +_\mathbb{Z} 0_\mathbb{Z} \\
            & = \delta +_\mathbb{Z} (\alpha +_\mathbb{Z} \beta) \\
            & = (\delta +_\mathbb{Z} \alpha) +_\mathbb{Z} \beta \\
            & = (\alpha +_\mathbb{Z} \delta) +_\mathbb{Z} \beta \\
            & = 0_\mathbb{Z} +_\mathbb{Z} \beta \\
            & = \beta +_\mathbb{Z} 0_\mathbb{Z}\\
            & = \beta
            \end{split}$
            
Multiplication is handled in much the same way as addition. First we look at the intuitive picture for hints as how to proceed. If $r = n - j$ and $s = k - i$ then $$r \times s = (n - j) \times (k -i) = ((n \times k) + (j \times i)) - ((j \times k) + (n \times i))$$ 

This gives us the idea for the multiplication rule.

__Lemma__ Let $n, j, k, i, n_1, j_1, k_1, i_1$ be natural numbers. If $(n, j) \sim (n_1, j_1)$ and $(k, i) \sim (k_1, i_1)$ then $$(n \times k + j \times i, j \times k + n \times i) \sim (n_1 \times k_1 + j_1 \times i_1, j_1 \times k_1 + n_1 \times i_1)$$

We need to prove that 

$$\underbrace{(n \times k + j \times i) + (j_1 \times k_1 + n_1 \times i_1)}_\text{A} = \underbrace{(n_1 \times k_1 + j_1 \times i_1) + (j \times k + n \times i)}_\text{B}$$

We have that:

- $(i) \; n + j_1 = n_1 + j$
- $(ii) \; k + i_1 = k_1 + i$

We proceed in the following way:

- Multiply $(i)$ by $k$

    + $(iii) \; n \times k + j_1 \times k = n_1 \times k + j \times k$
    
- Multiply $(ii)$ by $j_1$ and reverse the two sides of the equality

    + $(iv) \; j_1 \times k_1 + j_1 \times i = j_1 \times k + j_1 \times i_1$
    
- Multiply $(i)$ by $i$ and reverse the two sides of the equality

    + $(v) \; n_1 \times i + j \times i = n \times i + j_1 \times i$
    
- Multiply $(ii)$ by $n_1$

    + $(vi) \; n_1 \times k + n_1 \times i_1 = n_1 \times k_1 + n_1 \times i$
    
- Adding $(iii), (iv), (v), (vi)$

$$\begin{split}
   (n \times k + j_1 \times k) + (j_1 \times k_1 + j_1 \times i) + (n_1 \times i + j \times i) + (n_1 \times k + n_1 \times i_1) & = (n_1 \times k + j \times k) + (j_1 \times k + j_1 \times i_1) + (n \times i + j_1 \times i) + (n_1 \times k_1 + n_1 \times i) \\
   (n \times k + j \times i) + (j_1 \times k_1 + n_1 \times i_1) + (j_1 \times i + j_1 \times k) + (n_1 \times k + n_1 \times i) & = (n_1 \times k_1 + j_1 \times i_1) + (j \times k + n \times i) + (j_1 \times k + j_1 \times i) + (n_1 \times k + n_1 \times i) \\
   \underbrace{(n \times k + j \times i) + (j_1 \times k_1 + n_1 \times i_1)}_\text{A} + \underbrace{(j_1 \times i + j_1 \times k) + (n_1 \times k + n_1 \times i)}_\text{C} & = \underbrace{(n_1 \times k_1 + j_1 \times i_1) + (j \times k + n \times i)}_\text{B} + \underbrace{(j_1 \times i + j_1 \times k) + (n_1 \times k + n_1 \times i)}_\text{C} \\
   A + C & = B + C \\
   A & = B \\
   (n \times k + j \times i, j \times k + n \times i) & \sim (n_1 \times k_1 + j_1 \times i_1, j_1 \times k_1 + n_1 \times i_1)
   \end{split}$$

__Definition__ Given $\alpha$ and $\beta$ in $\mathbb{Z}$ take any $(n,j) \in \alpha$ and any $(k, i) \in \beta$. Let $\alpha \times_{\mathbb{Z}} \beta$ stand for $[(n \times k + j \times i, j \times k + n \times i)]$

__Theorem__ _(Properties of Multiplication)_ For any $\alpha, \beta, \gamma$ in $\mathbb{Z}$

- $\alpha \times_{\mathbb{Z}} \beta = \beta \times_{\mathbb{Z}} \alpha$ (Commutativity)

    + $\begin{split}
        \alpha \times_{\mathbb{Z}} \beta & = [(n,j)] \times_{\mathbb{Z}} [(k,i)] \\
        & = [(n \times k + j \times i, j \times k + n \times i)] \\
        & = [(k \times n + i \times j, k \times j + i \times n)] \\
        & = [(k \times n + i \times j, i \times n + k \times j)] \\
        & = \beta \times_{\mathbb{Z}} \alpha
        \end{split}$

- $\alpha \times_{\mathbb{Z}} ( \beta \times_{\mathbb{Z}} \gamma) = (\alpha \times_{\mathbb{Z}} \beta) \times_{\mathbb{Z}} \gamma$ (Associativity)

    + $\begin{split}
        \alpha \times_{\mathbb{Z}} ( \beta \times_{\mathbb{Z}} \gamma) & = [(n,j)] \times_{\mathbb{Z}} ([(k,i)] \times_{\mathbb{Z}} [(m,h)]) \\
        & = [(n,j)] \times_{\mathbb{Z}} ([(k \times m + i \times h,i \times m + k \times h)] \\
        & = [(n(k \times m + i \times h) + j(i \times m + k \times h), j(k \times m + i \times h) + n(i \times m + k \times h))] \\
        & = [(n(k \times m) + n(i \times h) + j(i \times m) + j(k \times h), j(k \times m) + j(i \times h) + n(i \times m) + n(k \times h))] \\
        & = [((n \times k + j \times i)m + (n \times i + j \times k)h, (j \times k + n \times i)m + (j \times i + n \times k)h)] \\
        & = [((n \times k + j \times i)m + (j \times k + n \times i)h, (j \times k + n \times i)m + (n \times k + j \times i)h)] \\
        & = [(n \times k + j \times i, j \times k + n \times i)] \times_\mathbb{Z} [(m, h)] \\
        & = ([(n, j)] \times_\mathbb{Z} [(k, i)]) \times_\mathbb{Z} [(m, h)] \\
        & = (\alpha \times_\mathbb{Z} \beta) \times_\mathbb{Z} \gamma
        \end{split}$
        
- $\alpha \times_\mathbb{Z} (\beta +_\mathbb{Z} \gamma) = (\alpha \times_\mathbb{Z} \beta) +_\mathbb{Z} (\alpha \times_\mathbb{Z} \gamma)$ (Distributivity)

    + $\begin{split}
        \alpha \times_\mathbb{Z} (\beta +_\mathbb{Z} \gamma) & = [(n,j)] \times_\mathbb{Z} ([(k,i)] +_\mathbb{Z} [(m, h)]) \\
        & = [(n,j)] \times_\mathbb{Z} [(k + m, i + h)] \\
        & = [(n(k + m) + j(i + h), j(k + m) + n(i + h))] \\
        & = [(n \times k + n \times m + j \times i + j \times h, j \times k + j \times m + n \times i + n \times h)] \\
        & = [(n \times k + j \times i + n \times m + j \times h, j \times k + n \times i + j \times m + n \times h)] \\
        & = [(n \times k + j \times i, j \times k + n \times i)] +_\mathbb{Z} [(n \times m + j \times h, j \times m + n \times h)] \\
        & = ([(n, j)] \times_{\mathbb{Z}} [(k, i)]) +_\mathbb{Z} ([(n,j)] \times_{\mathbb{Z}} [(m, h)]) \\
        & = (\alpha \times_{\mathbb{Z}} \beta ) +_\mathbb{Z} (\alpha  \times_{\mathbb{Z}} \gamma)
        \end{split}$

- $\alpha \times_\mathbb{Z} 1_{\mathbb{Z}} = \alpha$

    + $\begin{split}
        \alpha \times_\mathbb{Z} 1_{\mathbb{Z}} & = [(n, j)] \times_\mathbb{Z} [(2,1)] \\
        & = [(n \times 2 + j \times 1, j \times 2 + n \times 1)] \\
        & = [(2 \times n + j , 2 \times j + n)] \\
        & = [(n, j)] \text{ because } (2 \times n + j , 2 \times j + n) \sim (n, j) \\
        & = [(n, j)] \text{ because } 2 \times n + 2 \times j = 2 \times j + 2 \times n \\
        & = \alpha
        \end{split}$
        
- $[\alpha \neq 0_{\mathbb{Z}} \land \beta \neq 0_{\mathbb{Z}}] \implies \alpha \times_{\mathbb{Z}} \beta \neq 0_{\mathbb{Z}}$

    + $\begin{split}
        \alpha \neq 0_{\mathbb{Z}} \land \beta \neq 0_{\mathbb{Z}} & = [(n, j)] \land [(k, i)] \text{ where } n \neq j \land k \neq i \\
        & \implies [(n, j)] \land (k < i \lor k > i) \\
        & \implies [(n, j)] \land (i = k + w \lor k = i + u) \\
        & \implies ((n, j) \times_\mathbb{Z} (k, k + w)) \lor ((n, j) \times_\mathbb{Z} (i + u, i)) \\
        & \implies (n \times k + j \times (k + w), j \times k + n \times (k + w)) \lor (n \times (i + u) + j \times i, j \times (i + u) + n \times i) \\
        & \implies (n \times k + j \times k + j \times w, j \times k + n \times k + n \times w) \lor (n \times i + n \times u + j \times i, j \times i + j \times u + n \times i) \\
        & \implies (n \times k + j \times k + j \times w, n \times k + j \times k + n \times w) \lor (n \times i + j \times i + n \times u, n \times i + j \times i + j \times u) \\
        & \implies ((n + j) \times k + j \times w, (n + j) \times k + n \times w) \lor ((n + j) \times i + n \times u, (n + j) \times i + j \times u) \\
        & \implies (((n + j) \times k + j \times w, (n + j) \times k + n \times w) \text{ where } (j \neq n \implies j \times w \neq n \times w)) \lor (((n + j) \times i + n \times u, (n + j) \times i + j \times u) \text{ where } (n \neq j \implies n \times u \neq j \times u)) \\
        & \implies \alpha \times_{\mathbb{Z}} \beta \neq 0_{\mathbb{Z}} \lor \alpha \times_{\mathbb{Z}} \beta \neq 0_{\mathbb{Z}} \\
        & \implies \alpha \times_{\mathbb{Z}} \beta \neq 0_{\mathbb{Z}}
        \end{split}$

## Rings and integral domains

We shall study class of algebraic structures, of which the integers form just one example, and we shall prove a large number of theorems about such structures. All of these theorems will, in particular, hold true for the integers, but they will also be applicable to many other mathematical structures.

__Definition__ By a __ring__ we mean a triple $(R, +, \times)$ such that $R$ is a set and $+$ and $\times$ are binary operations in $R$ satisfying:

- $x + (y + z) = (x + y) + z$ (Associativity of $+$)

- $x + y = y + x$ (Commutativity of $+$)

- There is an element $0$ in $R$ such that

    + $(\forall x)(x + 0 = x)$
    
        + Such and element $0$ is unique and is called the __zero element__ of the ring. For, if there were another such element $\square$, $(\forall x)(x + \square = x)$,then: $$\begin{split}
   \square & = \square + 0 \\
   & = 0 + \square \\
   & = 0
   \end{split}$$
    
    + $(\forall x)(\exists y)(x + y = 0)$
   
        + Such and element $y$ is denoted $-x$ and is called the __additive inverse__ of x. Also in $(\forall x)(\exists y)(x + y = 0)$ implies that $y$ is unique such that $(\forall x)(\exists !y)(x + y = 0)$, because if $x + y = 0 = x + z$: $$\begin{split}
   y & = 0 + y \\
   & = (x + z) + y \\
   & = x + (z + y) \\
   & = x + (y + z) \\
   & = (x + y) + z \\
   & = 0 + z \\
   & = z 
   \end{split}$$
   
- $x \times (y \times z) = (x \times y) \times z$ (Associativity of $\times$)

- $\begin{split}
    x \times (y + z) & = x \times y + x \times z \\
    (y + z) \times x & = y \times x + z \times x 
    \end{split}$ (Distributivity of $\times$ with respect to $+$)
   
__Definition__ A __ring with unit element__ is a ring $(R, + , \times)$ such that:

- There is an element $1 \in R$ such that:

    + $(\forall x)(x = x \times 1 = 1 \times x)$
    
        + Such an element $1$ is unique. For, if there were another element $1^*$ such that $(\forall x)(x = x \times 1^* = 1^* \times x)$. Then $1^* = 1^* \times 1 = 1$. The unique element $1$ is called the __unit element__ of $R$
        
__Definition__ A ring $(R, +, \times)$ is __commutative__ if and only if:

- $(\forall x)(\forall y)(x \times y = y \times x)$

__Examples__

1. The system of integers $(\mathbb{Z}, +_{\mathbb{Z}}, \times_{\mathbb{Z}})$ is a commutative ring with unit element

2. The rational numbers under ordinary addition and multiplication form a commutative ring with unit element

3. The real numbers under ordinary addition and multiplication form a commutative ring with unit element

4. The even integers under ordinary addition and multiplication form a commutative ring without a unit element. Take into account that there is no even integer $u$ such that $x \times u = x$ for all even integers $x$.

5. Let $R$ be the set of all $2 \times 2$ matrices $\begin{pmatrix}
                                                     a & b \\
                                                     c & d
                                                     \end{pmatrix}$, that is, $a, b, c, d$ are integers. Let $$\begin{pmatrix}
       a & b \\
       c & d
       \end{pmatrix} + 
      \begin{pmatrix}
       a^* & b^* \\
       c^* & d^*
       \end{pmatrix} = 
      \begin{pmatrix}
       a + a^* & b + b^* \\
       c + c^* & d + d^*
       \end{pmatrix}$$ and $$\begin{pmatrix}
                              a & b \\
                              c & d
                              \end{pmatrix} \times
                             \begin{pmatrix}
                             a^* & b^* \\
                             c^* & d^*
                              \end{pmatrix} =
                             \begin{pmatrix}
                             aa^* + bc   & ab^* + bd^* \\
                             ca^* + dc^* & cb^* + dd^*
                              \end{pmatrix}$$ where $(R, +, \times)$ is a noncommutative ring with unit element. The unit element is $\begin{pmatrix}
                               1 & 0 \\
                               0 & 1
                               \end{pmatrix}$. To show noncommutativity, consider $\begin{pmatrix}
                                                                                    1 & 0 \\
                                                                                    0 & 0
                                                                                    \end{pmatrix}$ and $\begin{pmatrix}
                                                                                                         0 & 1 \\
                                                                                                         0 & 0
                                                                                                         \end{pmatrix}$ where $$\begin{pmatrix}
                        1 & 0 \\
                        0 & 0
                        \end{pmatrix} \times
                       \begin{pmatrix}
                        0 & 1 \\
                        0 & 0
                        \end{pmatrix} = 
                       \begin{pmatrix}
                        1 & 1 \\
                        0 & 0
                        \end{pmatrix}$$ and $$\begin{pmatrix}
                                               0 & 1 \\
                                               0 & 0
                                               \end{pmatrix} \times
                                              \begin{pmatrix}
                                               1 & 0 \\
                                               0 & 0
                                               \end{pmatrix} =
                                              \begin{pmatrix}
                                               0 & 0 \\
                                               0 & 0
                                               \end{pmatrix}$$

__Theorem__ _(Properties of Rings)_

- $x + y = x + z \implies y = z$

    + $\begin{split}
        x + y = x + z & \implies -x + (x + y) = -x + (x + z) \\
        & \implies (-x + x) + y = (-x + x) + z \\
        & \implies 0 + y = 0 + z \\
        & \implies y = z
        \end{split}$
        
- $-0 = 0$

    + $\begin{split}
        -0 & = -0 + 0 \\
        & = 0 + -0 \\
        & = 0
        \end{split}$
        
- $-(-x) = x$

    + $\begin{split}
        -(-x) & = -(-x) + 0 \\
        & = -(-x) + (x + -x) \\
        & = -(-x) + (-x + x) \\
        & = (-(-x) + -x) + x \\
        & = (-x + -(-x)) + x \\
        & = 0 + x \\
        & = x + 0 \\
        & = x
        \end{split}$
        
- $-x = -y \implies x = y$

    + $\begin{split}
        x + -x & = 0 \\
        x + -y & = 0 \\
        (x + -y) + y & = 0 + y \\
        x + (-y + y) & = y + 0 \\
        x + (y + -y) & = y \\
        x + 0 & = y \\
        x & = y
        \end{split}$
        
- $-(x + y) = (-x) + (-y)$

    + $(x + y) + -(x + y) = 0$
    
    + $\begin{split}
        (x + y) + (-x + -y) & = (x + y) + (-y + -x) \\
        & = ((x + y) + -y) + -x \\
        & = (x + (y + -y)) + -x \\
        & = (x + 0) + -x \\
        & = x + -x \\
        & = 0
        \end{split}$
    
        + Therefore $-(x + y) = -x + -y$
        
- $x + y = x \implies y = 0$

    + $\begin{split}
        x + y & = x \\
        x + y & = x + 0 \\
        y & = 0
        \end{split}$

- $x \times 0 = 0 = 0 \times x$

    + $\begin{split}
        x \times 0 & = x \times (0 + 0) \\
        & = x \times 0 + x \times 0 \\
        & \implies x \times 0 = 0
        \end{split}$
        
    + $\begin{split}
        0 \times x & = (0 + 0) \times x \\
        & = 0 \times x + 0 \times x \\
        & \implies 0 \times x = 0
        \end{split}$
        
    + Therefore $x \times 0 = 0 \times x$
    
- $-(x \times y) = (-x) \times y = x \times (-y)$

    + $x \times y + -(x \times y) = 0$
    
    + $\begin{split}
        x \times y + -x \times y & = (x  + -x) \times y \\
        & = 0 \times y \\
        & = 0
        \end{split}$
        
    + $\begin{split}
        x \times y + x \times -y & = x \times (y + -y) \\
        & = x \times 0 \\
        & = 0
        \end{split}$
        
        + Therefore $-(x \times y) = -x \times y = x \times -y$

- $(-x) \times (-y) = x \times y$

    + $\begin{split}
        x \times y + -x \times y & = 0 \\
        -x \times y + x \times y & = 0
        \end{split}$
    
    + $\begin{split}
        -x \times y + -x \times -y & = -x \times (y + -y) \\
        & = -x \times (y + -y) \\
        & = -x \times 0 \\
        & = 0
        \end{split}$
        
        + Therefore $x \times y = -x \times -y$
        
__Definition__ In any ring let $x - y = x + (-y)$. The operation $-$ is called _substraction_

__Theorem__ _(Properties of Substraction)_

- $x - x = 0$

    + $\begin{split}
        x - x & = x + - x \\
        & = 0
        \end{split}$
        
- $-(x - y) = y - x$

    + $(x - y) + -(x - y) = 0$
    
    + $\begin{split}
        (x - y) + (y - x) & = (x - y) + (y + -x) \\
        & = ((x - y) + y) + -x \\
        & = ((x + - y) + y) + -x \\
        & = (x + (- y + y)) + -x \\
        & = (x + (y + -y)) + -x \\
        & = (x + 0) + -x \\
        & = x + -x \\
        & = 0
        \end{split}$
        
        + Therefore $-(x - y) = y - x$
        
- $x - y = 0 \iff x = y$

    + $\begin{split}
        x - y & = 0 \\
        x + -y & = 0 \\
        (x + -y) + y & = 0 + y \\
        x + (-y + y) & = y + 0 \\
        x + (y + -y) & = y + 0 \\
        x + 0 & = y + 0 \\
        x & = y
        \end{split}$
        
- $\begin{split}
    x \times (y - z) & = (x \times y) - (x \times z) \\
    (y - z) \times x & = (y \times x) - (z \times x)
    \end{split}$ (Distributivity of $\times$ over $-$)
    
    + $\begin{split}
        x \times (y - z) & = x \times (y + -z) \\
        & = x \times y + x \times -z \\
        & = x \times y - x \times z
        \end{split}$
        
    + $\begin{split}
        (y - z) \times x & = (y + -z) \times x \\
        & = y \times x + -z \times x \\
        & = y \times x - z \times x
        \end{split}$
        
- In a ring with unit element $(-1) \times x = -x = x \times (-1)$ and $(-1) \times (-1) = 1$

    + $\begin{split}
        -1 \times x & = 1 \times -x \\
        & = -(1 \times x) \\
        & = -x
        \end{split}$
        
    + $\begin{split}
        -1 \times -1 & = 1 \times 1 \\
        & = 1
        \end{split}$
        
## Rings and integral domains exercises and solutions

1. In any ring, prove $(\forall x)(\forall y)(\exists !z)(x + z = y)$

    + Assume $x + z = y$ and $x + w = y$ therefore $x + z = x + w$ which implies that $z = w$. Therefore $z$ is unique
    
2. In a ring, is $+$ distributive over $\times$, that is does $x + (y \times z) = (x + y) \times (y + z)$ hold for all $x, y, z$? (Hint: Consider the ring of integers)

    + Let $x = 1, y = 1$ and $z = 1$ then if $+$ distributive over $\times$ holds we will have that $2 = 1 + (1 \times 1) = (1 + 1) \times (1 + 1) = 4$ which is false
    
3. Prove $(x - y) + (y - z) = x - z$

    + $\begin{split}
        (x - y) + (y - z) & = (x + -y) + (y + -z) \\
        & = ((x + -y) + y) + -z \\
        & = (x + (y + -y)) + -z \\
        & = (x + 0) + -z \\
        & = x + -z \\
        & = x - z
        \end{split}$
        
4. Prove $0 - x = -x$

    + $\begin{split}
        0 - x & = 0 + -x \\
        & = -x + 0 \\
        & = -x
        \end{split}$

5. Prove $x + (y - x) = y$

    + $\begin{split}
        x + (y - x) & = x + (y + -x) \\
        & = x + (-x + y) \\
        & = (x + -x) + y \\
        & = 0 + y \\
        & = y + 0 \\
        & = y
        \end{split}$

6. Prove $(y + z) - (x + z) = y - x$

    + $\begin{split}
        (y + z) - (x + z) & = (y + z)  + -(x + z) \\
        & = (y + z) + (-z + -x) \\
        & = ((y + z) + -z) + -x \\
        & = (y + (z + -z)) + -x \\
        & = (y + 0) + -x \\
        & = y + -x \\
        & = y - x
        \end{split}$

__Definition__ By and __integral domain__ we mean a commutative ring with unit element $(R, +, \times)$ satisfying the following additional axiom:

- $[x \neq 0 \land y \neq 0] \implies x \times y \neq 0$ for any $x$ and $y$ in $R$

    + An equivalent form is $x \times y = 0 \implies [x = 0 \lor y = 0]$
    
__Terminology__ By a __zero divisor__ we mean a nonzero element $x$ for which there exists some nonzero element $y$ such that $x \times y = 0$. Thus, the previous axiom is equivalent to saying that $R$ contains no zero divisors

__Theorem__ For a commutative ring with unit element, the prevously axiom is equivalent to:

- For any $x, y, z$ in $R$, $[x \times y =  x \times z \land x \neq 0] \implies y = z$ (Cancellation Law for $\times$)

    + $\begin{split}
        \end{split}$

# Bibliography