---
title: "The integers"
date: 2020-07-09 18:51:36
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = FALSE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}

```

# The integers

This section is based on [@mendelson_number_2008, Chapter 3, pp 86-155]

- In this part we present a way to construct the integers from the natural numbers, that is from a peano system.

- We may assume the existence of a peano system $(P, S, 1)$ where the elements of $P$ will be called __natural numbers__

- From an intuitive knowledge of the integers we can represent $r$ as the difference $n - k$ of some natural numbers $n$ and $k$. For example:

    + $\begin{split}
        3 &  = 4 - 1 \\
        0 &  = 1 - 1 \\
        -7 & = 1 - 8
        \end{split}$
        
- Of course, the same integer can be represented as the difference of many different pairs:

    + $\begin{split}
        3 &  = 4 - 1 & = 5 - 2 & = 6 - 3 \\
        0 &  = 1 - 1 & = 1 - 1 & = 3 - 3 \\
        -7 & = 1 - 8 & = 2 - 9 & = 3 - 10
        \end{split}$
        
        + It then appears plausible to define the integer $r$ to be the set of all pairs of natural numbers $(n, k)$ such that $r = n - k$
        
            + For example, $-7$ will be the set of all ordered pairs $(n, k)$ of natural numbers such that $-7 = n - k$. However, we cannot proceed directly as indicated because:
            
                + We have not defined $-7$ so the meaning $-7 = n - k$ is still not clear
                
                + We have not defined the substraction operation
                
                    + Therefore we shall avoid these difficulties with the following definition

__Definition__ For any natural numbers $n, k, j$ and $i$ let $$(n, j) \sim (k, i)$$ mean that $$n + i = k + j$$ This defines a relation $\sim$ 

- Examples

    + $(1, 3) \sim (5, 7)$ because $1 + 7 = 5 + 3$
    
    + $(2, 2) \sim (3, 3)$ because $2 + 3 = 3 + 2$
    
    + For any natural numbers $n$ and $j$: $(n, j) \sim (1, 1) \iff n = j$
    
        + $\begin{split}
            (n, j) \sim (1, 1) & \iff n + 1 = 1 + j \\
            & \iff n + 1 = j + 1 \\
            & \iff n = j \\
            \end{split}$
            
    + For any natural numbers $n$ and $j$: $(n, j) \sim (2, 1) \iff n = j + 1$
    
        + $\begin{split}
            (n, j) \sim (2, 1) & \iff n + 1 = 2 + j \\
            & \iff n + 1 = j + 2 \\
            & \iff n + 1 = (j + 1) + 1 \\
            & \iff n = j + 1 
            \end{split}$

__Theorem__ $\sim$ is an equivalence relation in the set $P \times P$ of ordered pairs of natural numbers, that is, for all natural numbers $h, i, j, k, m$ and $n$:

- $(h, i) \sim (h, i)$

    + $h + i = h + i \iff (h, i) \sim (h, i)$
    
- $(h, i) \sim (j, k) \implies (j, k) \sim (h, i)$

    + \begin{split}
       (h, i) \sim (j, k) & \iff h + k = j + i \\
       & \iff j + i = h + k \\
       & \iff (j, k) \sim (h, i)
       \end{split}
       
- 

# Bibliography