---
title: "Appendix A Equality"
date: 2020-04-10 18:21:56
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
---

<style>

body {
text-align: justify}

</style>

```{r libraries}


```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

# Equality in mathematics

This section is based on [@mendelson_number_2008, Appendix A, pp 279-280]

In almost all mathematical theories equality is taken to be a primitive undefined relation

## Properties

__Assumption I__  $(\forall x)(x = x) \text{ (Reflexivity)}$

__Assumption II__ $(\forall x)(\forall y) (x = y \implies (\mathscr{A}(x,x) \implies \mathscr{A}(x,y))) \text{ (Substitutivity)}$

- Where $\mathscr{A}(x,x)$ refers to any formula of the given theory and $\mathscr{A}(x,y)$ results from $\mathscr{A}(x,x)$ by replacing some occurrences of $x$ by $y$

    + $\mathscr{A}(x,x)$ may contain other variables in addition to $x$
    
__Theorem A1__ (Symmetry) $x = y \implies y = x$

__*Proof*__ Let $\mathscr{A}(x,x)$ be $x = x$ and $\mathscr{A}(x,y)$ be $y = x$ then using __Assumption II__

$$x = y \implies (x = x \implies y = x)$$
Because $A \implies (B \implies C)$ is equivalent to $B \implies (A \implies C)$ we have that $x = x \implies (x = y \implies y = x)$. By __Assumption I__ we have that $x = x \implies y = x$ $\square$

- In the last part we use a rule of inference called __Modus ponens__ [@bilaniuk_problem_2009, Chapter 3, p 11]
    
    + $\{\varphi, \varphi \implies \psi \} \vdash \psi$
    
    + Given the formulas $\varphi$ and $\varphi \implies \psi$ one may infer $\psi$

__Theorem A2__ (Transitivity) $(x = y \land y = z) \implies x = z$

__*Proof*__  By __Assumption II__ and interchanging $x$ and $y$

$$y = x \implies (\mathscr{A}(y,y) \implies \mathscr{A}(y,x))$$
Let $\mathscr{A}(y,y)$ be $y = z$ and $\mathscr{A}(x,y)$ be $x = z$ then $y = x \implies (y = z \implies x = z)$. Because $A \implies (B \implies C)$ is equivalent to $(B \land A) \implies C$ we have that $(y = x \land y = z) \implies x = z$. Let's assume that $x = y \land y = z$. Then if $x = y$ by __Theorem A1__ and $y = z$ we have that $x = z$ $\square$

- In the last part we use the __Deduction Theorem__ [@bilaniuk_problem_2009, Chapter 3, p 13]

    + If $\Sigma$ is  any  set  of  formulas and $\alpha$ and $\beta$ are any formulas, then $\Sigma \vdash \alpha \implies \beta$ if and only if $\Sigma \cup \{\alpha\}\vdash \beta$
    
    + In our case $\Sigma$ is $\emptyset$, $\alpha$ is $x = y \land y = z$ and $\beta$ is $x = z$. Therefore we can infer $(x = y \land y = z) \implies x = z$

# Bibliography

