---
title: "Appendix D Finite, infinite and denumerable sets"
date: 2020-05-28 11:04:33
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}


```

# Finite, infinite and denumerable sets

This section is based on [@mendelson_number_2008, Appendix D, pp 293-313]

## Equinumerous sets

$A$ and $B$ are __equinumerous__ if and only if $(\exists)(F)(F: A \xrightarrow[\text{onto}]{\text{1-1}} B)$. We write $A \approx B$ to mean that $A$ and $B$ are equinumerous

__Theorem__ 

- $A \approx A$

    + $I_A : A \xrightarrow[\text{onto}]{\text{1-1}} A$

- $A \approx B \implies B \approx A$

    + $\begin{split}
        A \approx B & \iff F: A \xrightarrow[\text{onto}]{\text{1-1}} B \\ 
        & \iff F^{-1}: B \xrightarrow[\text{onto}]{\text{1-1}} A \\
        & \iff B \approx A
        \end{split}$

- $(A \approx B \land B \approx C) \implies A \approx C$

    + $\begin{split}
        A \approx B \land B \approx C & \iff (F_1: A \xrightarrow[\text{onto}]{\text{1-1}} B \land F_2: B \xrightarrow[\text{onto}]{\text{1-1}} C) \\
        & \implies F_2 \circ F_1: A \xrightarrow[\text{onto}]{\text{1-1}} C \\
        & \iff A \approx C
        \end{split}$

Two sets which are equinumerous are said to have the same __cardinal number__. We assume that each set $A$ has an associated cardinal number $\mathscr{K}(A)$. Hence, $\mathscr{K}(A) = \mathscr{K}(B) \iff A \approx B$

__Definitions__ $$A \leq B \text{ means } (\exists F)(F: A \xrightarrow[]{\text{1-1}} B)$$ $$A < B \text{ means } A \leq B \land A \not\approx B$$

__Theorem__ 

- $A \leq A$
    
    + $I_A: A \xrightarrow[]{\text{1-1}} A$
    
- $(A \leq B \land B \leq C) \implies A \leq C$

    + $\begin{split}
       A \leq B \land B \leq C & \iff F: A \xrightarrow[]{\text{1-1}} B \land F: B \xrightarrow[]{\text{1-1}} C \\ 
       & \implies F: A \xrightarrow[]{\text{1-1}} C
       \end{split}$
       
- $A \not< A$

    + $\begin{split}
       A \not< A & \iff \neg(A \leq B \land A \not\approx A) \\
       & \iff \neg(A \leq B) \lor \neg(A \not\approx A) \\ 
       & \iff \neg(\exists F)(F: A \xrightarrow[]{\text{1-1}} A) \lor A \approx A \\ 
       & \iff A \approx A
       \end{split}$
       
- $A \subseteq B \implies A \leq B$

    + $I_A: A \xrightarrow[]{\text{1-1}} B$
    
        + $\mathscr{R}(I_A) = A \subseteq B$
        
- $A \leq B \iff (A < B \lor A \approx B)$

    + $\begin{split}
       A < B \lor A \approx B & \iff (A \leq B \land A \not\approx B) \lor A \approx B  \\
       & \iff (A \leq B \lor A \approx B) \land (A \not\approx B \lor A \approx B) \\
       & \iff (A \leq B \lor A \approx B) \\
       & \iff (\exists F)(F: A \xrightarrow[]{\text{1-1}} B) \\
       & \iff A \leq B
       \end{split}$
       
- $A \leq B \iff (\exists W)(W \subseteq B \land A \approx W)$

    + $\begin{split}
       (\exists W)(W \subseteq B \land A \approx W) 
       & \iff (\exists W)(W \subseteq B \land (\exists F)(F: A \xrightarrow[\text{onto}]{\text{1-1}} W)) \\ 
       & \iff (\exists W)(\exists F)(\mathscr{R}(F) = W \subseteq B \land (F: A \xrightarrow[]{\text{1-1}} B)) \\
       & \iff A \leq B
       \end{split}$
       
- $(A \leq B \land B \approx C) \implies A \leq C$

    + $\begin{split}
       (A \leq B \land B \approx C) & \iff F: A \xrightarrow[]{\text{1-1}} B \land G: B \xrightarrow[\text{onto}]{\text{1-1}} C \\
       & \implies G \circ F: A \xrightarrow[]{\text{1-1}} C
       & \iff A \leq C
       \end{split}$

- $(A \leq B \land A \approx C) \implies C \leq B$

    + $\begin{split}
       (A \leq B \land A \approx C) & \iff F: A \xrightarrow[]{\text{1-1}} B \land G: A \xrightarrow[\text{onto}]{\text{1-1}} C \\
       & \iff F: A \xrightarrow[]{\text{1-1}} B \land G^{-1}: C \xrightarrow[\text{onto}]{\text{1-1}} A \\
       & \implies F \circ G^{-1}: C \xrightarrow[]{\text{1-1}} B  
       \end{split}$
       
Let $\mathscr{K}$ be the set of cardinal numbers. We can define the binary relations $<_\mathscr{K}$ and  $\leq_\mathscr{K}$ on $\mathscr{K}$. Let $\mathfrak{m}$ and $\mathfrak{n}$ be cardinal numbers. Take any sets $A$ and $B$ such that $\mathscr{K}(A) = \mathfrak{m}$ and $\mathscr{K}(B) = \mathfrak{n}$. Then:

$$\mathfrak{m} <_\mathscr{K} \mathfrak{n} \iff A < B$$
$$\mathfrak{m} \leq_\mathscr{K} \mathfrak{n} \iff A \leq B$$
__Theorem__ (_Cantor's Theorem_) $A < \mathscr{P}(A)$

- $\theta(a) = \{a\}$ for all $a \in A$

    + $\theta: A \xrightarrow[]{\text{1-1}} \mathscr{P}(A)$
    
- We need to prove that $A \not\approx \mathscr{P}(A) \iff \neg(\exists G)(G: A \xrightarrow[\text{onto}]{\text{1-1}} \mathscr{P}(A))$

    + Let as assume that $(\exists G)(G: A \xrightarrow[\text{onto}]{\text{1-1}} \mathscr{P}(A))$
    
        + Therefore if $a \in A$ then $G(a)$ is a set. In that sense, $(\forall a)(a \in A \implies a \in G(a) \lor a \notin G(a))$
        
            + Let $B = \{b: b \in A \land b \notin G(b) \}$. We have that $B \neq \emptyset$ because there most be some $c \in A$ such that $G(c) = \emptyset$ where $c \notin \emptyset$

                + $\begin{split}
                    B \subset A & \iff B \in \mathscr{P}(A) \\
                    & \implies (\exists x)(x \in A \land G(x) = B)  
                    \end{split}$
                    
                    + If $x \in B$ then $x \notin G(x)$ but $G(x) = B$ so $x \notin G(x)$ where this is a contradiction
                    
                    + If $x \notin B$ then $x \notin G(x)$ so $x \in B$ where this is a contradiction
                    
        + The assumption $(\exists G)(G: A \xrightarrow[\text{onto}]{\text{1-1}} \mathscr{P}(A))$ generates a contradiction so $\neg(\exists G)(G: A \xrightarrow[\text{onto}]{\text{1-1}} \mathscr{P}(A))$
        
__Theorem__ (_Schröder–Bernstein Theorem_) $(A \leq B \land B \leq A) \implies A \approx B$ [See [@hinkis_proofs_2013] for a compilation of proofs of this theorem]

- Let assume that $A \cap B = \emptyset$

    + If this is not the case we could use $A_1 = A \times {0}$ and $B_1 = B \times {1}$ where 
    $A_1 \cap B_1 = \emptyset$, $A_1 \approx A$ and $B_1 \approx B$
    
- $(A \leq B \land B \leq A) \iff F: A \xrightarrow[]{\text{1-1}} B \land G: B \xrightarrow[]{\text{1-1}} A$

    + By a __B-thread__ we mean a function $f: \mathbb{P} \longrightarrow A \cup B$ such that:
    
        + $y_1 = f(1) \in B - \mathscr{R}(F)$
        
        + $f(n) \in B \implies f(n + 1) = G(f(n))$
        
        + $f(n) \in A \implies f(n + 1) = F(f(n))$
        
            + $\mathscr{D}(f) = \mathbb{P}$
            
            + $\mathscr{R}(f) \subseteq A \cup B$
            
            + $f$ is a function because $A \cap B = \emptyset$, $G$ is a function and $F$ is also a function  
    
    + Therefore a B-thread is an infinite sequence of points starting with a point $y_1 \in B \land y_1 \notin \mathscr{R}(F)$ and then using the functions $G$ and $F$ alternately
            
        + Let $W = \{x: x \in A \land (\exists f)(x \in \mathscr{R}(f)) \}$. Therefore $A = W \cup A - W$
        
            + If $f(1) \in B - \mathscr{R}(F)$ then $f(2) = G(f(1)) \in A$ so $G(f(1)) \in W$. Therefore $W \neq \emptyset$
            
                + In the case of $B = \emptyset$ we will have that $F: A \xrightarrow[]{\text{1-1}} \emptyset$ where $\mathscr{R}(F) = \emptyset$ so $F: A \xrightarrow[\text{onto}]{\text{1-1}} \emptyset$ so there would be nothing to prove    
            
        + Let $H: A \longrightarrow B$ such that:
            
            + $H(x) = \begin{cases}
                       G^{-1}(x) & \text{ if } x \in W \\
                       F(x)      & \text{ if } x \in A - W
                       \end{cases}$
                           
            + $\mathscr{D}(H) = W \cup A - W = A$
                
            + $\mathscr{R}(H) \subseteq B$
                
            + $H$ is a function because $W \cap A - W = \emptyset$, $G^{-1}$ is a function and $F$ is also a function
                
                + What we want to prove is that $H: A \xrightarrow[\text{onto}]{\text{1-1}} B$
                    
                    + Let's prove that $\mathscr{R}(H) = B$
                        
                        + If $b \in B$ then $G(b)$ and there will be two possible cases:
                            
                            + $\begin{split}
                                G(b) \in W & \implies H(G(b)) = G^{-1}(G(b)) = b \\ 
                                & \implies b \in \mathscr{R}(H)
                                \end{split}$ 
                                
                            + $\begin{split}
                                G(b) \in A - W & \implies (\exists x)(F(x) = b) \land G(b) \notin W\\
                                & \implies x \in A \land b \notin W \\
                                & \implies x \in A - W \\
                                & \implies H(x) = b \\
                                & \implies b \in \mathscr{R}(H)
                                \end{split}$
                                
                            + Therefore $\mathscr{R}(H) = B$
                            
                        + Let's prove that $H(x) = H(y) \implies x = y$ which is equivalent to $\neg (H(x) = H(y) \land x \neq y)$
                        
                            + Suppose instead that $H(x) = H(y) \land x \neq y$
                            
                                + $\begin{split}
                                    H(x) = H(y) \land x \neq y & \implies x \in W \land y \in A - W \\
                                    & \implies G^{-1}(x) = F(y) \land y \notin W \\
                                    & \implies G(G^{-1}(x)) = G(F(y)) \land y \notin W \\
                                    & \implies x = G(F(y)) \land y \notin W \\
                                    & \implies G(F(y)) \in W \land y \notin W \\
                                    & \implies G(F(y)) \in \mathscr{R}(f) \land y \notin W \\
                                    & \implies F(y) \in \mathscr{R}(f) \land y \notin W \\
                                    & \implies y \in \mathscr{R}(f) \land y \notin W \\
                                    & \implies y \in W \land y \notin W \\
                                    & \implies \text{There is a contradiction}
                                    \end{split}$
                                    
                                + Therefore $H(x) = H(y) \land x \neq y$ is false so $\neg (H(x) = H(y) \land x \neq y)$ is true 

__Corollary__

- $A \leq B \implies B \not < A$

    + $\begin{split}
        (A \leq B \implies B \not < A) & \iff (A \leq B \implies \neg(B < A)) \\
        & \iff \neg(A \leq B \land \neg\neg(B < A)) \\
        & \iff \neg(A \leq B \land B < A)
        \end{split}$
        
        + Suppose instead that $A \leq B \land B < A$
        
            + $\begin{split}
                A \leq B \land B < A & \iff A \leq B \land B \leq A \land A \not\approx B \\
                & \iff A \approx B \land A \not\approx B \\
                & \iff \text{There is a contradiction}
                \end{split}$
                
            + Therefore $A \leq B \land B < A$ is false so $\neg(A \leq B \land B < A)$ is true
            
- $(A \leq B \land B < C) \implies A < C$

    + $\begin{split}
        A \leq B \land B < C & \iff A \leq B \land B \leq C \land B \not\approx C \\
        & \implies A \leq C \land B \not\approx C \\
        & \implies A \leq C \land (\neg(B \leq C) \lor \neg(C \leq B)) \\
        & \implies A \leq C \land \neg(C \leq B)
        \end{split}$
        
        + Suppose instead that $A \approx C$
        
            + $\begin{split}
                A \approx C & \implies C \leq A \\
                & \implies C \leq B
                \end{split}$
                
                + However $A \leq B \land B < C$ implies $A \leq C \land \neg(C \leq B)$. Therefore if $A \leq B \land B < C$ it must be the case that $A \not\approx C$. Thus, we have that $A \leq C \land A \not\approx C$ which implies $A < C$
                
- $(A < B \land B \leq C) \implies A < C$

    + $\begin{split}
        A < B \land B \leq C & \iff A \leq B \land B \leq C \land A \not\approx B \\
        & \implies A \leq C \land A \not\approx B \\
        & \implies A \leq C \land (\neg(A \leq B) \lor \neg(B \leq A)) \\
        & \implies A \leq C \land \neg(B \leq A)
        \end{split}$
        
        + Suppose instead that $A \approx C$
        
            + $\begin{split}
                A \approx C & \implies C \leq A \\
                & \implies C < B \\
                & \implies C < C
                \end{split}$
                
                + However $C < C$ is false. Therefore if $A < B \land B \leq C$ it must be the case that $A \not\approx C$. Thus, we have that $A \leq C \land A \not\approx C$ which implies $A < C$
                
- $(A < B \land B < C) \implies A < C$

    + $\begin{split}
        A < B \land B < C & \implies A < B \land B \leq C \\
        & \implies A < C
        \end{split}$
                   
## Equinumerous sets exercises and solutions

1. If $A \approx \emptyset$, prove that $A = \emptyset$

    + $\begin{split}
        A \approx \emptyset & \iff F: A \xrightarrow[\text{onto}]{\text{1-1}} \emptyset \\
        & \implies \mathscr{D}(F) = A \land \mathscr{D}(F) = \emptyset \land F \subseteq \mathscr{D}(F) \times \mathscr{R}(F) = \emptyset \\
        & \implies A = \{y: (\exists x)((x,y) \in F) \} \land F = \emptyset \\
        & \implies A = \{y: (\exists x)((x,y) \in \emptyset) \} \\
        & \implies A = \emptyset
        \end{split}$

2. Prove $A \approx A \times \{ y \}$ (Hint: Let $\theta(x) = (x,y)$ for all $x$ in $A$)

    + $\begin{split}
        \theta(a) = \theta(b) & \implies (a,y) = (b, y) \\
        & \implies a = b \land y = y \\
        & \implies a = b
        \end{split}$
        
    + $\mathscr{R}(\theta) = A \times \{ y \}$
        
        + $\mathscr{R}(\theta) = \{ (a,y): (\exists x)(\theta(x) = (a,y)) \}$
    
        + $A \times \{ y \} = \{ (a,y): a \in A \land y \in \{ y \} \}$
        
            + $\mathscr{R}(\theta) \subseteq A \times \{ y \}$

            + $\begin{split}
                (a,y) \in A \times \{ y \} & \implies \theta(a) = (a,y) \\
                & \implies (a,y) \in \mathscr{R}(\theta) 
                \end{split}$

3. Prove that there is no universal set $U$ of all sets. (Hint: $\mathscr{P}(U) \leq U$ since $\mathscr{P}(U) \subseteq U$. Use $A \leq B \implies B \not < A$ and $A < \mathscr{P}(A)$)

    + Assume that there is universal set $U$ of all sets 

        + $\begin{split}
            \mathscr{P}(U) \subseteq U & \implies \mathscr{P}(U) \leq U \\
            & \implies U \not < \mathscr{P}(U)
            \end{split}$
            
        + Also by the _Cantor's Theorem_ $U < \mathscr{P}(U)$
    
            + Therefore we have that $U \not < \mathscr{P}(U) \land U < \mathscr{P}(U)$ which is a contradiction. Thus, by _reductio ad absurdum_ there is no universal set $U$ of all sets
            
__Observation__: Another apparently simple fact about sets is that for any sets $A$ and $B$ we have that $A \leq B$ or $B \leq A$.

- In terms of cardinal numbers this reads as: for any cardinal numbers $\mathfrak{m}$ and $\mathfrak{n}$ then $\mathfrak{m} \leq_{\mathscr{K}} \mathfrak{n}$ or $\mathfrak{n} \leq_{\mathscr{K}} \mathfrak{m}$    

This result is known as _Trichotomy Law_ and it turns out to be equivalent to a proposition of set theory known as _Axiom of Choice_.

- A _choice function_ on a collection $A$ of nonempty sets is a function $F$ with $\mathscr{D}(F) = A$ such that $F(B) \in B$ for every $B \in A$

    + For example if $A = \{ \{ 0 \}, \{ 1 \}, {0,1}\}$ then the possible choice functions are:
    
        + $F_1(\{ 0 \}) = 0 \in \{ 0 \}, F_1(\{ 1 \}) = 1 \in \{ 1 \}, F_1(\{ 0, 1 \}) = 0 \in \{ 0, 1 \}$
        
        + $F_2(\{ 0 \}) = 0 \in \{ 0 \}, F_2(\{ 1 \}) = 1 \in \{ 1 \}, F_2(\{ 0, 1 \}) = 1 \in \{ 0, 1 \}$
        
        + In the example $A$ is a finite set where we will define later the meaning of $A$ being finite. What happen when $A$ is not finite? See for example [@houston-edwards_how_2017] of how the axiom of choice is used
        
__Notation__ If $n$ is a natural number, let $$P_n = \{k: k \in \mathbb{P} \land k \leq n \}$$
For example $P_1 = \{ 1\}$, $P_2 = \{ 1,2 \}$ and $P_3 = \{ 1,2,3 \}$

__Definition__ We say that $A$ is __finite__ if and only if $A = \emptyset$ or $(\exists n)(A \approx P_n)$

__Definition__ We say that $A$ is __infinite__ if and only if $A$ is not finite:

- $\begin{split}
    A \text{ is not finite} & \iff \neg(A = \emptyset \lor (\exists n)(A \approx P_n) \\
    & \iff A \neq \emptyset \land \neg(\exists n)(A \approx P_n) \\
    & \iff A \neq \emptyset \land (\forall n)(\neg(\exists F)(F: A \xrightarrow[\text{onto}]{\text{1-1}} P_n))
    \end{split}$

__Lemma__ $(\forall A)(A \subsetneq P_n \implies A \not\approx P_n)$

- Some definitions and results before the proof 

    + $S(P_n) = \{ A: A \subsetneq P_n \}$
    
    + $S(P_{n+1}) = \mathscr{P}(P_n) \cup \bigcup_{S_i \in S(P_n)} S_i \cup \{n+1\}$

        + For example:
    
            + $S(P_1) = \{ \emptyset \}$
            
            + $\mathscr{P}(P_1) = \{ \emptyset, \{ 1 \} \}$
      
            + $S(P_2) = \{ \emptyset, \{ 1 \}, \{ 2 \} \} = \{ \emptyset, \{ 1 \} \} \cup \{\{ 2 \} \} = \mathscr{P}(P_1) \cup \bigcup_{S_i \in S(P_1)} S_i \cup \{2\}$
      
    + $S(P_n) \subseteq S(P_{n+1})$
    
- Proof by induction

    + _Base step_ $(n = 1)$

        + $P_1 = \{ 1 \} \land \mathscr{P}(P_1) - P_1 = \{ \emptyset \}$
    
            + $\emptyset \not\approx P_1$ because if $(\exists F)(F: \emptyset \longrightarrow P_1)$ then $\mathscr{R}(F) = \emptyset \neq P_1$. Therefore $F$ is not a function from $\emptyset$ onto $P_1$
        
    + _Induction hypothesis_ $(n = k)$:
        + $\begin{split}
            (\forall A)(A \subsetneq P_{k + 1} \implies A \not\approx P_{k + 1}) & \iff (\forall A)\neg(A \subsetneq P_{k+1} \land A \approx P_{k+1}) \\
            & \iff (\forall A)\neg(A \subsetneq P_{k+1} \land P_{k+1} \approx A) \\
            & \iff (\forall A)\neg(A \subsetneq P_{k+1} \land (\exists F)(F: P_{k+1} \xrightarrow[\text{onto}]{\text{1-1}} A))
            \end{split}$

    + _Induction step_ $(n = k + 1)$:

        + Suppose instead that:
    
            + $\begin{split}
               \neg(\forall A)(A \subsetneq P_{k + 1} \implies A \not\approx P_{k + 1}) & \iff (\exists A)(A \subsetneq P_{k+1} \land A \approx P_{k+1}) \\
               & \iff (\exists A)(A \subsetneq P_{k+1} \land P_{k+1} \approx A) \\
               & \iff (\exists A)(A \subsetneq P_{k+1} \land (\exists F)(F: P_{k+1} \xrightarrow[\text{onto}]{\text{1-1}} A))
            \end{split}$
            
                + __Case 1__: $F(i) \neq k + 1$ $(\forall i)(i \in P_{k+1} \land i \leq k)$
                
                    + If $F(i) \neq k + 1$ and $F: P_{k+1} \xrightarrow[\text{onto}]{\text{1-1}} A$ then the restriction of $F$ to $P_n$ is a one-one correspondence between $P_n$ and $A - \{ n + 1 \} = S_i \in S(P_n)$
                    
                        + The restriction of $F$ to $P_n$ is a function G such that $\mathscr{D} = P_n$, and $(\forall i)(i \in P_n \implies G(i) = F(i))$. Also because $F(i) \neq k + 1$ we have that $\mathscr{R}(G) = A - {n + 1} = S_i \in S(P_n)$. Therefore $G$ is a one-one correspondence between $P_n$ and $A - \{ n + 1 \}$   
                        
                        + This result contradicts the induction hypothesis so it is not the case that $F: P_{k+1} \xrightarrow[\text{onto}]{\text{1-1}} A$ when $F(i) \neq k + 1$ $(\forall i)(i \in P_{k+1} \land i \leq k)
                        
                + __Case 2__: $F(i) = k + 1$ for some $i \leq k$ that belongs to $A$
                
                    + We can define the following function
                    
                        + $H(j) = \begin{cases}
                                   F(i)     & \text{if } j \leq k \land j \neq i \\
                                   F(k + 1) & \text{if } j = i
                                   \end{cases}$
                                   
                        + We have that $\mathscr{D}(H) = P_n$, $\mathscr{R}(H) = A - \{ n + 1 \} = S_i \in S(P_n)$ and is a one-one correspondence between $P_n$ and $A - \{ n + 1 \} = S_i \in S(P_n)$ because $F$ is a one-one correspondence.
                        
                            + This result contradicts the induction hypothesis so it is not the case that $F: P_{k+1} \xrightarrow[\text{onto}]{\text{1-1}} A$ when $F(i) \neq k + 1$ for some $i \leq k$ that belongs to $A$
                        
__Theorem__ If $B$ is finite then $B$ is not equinumerous with any of its proper subsets.

- If $B = \emptyset$ then $B$ has no proper subsets

- If $B = \emptyset$ and finite then:

    + $\begin{split}
        B \text{ is finite} \land (\exists C)(C \subsetneq B)& \iff (\exists n)(B \approx P_n) \land (\exists C)(C \subsetneq B) \\
        & \iff (\exists n)(F: B \xrightarrow[\text{onto}]{\text{1-1}} P_n) \land (\exists C)(C \subsetneq B) \\
        & \implies \text{The restriction } F \text{ to } C \text{ is a subset of } P_n
        & \implies C \not\approx P_n \\
        & \implies C \not\approx B
        \end{split}$

# Bibliography

