---
title: "Appendix D Finite, infinite and denumerable sets"
date: 2020-05-28 11:04:33
output:
  html_document:
    code_folding: hide
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: flatly
    toc: yes
    toc_depth: 2
bibliography: number_systems.bib
link-citations: yes
---

<style>

body {
text-align: justify}

</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE, 
                      warning   = FALSE, 
                      message   = FALSE,
                      fig.align = "center")
```

```{r libraries}


```

# Finite, infinite and denumerable sets

This section is based on [@mendelson_number_2008, Appendix D, pp 293-313]

## Equinumerous sets

$A$ and $B$ are __equinumerous__ if and only if $(\exists)(F)(F: A \xrightarrow[\text{onto}]{\text{1-1}} B)$. We write $A \approx B$ to mean that $A$ and $B$ are equinumerous

__Theorem__ 

- $A \approx A$

    + $I_A : A \xrightarrow[\text{onto}]{\text{1-1}} A$

- $A \approx B \implies B \approx A$

    + $\begin{split}
        A \approx B & \iff F: A \xrightarrow[\text{onto}]{\text{1-1}} B \\ 
        & \iff F^{-1}: B \xrightarrow[\text{onto}]{\text{1-1}} A \\
        & \iff B \approx A
        \end{split}$

- $(A \approx B \land B \approx C) \implies A \approx C$

    + $\begin{split}
        A \approx B \land B \approx C & \iff (F_1: A \xrightarrow[\text{onto}]{\text{1-1}} B \land F_2: B \xrightarrow[\text{onto}]{\text{1-1}} C) \\
        & \implies F_2 \circ F_1: A \xrightarrow[\text{onto}]{\text{1-1}} C \\
        & \iff A \approx C
        \end{split}$

Two sets which are equinumerous are said to have the same __cardinal number__. We assume that each set $A$ has an associated cardinal number $\mathscr{K}(A)$. Hence, $\mathscr{K}(A) = \mathscr{K}(B) \iff A \approx B$

__Definitions__ $$A \leq B \text{ means } (\exists F)(F: A \xrightarrow[]{\text{1-1}} B)$$ $$A < B \text{ means } A \leq B \land A \not\approx B$$

## Equinumerous sets exercises and solutions

1. If $A \approx \emptyset$, prove that $A = \emptyset$

    + $\begin{split}
        A \approx \emptyset & \iff F: A \xrightarrow[\text{onto}]{\text{1-1}} \emptyset \\
        & \implies \mathscr{D}(F) = A \land \mathscr{D}(F) = \emptyset \land F \subseteq \mathscr{D}(F) \times \mathscr{R}(F) = \emptyset \\
        & \implies A = \{y: (\exists x)((x,y) \in F) \} \land F = \emptyset \\
        & \implies A = \{y: (\exists x)((x,y) \in \emptyset) \} \\
        & \implies A = \emptyset
        \end{split}$

2. Prove $A \approx A \times \{ y \}$ (Hint: Let $\theta(x) = (x,y)$ for all $x$ in $A$)

    + $\begin{split}
        \theta(a) = \theta(b) & \implies (a,y) = (b, y) \\
        & \implies a = b \land y = y \\
        & \implies a = b
        \end{split}$
        
    + $\mathscr{R}(\theta) = A \times \{ y \}$
        
        + $\mathscr{R}(\theta) = \{ (a,y): (\exists x)(\theta(x) = (a,y)) \}$
    
        + $A \times \{ y \} = \{ (a,y): a \in A \land y \in \{ y \} \}$
        
            + $\mathscr{R}(\theta) \subseteq A \times \{ y \}$

            + $\begin{split}
                (a,y) \in A \times \{ y \} & \implies \theta(a) = (a,y) \\
                & \implies (a,y) \in \mathscr{R}(\theta) 
                \end{split}$
    

3. Prove that there is no universal set $U$ of all sets. (Hint: $\mathscr{P}(U) \leq U$ since $\mathscr{P}(U) \subseteq U$. Use $A \leq B \implies B \not < A$ and $A < \mathscr{P}(A)$)

# Bibliography

